---
title: "Geodesic Pathwise Polynomial Annihilator Experiments"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.5,
  fig.height = 6.2,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = "ragg_png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "geodesic_annihilator_experiments.R"))
library(ggplot2)

full.path <- file.path(report.dir, "geodesic_annihilator_full.rds")
smoke.path <- file.path(report.dir, "geodesic_annihilator_smoke.rds")
requested.mode <- Sys.getenv("GFLOW_GEO_ANNIHILATOR_MODE", unset = "")
result.path <- if (requested.mode == "smoke" && file.exists(smoke.path)) {
  smoke.path
} else if (requested.mode == "full" && file.exists(full.path)) {
  full.path
} else if (file.exists(full.path)) {
  full.path
} else {
  smoke.path
}
diagnostics <- readRDS(result.path)
summary.table <- diagnostics$results
nullity.table <- diagnostics$nullity
vertex.table <- diagnostics$vertex.table
path.table <- diagnostics$path.table
local.dimension.table <- diagnostics$local.dimension.table %||% data.frame()
local.dimension.vote.table <- diagnostics$local.dimension.vote.table %||% data.frame()
score.table <- diagnostics$score.table
if (!"disk.radius.rule" %in% names(summary.table)) {
  summary.table$disk.radius.rule <- "fixed.k"
}
if (!"disk.radius.rule" %in% names(nullity.table)) {
  nullity.table$disk.radius.rule <- "fixed.k"
}
if (!"disk.radius.rule" %in% names(vertex.table)) {
  vertex.table$disk.radius.rule <- "fixed.k"
}
if (!"selected.disk.k" %in% names(vertex.table)) {
  vertex.table$selected.disk.k <- vertex.table$disk.k
}
if (!"disk.radius.rule" %in% names(score.table)) {
  score.table$disk.radius.rule <- "fixed.k"
}
if (nrow(local.dimension.table) > 0L && !"disk.radius.rule" %in% names(local.dimension.table)) {
  local.dimension.table$disk.radius.rule <- "fixed.k"
}
if (nrow(local.dimension.vote.table) > 0L && !"disk.radius.rule" %in% names(local.dimension.vote.table)) {
  local.dimension.vote.table$disk.radius.rule <- "fixed.k"
}
if (!"analysis.group" %in% names(summary.table)) {
  summary.table$analysis.group <- "smoke"
}
if (!"path.family" %in% names(summary.table)) summary.table$path.family <- "local.composite.geodesics"
if (!"path.family" %in% names(vertex.table)) vertex.table$path.family <- "local.composite.geodesics"
if (!"path.family" %in% names(path.table)) path.table$path.family <- "local.composite.geodesics"
if (!"path.family" %in% names(score.table)) score.table$path.family <- "local.composite.geodesics"
if (!"row.selection" %in% names(summary.table)) summary.table$row.selection <- "none"
if (!"row.selection" %in% names(vertex.table)) vertex.table$row.selection <- "none"
if (!"row.selection" %in% names(path.table)) path.table$row.selection <- "none"
if (!"skeleton.coverage" %in% names(vertex.table)) vertex.table$skeleton.coverage <- NA_integer_
if (!"skeleton.distance" %in% names(vertex.table)) vertex.table$skeleton.distance <- NA_real_
if (!"center.order" %in% names(summary.table)) summary.table$center.order <- "natural"
if (!"center.order" %in% names(vertex.table)) vertex.table$center.order <- "natural"
if (!"center.order" %in% names(path.table)) path.table$center.order <- "natural"
if (!"candidate.cap.mode" %in% names(summary.table)) summary.table$candidate.cap.mode <- "random"
if (!"candidate.cap.mode" %in% names(vertex.table)) vertex.table$candidate.cap.mode <- "random"
if (!"candidate.cap.mode" %in% names(path.table)) path.table$candidate.cap.mode <- "random"
if (!"candidate.operator.rows" %in% names(vertex.table)) {
  vertex.table$candidate.operator.rows <- vertex.table$n.operator.rows
}
if (!"selected.operator.rows" %in% names(vertex.table)) {
  vertex.table$selected.operator.rows <- vertex.table$n.operator.rows
}
for (nm in c("thinning.local.dim", "thinning.polynomial.dim", "thinning.rank",
             "thinning.coverage", "thinning.residual.min",
             "thinning.residual.median", "thinning.residual.max")) {
  if (!nm %in% names(vertex.table)) vertex.table[[nm]] <- NA_real_
}
if (nrow(local.dimension.table) > 0L && !"row.selection" %in% names(local.dimension.table)) {
  local.dimension.table$row.selection <- "none"
}
if (nrow(local.dimension.table) > 0L && !"center.order" %in% names(local.dimension.table)) {
  local.dimension.table$center.order <- "natural"
}
if (nrow(local.dimension.table) > 0L && !"candidate.cap.mode" %in% names(local.dimension.table)) {
  local.dimension.table$candidate.cap.mode <- "random"
}
if (nrow(local.dimension.vote.table) > 0L && !"row.selection" %in% names(local.dimension.vote.table)) {
  local.dimension.vote.table$row.selection <- "none"
}
if (nrow(local.dimension.vote.table) > 0L && !"center.order" %in% names(local.dimension.vote.table)) {
  local.dimension.vote.table$center.order <- "natural"
}
if (nrow(local.dimension.vote.table) > 0L && !"candidate.cap.mode" %in% names(local.dimension.vote.table)) {
  local.dimension.vote.table$candidate.cap.mode <- "random"
}
ok.summary <- subset(summary.table, status == "ok")

wrap.graph <- function(x) {
  gsub("delaunay.1skeleton", "Delaunay\n1-skeleton",
       gsub("symmetric.knn", "symmetric\nkNN", x))
}
wrap.target <- function(x) {
  x <- gsub("intrinsic.degree.leq.d", "intrinsic\ndegree <= d", x)
  x <- gsub("intrinsic.exact.degree.d.plus.1", "intrinsic\nexact degree d+1", x)
  x <- gsub("ambient.linear", "ambient\nlinear", x)
  x <- gsub("ambient.quadratic", "ambient\nquadratic", x)
  x
}
wrap.radius.rule <- function(x) {
  x <- gsub("geodesic.skeleton", "geodesic\nskeleton", x)
  gsub("min.composite.geodesics", "adaptive\nmin geodesics",
       gsub("fixed.k", "fixed\nk_disk", x))
}
radius.rule.levels <- wrap.radius.rule(c(
  "fixed.k", "min.composite.geodesics", "geodesic.skeleton"
))
wrap.path.family <- function(x) {
  x <- gsub("geodesic.skeleton.cross.enriched", "skeleton\n+ cross", x)
  x <- gsub("geodesic.skeleton", "skeleton", x)
  gsub("local.composite.geodesics", "local\ncomposite", x)
}
base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(margin = margin(b = 8)),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 10)),
      strip.text = element_text(face = "bold", size = rel(0.85)),
      legend.position = "bottom",
      plot.margin = margin(10, 24, 10, 24)
    )
}
```

## Purpose

This report evaluates an experimental graph operator intended to annihilate
intrinsic low-degree polynomials by sampling geodesic paths in local graph
disks. For a weighted graph, the disk around vertex \(v\) is

\[
B_G(v,\varepsilon)=\{u\in V : d_G(u,v)\le \varepsilon\},
\]

where \(d_G\) is graph-geodesic distance. The implementation samples composite
geodesic paths through \(v\), constructs one-dimensional polynomial annihilator
rows along those paths, and stacks the rows into a sparse matrix \(A\). If the
construction behaves as an intrinsic polynomial annihilator on a sampled
surface, then sampled intrinsic polynomials \(p\) of degree at most \(d\) should
satisfy approximately

\[
A p \approx 0.
\]

The experiments ask two questions. First, on sampled \([0,1]^2\), does the
estimated nullity of \(A\) stabilize near the dimensions

\[
\dim \mathcal P_{\le d}(\mathbb R^2)=\binom{d+2}{2}=3,6,10
\quad\text{for}\quad d=1,2,3?
\]

Second, on sampled quadrics, does the operator respond to intrinsic polynomial
structure in the surface parameters rather than merely to ambient coordinate
polynomials in \((X,Y,Z)\)?

The graph families tested here are the Delaunay 1-skeleton, symmetric kNN,
radius graphs, and cKNN graphs. Two radius rules are compared. The fixed rule
uses the graph distance to the \(k_{\mathrm{disk}}\)-th graph-geodesic neighbor.
The adaptive rule increases \(k_{\mathrm{disk}}\) until the number of available
composite geodesics reaches a specified minimum, or until a maximum search size
is reached. Detailed run tables and reproducibility code are collected in the
appendix so the main text can focus on the figures.

## Disk Size

The disk radius controls what local geometry the operator sees. If disks are too
small, there may be too few usable geodesic paths to identify multivariate
polynomial structure. If disks are too large, the operator may mix geometry from
regions where a local polynomial approximation is no longer appropriate. The
first diagnostic therefore measures the distribution of

\[
|B_G(v,\varepsilon_v)|
\]

across vertices. Under the adaptive rule, \(\varepsilon_v\) is selected as

\[
\varepsilon_v
=\min \{\varepsilon_{v,k}: |\Gamma(v,\varepsilon_{v,k})|\ge m_{\min}\},
\]

subject to the maximum allowed search size.

```{r disk-histogram, fig.height=7.0}
disk.plot.data <- transform(
  vertex.table,
  graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
    "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
  ))),
  radius.rule.label = factor(wrap.radius.rule(disk.radius.rule),
                             levels = radius.rule.levels)
)
ggplot(disk.plot.data, aes(x = n.disk.vertices, fill = dataset)) +
  geom_histogram(bins = 26, color = "white", alpha = 0.82, position = "identity") +
  facet_grid(graph.label ~ radius.rule.label, scales = "free_y") +
  labs(
    title = "Distribution of graph-disk sizes",
    subtitle = "Adaptive rule increases the disk until enough composite geodesics are available.",
    x = "vertices in disk",
    y = "number of centers",
    fill = "dataset"
  ) +
  base.theme()
```

The adaptive rule often selects larger local disks than the fixed rule. This is
expected: it is no longer trying to equalize disk size, but instead to guarantee
a minimum supply of usable composite geodesics. The figure also shows that the
extra disk growth is graph-dependent, which means the selected radius itself is
an important diagnostic of local graph geometry.

## Selected Radius Search Size

The adaptive rule changes the radius by increasing the neighbor index
\(k_{\mathrm{disk}}\). The selected value is

\[
k_v^\star=\min\{k: |\Gamma(v,\varepsilon_{v,k})|\ge m_{\min}\}.
\]

If \(k_v^\star\) is frequently near the maximum allowed value, the graph does
not provide enough local composite geodesics at the tested scale.

```{r selected-k-histogram, fig.height=7.0}
ggplot(disk.plot.data, aes(x = selected.disk.k, fill = dataset)) +
  geom_histogram(bins = 28, color = "white", alpha = 0.82, position = "identity") +
  facet_grid(graph.label ~ radius.rule.label, scales = "free_y") +
  labs(
    title = "Distribution of selected radius search size",
    subtitle = "For the adaptive rule, selected k_disk is the first value reaching the minimum composite-geodesic count.",
    x = "selected k_disk",
    y = "number of centers",
    fill = "dataset"
  ) +
  base.theme()
```

The selected-\(k_{\mathrm{disk}}\) distributions show how much each graph family
must expand its local disk to obtain enough composite geodesics. A concentration
at small values means the local graph already supplies adequate path geometry.
A heavy right tail means the graph requires larger neighborhoods before the
operator can even be assembled with the requested path availability.

## Composite-Geodesic Availability

The operator is built from pathwise annihilator rows. The relevant local sample
size is therefore not just the disk size but the number of composite geodesics
available through each center vertex. For each center \(v\), we count candidate
paths \(\gamma\in\Gamma(v,\varepsilon_v)\) that pass through \(v\) and satisfy the
composite-geodesic test. This diagnostic measures

\[
|\Gamma(v,\varepsilon_v)|.
\]

```{r composite-histogram, fig.height=7.0}
ggplot(disk.plot.data, aes(x = n.composite.available, fill = dataset)) +
  geom_histogram(bins = 30, color = "white", alpha = 0.82, position = "identity") +
  facet_grid(graph.label ~ radius.rule.label, scales = "free_y") +
  labs(
    title = "Distribution of composite-geodesic counts per vertex",
    subtitle = "Counts are measured before the per-vertex cap on sampled paths is applied.",
    x = "available composite geodesics",
    y = "number of centers",
    fill = "dataset"
  ) +
  base.theme()
```

This figure is the main check on whether the adaptive radius rule succeeded.
Compared with the fixed rule, the adaptive rule shifts the composite-geodesic
count distribution away from zero and toward the requested minimum. Remaining
low-count centers identify vertices where even the maximum allowed search size
does not produce enough local path geometry.

## Geodesic Skeleton Coverage

The skeleton row family asks a different question from the local composite
operator. Instead of constructing a disk at every vertex, it first selects a
small set of long graph-geodesic paths. The first path is a graph diameter.
Subsequent passes attach vertices that are poorly covered by the current
skeleton. In the cross-enriched variant, additional local connector geodesics
are added between nearby skeleton paths. The diagnostic below measures the
coverage multiplicity

\[
c(v)=\#\{\gamma_s: v\in\gamma_s\}
\]

and the distance from each vertex to the selected skeleton. A good first
skeleton should cover most vertices at least once; a useful enriched skeleton
should raise coverage in regions that need transverse constraints.

```{r skeleton-coverage-diagnostics, fig.height=6.8}
skeleton.plot <- subset(vertex.table, path.family != "local.composite.geodesics")
if (nrow(skeleton.plot) > 0L) {
  skeleton.plot <- transform(
    skeleton.plot,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    path.family.label = factor(wrap.path.family(path.family),
                               levels = wrap.path.family(c(
                                 "geodesic.skeleton",
                                 "geodesic.skeleton.cross.enriched"
                               )))
  )
  ggplot(skeleton.plot, aes(x = skeleton.coverage, fill = dataset)) +
    geom_histogram(binwidth = 1, boundary = -0.5, color = "white", alpha = 0.82) +
    facet_grid(graph.label ~ path.family.label, scales = "free_y") +
    labs(
      title = "Coverage multiplicity of the geodesic skeleton",
      subtitle = "Cross-enrichment adds connector geodesics between nearby selected skeleton paths.",
      x = "skeleton paths containing vertex",
      y = "number of vertices",
      fill = "dataset"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No geodesic-skeleton runs were recorded.")
}
```

The skeleton family is intentionally sparse. Its first success criterion is not
matching the local composite operator, but obtaining deterministic path coverage
that can support ordinary one-dimensional sliding-window trend-filter rows. If
cross-enrichment improves coverage without making the null space collapse, it is
a promising row-generation strategy for the next round of operator tests.

## Polynomial-Residual Row Thinning

The unthinned operator keeps every pathwise annihilator row that passes the
local geodesic and mirror-extension rules.  The experimental thinning rule first
builds a local metric-MDS polynomial probe matrix \(P_v\) on each graph disk and
then greedily keeps rows with small

\[
\frac{\|aP_v\|_2}{\|P_v\|_F}
\]

while preserving row rank and vertex coverage.  This diagnostic asks how
aggressively the rule reduces the local row stack before we inspect nullity and
polynomial residuals.

The smoke and full grids now use deterministic thinning: centers are processed
in increasing graph-degree order, candidate path caps are resolved by stable
path-length and canonical-path ordering, and residual-score ties are broken by
path length and canonical path label.  This makes repeated runs comparable
without depending on random path subsampling.

```{r row-thinning-diagnostics, fig.height=6.8}
if (all(c("candidate.operator.rows", "selected.operator.rows") %in% names(vertex.table))) {
  thinning.plot <- transform(
    vertex.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    selected.fraction = ifelse(candidate.operator.rows > 0,
                               selected.operator.rows / candidate.operator.rows,
                               NA_real_)
  )
  ggplot(subset(thinning.plot, !is.na(selected.fraction)),
         aes(x = selected.fraction, fill = row.selection)) +
    geom_histogram(bins = 24, color = "white", alpha = 0.82, position = "identity") +
    facet_grid(graph.label ~ row.selection, scales = "free_y") +
    labs(
      title = "Fraction of candidate rows retained per graph disk",
      subtitle = "Polynomial-residual thinning keeps rows that preserve local polynomial probes while maintaining rank and coverage.",
      x = "selected rows / candidate rows",
      y = "number of centers",
      fill = "row selection"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No row-thinning diagnostics were recorded.")
}
```

If this distribution is near one, thinning is behaving conservatively.  If it is
near zero, the local polynomial probe is rejecting most path rows and the
operator may become underconstrained.  The useful regime should reduce redundant
rows while preserving enough coverage for nullity and residual tests.

## Local Coordinate Dimension Diagnostics

The pathwise operator is defined from geodesic paths, but the implementation
roadmap also asks whether each disk looks locally one-, two-, or
three-dimensional after metric embedding. For a local disk \(B_G(v,\varepsilon_v)\),
we embed the induced weighted graph in dimensions \(q=1,\ldots,q_{\max}\) and compute the
relative edge-length error

\[
E_q(v)
=
\left(
\frac{\sum_{(i,j)\in E_v}(\|z_i-z_j\|-\ell_{ij})^2}
     {\sum_{(i,j)\in E_v}\ell_{ij}^2}
\right)^{1/2}.
\]

The fast diagnostic backend uses local graph-geodesic `cmdscale`; selected smoke
runs additionally exercise the edge-KK diagnostic path. For dimensions 2 and 3,
that path can use weighted GRIP as an initializer. For dimensions above 3, it
uses metric MDS followed by edge-KK when the installed `grip` namespace exposes
the newer high-dimensional edge-KK functions. The selected local dimension is a
stress-elbow diagnostic, not a theorem about the null space.

```{r local-dimension-stress, fig.height=6.8}
if (nrow(local.dimension.table) > 0L) {
  local.dim.plot <- transform(
    local.dimension.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    radius.rule.label = factor(wrap.radius.rule(disk.radius.rule),
                               levels = radius.rule.levels),
    method.label = normalize.local.dimension.method.label(local.embedding.method)
  )
  ggplot(subset(local.dim.plot, status == "ok"),
         aes(x = factor(embedding.dim), y = edge.rrmse,
             group = interaction(run.id, center), color = graph.label)) +
    geom_line(alpha = 0.25) +
    geom_point(alpha = 0.55, size = 1.5) +
    facet_grid(method.label ~ radius.rule.label) +
    labs(
      title = "Local edge-length stress by embedding dimension",
      subtitle = "Each line is one sampled graph disk; lower stress means the local embedding realizes edge lengths more accurately.",
      x = "embedding dimension",
      y = "relative edge-length error",
      color = "graph"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No local dimension diagnostics were recorded.")
}
```

This diagnostic separates a geometric question from the annihilator row-stack
question. If stress drops sharply from dimension one to two and then stabilizes,
the local disk behaves like a sampled surface patch. If stress remains large in
dimension three, the current local coordinate engine is at its useful cap and
the downstream thinning rules should treat the disk cautiously.

The stress elbow is only one estimator. The diagnostic table also records a
metric-MDS spectrum elbow and a graph-Laplacian Weyl-slope estimate. Their
agreement or disagreement is useful because local graph disks can be irregular:
one estimator may detect metric embedding dimension, while another detects
diffusive graph dimension.

```{r local-dimension-votes, fig.height=6.2}
if (nrow(local.dimension.vote.table) > 0L) {
  vote.plot <- transform(
    local.dimension.vote.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    method.label = normalize.local.dimension.method.label(local.embedding.method),
    estimator.label = gsub("graph.laplacian.weyl", "Laplacian\nWeyl",
                           gsub("mds.spectrum.elbow", "MDS\nspectrum",
                                gsub("edge.stress.elbow", "edge stress\nelbow",
                                     dimension.method)))
  )
  ggplot(subset(vote.plot, status == "ok" & !is.na(estimated.dimension)),
         aes(x = factor(estimated.dimension), fill = estimator.label)) +
    geom_bar(position = "dodge") +
    facet_grid(method.label ~ graph.label, scales = "free_y") +
    labs(
      title = "Local dimension votes by estimator",
      subtitle = "Agreement is evidence of stable local geometry; disagreement flags disks needing caution before row thinning.",
      x = "estimated local dimension",
      y = "number of sampled graph disks",
      fill = "estimator"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No local dimension vote diagnostics were recorded.")
}
```

```{r local-dimension-selection, fig.height=5.8}
if (nrow(local.dimension.table) > 0L) {
  selected.dim <- unique(local.dimension.table[
    , c("run.id", "center", "dataset", "graph.type", "degree",
        "disk.radius.rule", "local.embedding.method", "selected.local.dim",
        "dimension.cap.hit", "status")
  ])
  selected.dim <- subset(selected.dim, status == "ok")
  selected.dim <- transform(
    selected.dim,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    method.label = normalize.local.dimension.method.label(local.embedding.method)
  )
  ggplot(selected.dim,
         aes(x = factor(selected.local.dim), fill = graph.label)) +
    geom_bar(position = "dodge") +
    facet_wrap(~ method.label) +
    labs(
      title = "Selected local dimension from the stress-elbow rule",
      subtitle = "The selected dimension is a diagnostic used to guide later row thinning, not a final null-space claim.",
      x = "selected local dimension",
      y = "number of sampled disks",
      fill = "graph"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No local dimension diagnostics were recorded.")
}
```

## Nullity At A Fixed Numerical Tolerance

The formal null-space target for a degree-\(d\) intrinsic polynomial annihilator
on a two-dimensional domain is

\[
\dim \ker(A)\approx \binom{d+2}{2}.
\]

The next figure compares the estimated nullity at relative singular-value
threshold \(10^{-6}\) with the target values \(3,6,10\). The dashed line marks
perfect agreement.

```{r nullity-plot, fig.height=6.4}
nullity.plot.data <- transform(
  ok.summary,
  graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
    "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
  ))),
  radius.rule.label = factor(wrap.radius.rule(disk.radius.rule),
                             levels = radius.rule.levels)
)
ggplot(nullity.plot.data,
       aes(x = expected.nullity, y = nullity.1e.6,
           color = graph.label, shape = radius.rule.label)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray40") +
  geom_jitter(width = 0.12, height = 0.12, alpha = 0.78, size = 2.1) +
  scale_x_continuous(breaks = c(3, 6, 10)) +
  facet_grid(dataset ~ n) +
  labs(
    title = "Estimated nullity versus intrinsic polynomial dimension",
    subtitle = "Nullity is computed from singular values using relative tolerance 1e-6.",
    x = "target dimension",
    y = "estimated nullity",
    color = "graph",
    shape = "radius rule"
  ) +
  base.theme()
```

The fixed-threshold nullity still does not stabilize near \(3,6,10\). The
adaptive rule improves the most obvious underconstrained radius-graph failures
for lower orders by forcing more composite geodesics into the operator, but it
does not create the desired polynomial null space. Delaunay, cKNN, and
symmetric kNN configurations remain mostly overconstrained, with estimated
nullity near one. This answers the first nullity question in the negative for
the present construction: path availability is necessary, but not sufficient,
for recovering the expected intrinsic polynomial null-space dimension.

## Nullity Sensitivity

A single numerical tolerance can hide whether the singular spectrum contains a
near-null plateau. We therefore also compute

\[
\widehat{\dim\ker(A;\tau)}
= n - \#\{s_i(A): s_i(A)>\tau s_1(A)\}
\]

for several relative tolerances \(\tau\). A stable intrinsic null space would
appear as a plateau near \(3,6,10\) across a nontrivial tolerance range.

```{r nullity-by-tolerance, fig.height=8.0}
nullity.tol.data <- transform(
  nullity.table,
  graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
    "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
  ))),
  radius.rule.label = factor(wrap.radius.rule(disk.radius.rule),
                             levels = radius.rule.levels)
)
ggplot(nullity.tol.data,
       aes(x = rel.tol, y = nullity,
           color = graph.label,
           group = interaction(graph.label, disk.radius.rule, seed, n,
                               row.normalization, vertex.normalization))) +
  geom_line(alpha = 0.45) +
  geom_point(size = 0.9, alpha = 0.65) +
  geom_hline(aes(yintercept = choose(degree + 2L, 2L)),
             linetype = 2, color = "gray35") +
  scale_x_log10() +
  facet_grid(degree ~ radius.rule.label, scales = "free_y") +
  labs(
    title = "Nullity sensitivity to singular-value tolerance",
    subtitle = "Dashed horizontal lines show the target dimensions 3, 6, and 10.",
    x = "relative singular-value tolerance",
    y = "estimated nullity",
    color = "graph"
  ) +
  base.theme()
```

The sensitivity plot reinforces the fixed-tolerance result. There is no broad,
stable plateau at the target dimensions. The adaptive rule reduces some
large-nullity artifacts by adding rows where the fixed rule had too few paths,
but the resulting spectrum still does not organize itself around the intrinsic
dimensions \(3,6,10\). This indicates that the current pathwise row stack is
better viewed as a residual diagnostic than as a calibrated null-space
construction.

## Polynomial Residual Ratios

Because exact numerical nullity is brittle on finite sampled graphs, we also
evaluate sampled polynomial residuals directly. For a design matrix \(P\), define

\[
R(P)=\frac{\|AP\|_F}{\|P\|_F}.
\]

For a degree-\(d\) annihilator, residuals for intrinsic monomials of degree at
most \(d\) should be smaller than residuals for exact degree \(d+1\) controls.

```{r residual-plot, fig.height=8.0}
residual.long <- rbind(
  transform(ok.summary, target = "intrinsic degree <= d",
            residual = intrinsic.residual),
  transform(ok.summary, target = "degree d+1 control",
            residual = control.residual)
)
residual.long$graph.label <- factor(wrap.graph(residual.long$graph.type),
                                    levels = wrap.graph(c(
                                      "delaunay.1skeleton", "symmetric.knn",
                                      "radius", "cknn"
                                    )))
residual.long$target <- factor(residual.long$target,
                               levels = c("intrinsic degree <= d",
                                          "degree d+1 control"))
residual.long$radius.rule.label <- factor(wrap.radius.rule(residual.long$disk.radius.rule),
                                          levels = radius.rule.levels)
ggplot(residual.long,
       aes(x = graph.label, y = residual, fill = target)) +
  geom_boxplot(outlier.alpha = 0.28, position = position_dodge(width = 0.75)) +
  scale_y_log10() +
  facet_grid(degree ~ radius.rule.label) +
  labs(
    title = "Polynomial residual ratios",
    subtitle = "Intrinsic degree <= d residuals should lie below degree d+1 controls.",
    x = "graph type",
    y = "residual ratio",
    fill = "polynomial family"
  ) +
  base.theme() +
  theme(axis.text.x = element_text(size = 9))
```

The residual-ratio experiment is more encouraging than the nullity experiment.
The adaptive rule generally lowers intrinsic residuals by supplying more path
constraints, and intrinsic degree-\(d\) residuals are usually smaller than
degree-\(d+1\) controls. This means the pathwise rows do encode useful local
polynomial information even though the singular spectrum does not yet produce
the desired nullity. Radius graphs remain the most delicate case: adaptive
expansion helps, but high-order radius-graph operators can still be
underconstrained when the maximum disk size is reached before enough usable
paths are found.

## Intrinsic Versus Ambient Structure On Quadrics

On a sampled quadric, the coordinates \((u,v)\) are intrinsic parameters, while
\((X,Y,Z)\) are ambient coordinates. The intended operator should respond to
polynomial behavior in intrinsic coordinates. For example, on a paraboloid,
\(Z\) is quadratic in \((u,v)\), so it should be treated differently by degree-1
and degree-2 annihilators. The following figure compares residuals for
intrinsic and ambient function families.

```{r score-table, fig.height=8.2}
if (nrow(score.table) > 0L) {
  score.plot <- subset(score.table, status == "ok" & dataset != "square")
  score.plot$graph.label <- factor(wrap.graph(score.plot$graph.type),
                                   levels = wrap.graph(c(
                                     "delaunay.1skeleton", "symmetric.knn",
                                     "radius", "cknn"
                                   )))
  score.plot$target.label <- factor(wrap.target(score.plot$target),
                                    levels = wrap.target(c(
                                      "intrinsic.degree.leq.d",
                                      "intrinsic.exact.degree.d.plus.1",
                                      "ambient.linear",
                                      "ambient.quadratic"
                                    )))
  ggplot(score.plot,
         aes(x = residual.ratio, y = target.label, color = graph.label)) +
    geom_point(position = position_jitter(height = 0.12, width = 0),
               alpha = 0.72, size = 1.7) +
    scale_x_log10() +
    facet_grid(degree ~ dataset) +
    labs(
      title = "Intrinsic and ambient residual ratios on sampled quadrics",
      subtitle = "Residuals compare parameter-polynomial targets with ambient-coordinate targets.",
      x = "residual ratio",
      y = "function family",
      color = "graph"
    ) +
    base.theme()
}
```

The quadric diagnostics suggest partial intrinsic behavior, but not yet a clean
intrinsic polynomial annihilator. Intrinsic targets usually have lower residuals
than degree-\(d+1\) controls, and degree-2 operators improve on quadratic
surface-coordinate behavior relative to degree-1 operators. However, the
ambient and intrinsic families are not cleanly separated across all graphs. This
is expected for a first pathwise construction because ambient coordinates and
intrinsic parameters are strongly correlated on the sampled surfaces, especially
at this sample size.

## Results Summary And Discussion

The experiments answer the main questions in a deliberately conservative way.
The estimated nullity does **not** yet stabilize near \(3,6,10\) for degrees
\(1,2,3\). The adaptive radius rule fixes one important construction problem:
the median number of composite geodesics increases to the requested target for
Delaunay, cKNN, and symmetric kNN graphs, and improves substantially for radius
graphs. However, the expected polynomial nullity still does not emerge. This
separates two issues that were previously entangled. The old fixed-radius rule
could fail simply because too few paths were available; the new adaptive rule
shows that even with adequate path availability, the current row stack is
usually too constraining.

The residual diagnostics are more promising. Even though the operator does not
produce the target null-space dimension, it often yields smaller residuals on
intrinsic degree-\(d\) polynomial families than on degree-\(d+1\) controls. The
adaptive rule usually improves these residuals, which supports the decision to
control path availability directly. The remaining gap is therefore not merely a
radius-selection problem; it is a row-selection or local-coordinate problem.

The quadric experiments also support a cautious interpretation. The operator
shows some sensitivity to intrinsic polynomial structure in the parameter domain,
but it does not yet sharply separate intrinsic polynomial behavior from ambient
coordinate polynomial behavior. At the present sample sizes and graph densities,
this is not surprising: the ambient coordinates are smooth functions of the
intrinsic coordinates, and the current operator is built from one-dimensional
path restrictions rather than from an explicit local tangent-coordinate model.

The graph comparison is informative. Delaunay 1-skeleton graphs are a useful
geometric baseline because they maintain planar local structure and avoid some
of the high-branching behavior of kNN graphs. Symmetric kNN and cKNN graphs give
more composite paths, but they can also impose many overlapping constraints.
Radius graphs are the least stable in this experiment because comparable disk
sizes do not guarantee enough composite geodesics.

The immediate methodological conclusion is that the next version should not aim
only to add more paths. The adaptive rule should remain part of the construction
because it prevents path-starved local operators, but it must be paired with a
path-selection or local-basis step whose goal is to preserve the expected
polynomial null space while maintaining adequate coverage. The new local
coordinate diagnostics make this more concrete: before thinning rows, we can ask
whether each disk has a stable low-dimensional metric realization and whether
the selected local dimension is consistent with the polynomial probe space being
tested. The most useful next experiments are local row thinning by rank
contribution, tangent-coordinate regression annihilators inside graph disks, and
explicit comparison of row normalization and vertex normalization rules. The
current report establishes the diagnostic framework needed to test those choices.

# Appendix

## Run Metadata

```{r appendix-metadata, echo=TRUE}
result.path
 diagnostics$mode
 diagnostics$created
```

## Run Summary Table

```{r appendix-summary-table, echo=TRUE}
summary.table[, c("run.id", "dataset", "n", "seed", "graph.type", "degree",
                  "path.family", "disk.k", "row.normalization", "vertex.normalization",
                  "row.selection", "center.order", "candidate.cap.mode",
                  "analysis.group", "n.edges", "n.operator.rows",
                  "nullity.1e.6", "expected.nullity", "intrinsic.residual",
                  "control.residual", "status")]
```

## Construction Summary Table

```{r appendix-construction-summary, echo=TRUE}
aggregate(
  cbind(n.disk.vertices, n.composite.available, n.paths.used,
        selected.disk.k, candidate.operator.rows, selected.operator.rows,
        n.operator.rows, used.mirror.proxy) ~
    dataset + graph.type + degree + path.family + disk.radius.rule + disk.k + row.selection +
    center.order + candidate.cap.mode,
  data = vertex.table,
  FUN = function(z) if (is.logical(z)) mean(z) else stats::median(z)
)
```

## Nullity Table

```{r appendix-nullity-table, echo=TRUE}
head(nullity.table, 40)
```

## Polynomial Score Table

```{r appendix-polynomial-score-table, echo=TRUE}
head(score.table, 40)
```

## Local Dimension Diagnostic Table

```{r appendix-local-dimension-table, echo=TRUE}
head(local.dimension.table, 60)
```

## Local Dimension Vote Table

```{r appendix-local-dimension-vote-table, echo=TRUE}
head(local.dimension.vote.table, 80)
```

## Reproducibility Commands

```{r appendix-commands, echo=TRUE, eval=FALSE}
# Smoke run
GFLOW_GEO_ANNIHILATOR_MODE=smoke \
  Rscript dev/graph-trend-filtering/render_geodesic_annihilator_experiments.R

# Full run
GFLOW_GEO_ANNIHILATOR_MODE=full \
  Rscript dev/graph-trend-filtering/render_geodesic_annihilator_experiments.R
```
