---
title: "Geodesic Skeleton Rank-Aware Cross-Enrichment Calibration"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.8,
  fig.height = 6.4,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = "ragg_png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "geodesic_skeleton_rank_calibration.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_GEO_SKELETON_RANK_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("geodesic_skeleton_rank_calibration_", mode, ".rds"))
diagnostics <- readRDS(rds.path)
summary.table <- diagnostics$summary
direction.table <- diagnostics$direction %||% data.frame()
candidate.table <- diagnostics$connector.candidates %||% data.frame()
skeleton.table <- diagnostics$skeleton %||% data.frame()
coverage.table <- diagnostics$coverage %||% data.frame()
global.candidate.table <- diagnostics$global.candidates %||% data.frame()
global.selection.table <- diagnostics$global.selection %||% data.frame()

ok.summary <- subset(summary.table, status == "ok")
rank.summary <- subset(ok.summary, row.family == "rank.cross")
global.summary <- subset(ok.summary, row.family %in% c("global.rank.cross",
                                                       "global.rank.rich"))
if (nrow(ok.summary) > 0L) {
  ok.summary$nullity.gap <- abs(ok.summary$nullity.1e.6 - ok.summary$expected.nullity)
}
if (nrow(rank.summary) > 0L) {
  rank.summary$nullity.gap <- abs(rank.summary$nullity.1e.6 - rank.summary$expected.nullity)
}
if (nrow(global.summary) > 0L) {
  global.summary$nullity.gap <- abs(global.summary$nullity.1e.6 - global.summary$expected.nullity)
}
path.ok <- subset(ok.summary, case.id == "path")
path.exact <- nrow(path.ok) > 0L &&
  all(path.ok$nullity.1e.6 == path.ok$expected.nullity, na.rm = TRUE)
rank.best <- if (nrow(rank.summary) > 0L) {
  aggregate(
    cbind(nullity.gap, intrinsic.residual, n.rank.connectors.added,
          mean.deficit.before, mean.deficit.after) ~ case.id + degree,
    data = rank.summary,
    FUN = function(z) {
      z <- z[is.finite(z)]
      if (length(z) < 1L) NA_real_ else min(z, na.rm = TRUE)
    }
  )
} else {
  data.frame()
}

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}

label.family <- function(x) {
  x <- gsub("global.rank.rich", "global-rank\nrich", x)
  x <- gsub("global.rank.cross", "global-rank\ncross", x)
  x <- gsub("rank.cross", "rank-aware\ncross", x)
  x <- gsub("coverage.cross", "coverage\ncross", x)
  gsub("skeleton.only", "skeleton\nonly", x)
}

row.family.levels <- c("skeleton.only", "coverage.cross", "rank.cross",
                       "global.rank.cross", "global.rank.rich")

best.rows.by.family <- function(tab) {
  if (!is.data.frame(tab) || nrow(tab) < 1L) return(data.frame())
  split.key <- interaction(tab$case.id, tab$degree, tab$row.family, drop = TRUE)
  out <- lapply(split(tab, split.key), function(x) {
    x <- x[order(x$nullity.gap, x$intrinsic.residual,
                 -x$n.rank.connectors.added), , drop = FALSE]
    x[1L, , drop = FALSE]
  })
  do.call(rbind, out)
}
```

# Purpose

This calibration report tests the rank-aware skeleton cross-enrichment rule on
controlled graphs where the expected local dimension is known. For each local
disk, the rule estimates an intrinsic dimension \(m_v\), measures the rank
\(r_v\) of the skeleton directions already present in that disk, and adds
connector geodesics when

\[
\mathrm{deficit}(v)=(m_v-r_v)_+>0.
\]

The goal is not to pick final defaults yet. The goal is to learn which control
settings produce connector additions, which settings overconstrain the graph,
and which settings preserve exact one-dimensional path behavior.

# Calibration Cases

The smoke calibration uses a path graph, a grid graph, a sampled paraboloid
graph, and a cube lattice. These cover expected local dimensions \(1,2,2,3\).
The full calibration adds cycle, nonuniform grid, and sampled square cases.

```{r case-overview, fig.height=5.2}
case.plot <- unique(summary.table[, c("case.id", "family", "graph.type",
                                      "n", "true.dimension")])
ggplot(case.plot,
       aes(x = reorder(case.id, true.dimension), y = n,
           fill = factor(true.dimension))) +
  geom_col(width = 0.7) +
  coord_flip() +
  labs(
    title = "Controlled graph cases for rank calibration",
    subtitle = "Fill color is the expected intrinsic dimension used as a calibration target.",
    x = NULL,
    y = "vertices",
    fill = "true dimension"
  ) +
  base.theme()
```

These cases intentionally mix idealized graphs and sampled geometric graphs.
The idealized path and grid expose exact structural behavior; the sampled
paraboloid and cube lattice test whether the rank rule remains meaningful when
graph geometry is less perfectly regular.

# Direction Deficit

The first diagnostic asks whether the rank-aware rule sees missing directions
and whether accepted connectors reduce the deficit

\[
(m_v-r_v)_+.
\]

```{r direction-deficit, fig.height=7.2}
if (nrow(direction.table) > 0L) {
  deficit.long <- rbind(
    transform(direction.table,
              stage = "before",
              direction.deficit = direction.deficit.before),
    transform(direction.table,
              stage = "after",
              direction.deficit = direction.deficit.after)
  )
  deficit.long$stage <- factor(deficit.long$stage, levels = c("before", "after"))
  ggplot(deficit.long,
         aes(x = direction.deficit, fill = stage)) +
    geom_histogram(binwidth = 1, boundary = -0.5,
                   color = "white", alpha = 0.78, position = "dodge") +
    facet_grid(true.dimension ~ case.id, scales = "free_y") +
    labs(
      title = "Local direction deficit before and after rank-aware enrichment",
      subtitle = "A left shift from before to after means connectors supplied missing local directions.",
      x = "direction deficit",
      y = "local disks",
      fill = "stage"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No direction diagnostics were recorded.")
}
```

This figure is the most direct calibration check. A path graph should usually
have zero deficit. Two- and three-dimensional graphs should show deficits when
the base skeleton is too sparse, and those deficits should decrease after
rank-aware connectors are added.

# Connector Additions

The connector rule is deterministic. Candidate connectors are scored using rank
gain, coverage gain, metric length, and a canonical path key. The next plot
shows how often connectors are accepted across the parameter sweep.

```{r connector-additions, fig.height=7.0}
if (nrow(rank.summary) > 0L) {
  ggplot(rank.summary,
         aes(x = factor(skeleton.max.cross.paths),
             y = n.rank.connectors.added,
             color = factor(disk.k),
             group = interaction(disk.k, skeleton.max.paths,
                                 skeleton.cross.iterations,
                                 skeleton.cross.rank.tol,
                                 skeleton.cross.length.penalty))) +
    geom_point(alpha = 0.75, size = 2) +
    geom_line(alpha = 0.35) +
    facet_grid(true.dimension + case.id ~ degree + skeleton.cross.iterations,
               scales = "free_y") +
    labs(
      title = "Accepted rank-aware connectors across control settings",
      subtitle = "Columns compare polynomial degree and the number of rank-repair passes.",
      x = "connector budget",
      y = "accepted connectors",
      color = "disk k"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No rank-aware rows were recorded.")
}
```

If this plot is zero everywhere, the rank test is too permissive, the base
skeleton is already direction-saturated, or the candidate connector pool is too
restricted. If connector counts are large everywhere, the rule may be too eager
and should be checked against nullity.

```{r deficit-by-iteration, fig.height=7.0}
if (nrow(direction.table) > 0L &&
    "skeleton.cross.iteration" %in% names(direction.table)) {
  iter.summary <- aggregate(
    cbind(direction.deficit.before, direction.deficit.after, connector.added) ~
      case.id + true.dimension + degree + skeleton.max.cross.paths +
      skeleton.cross.iterations + skeleton.cross.iteration,
    data = direction.table,
    FUN = function(z) mean(z, na.rm = TRUE)
  )
  ggplot(iter.summary,
         aes(x = skeleton.cross.iteration,
             y = direction.deficit.after,
             color = factor(skeleton.max.cross.paths),
             group = interaction(skeleton.max.cross.paths,
                                 skeleton.cross.iterations))) +
    geom_point(alpha = 0.75, size = 1.8) +
    geom_line(alpha = 0.45) +
    facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
    labs(
      title = "Mean direction deficit by rank-repair iteration",
      subtitle = "Successful iterative repair should lower the after-deficit as passes accumulate.",
      x = "rank-repair iteration",
      y = "mean after-deficit",
      color = "connector budget"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No iteration-level direction diagnostics were recorded.")
}
```

This figure separates connector budget from repeated local repair. If later
iterations do not lower the deficit, then additional passes are revisiting disks
that are already direction-saturated or unable to generate useful connectors.

# Nullity Calibration

For an intrinsic \(m\)-dimensional polynomial model of degree \(d\), the target
dimension is

\[
\dim \mathcal P_{\le d}(\mathbb R^m)=\binom{m+d}{m}.
\]

This is not a proof target for every graph, but it is the calibration reference
for detecting severe under- or over-constraint.

```{r nullity-calibration, fig.height=7.2}
nullity.plot <- transform(
  ok.summary,
  row.family.label = factor(label.family(row.family),
                            levels = label.family(row.family.levels))
)
ggplot(nullity.plot,
       aes(x = row.family.label, y = nullity.1e.6, fill = row.family.label)) +
  geom_boxplot(outlier.alpha = 0.35) +
  geom_hline(aes(yintercept = expected.nullity),
             linetype = "dashed", color = "grey25") +
  facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
  labs(
    title = "Estimated nullity versus polynomial dimension reference",
    subtitle = "Dashed line is choose(true dimension + degree, true dimension).",
    x = "row family",
    y = "estimated nullity at relative tolerance 1e-6",
    fill = "row family"
  ) +
  base.theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

High nullity means the skeleton remains underconstrained. Nullity far below the
reference suggests the operator may be destroying the intended polynomial-like
null space. The best calibration region should reduce the skeleton-only nullity
without collapsing to a trivial null space.

# Polynomial Residuals

The next diagnostic evaluates whether the operator approximately annihilates
known intrinsic coordinate polynomials:

\[
\frac{\|AP_d\|_F}{\|P_d\|_F}.
\]

The control residual uses exact degree \(d+1\) monomials, which should generally
be larger.

```{r polynomial-residuals, fig.height=7.2}
residual.plot <- rbind(
  transform(ok.summary, target = "degree <= d",
            residual = intrinsic.residual),
  transform(ok.summary, target = "exact degree d+1",
            residual = control.residual)
)
residual.plot$row.family.label <- factor(
  label.family(residual.plot$row.family),
  levels = label.family(row.family.levels)
)
residual.plot$target <- factor(residual.plot$target,
                               levels = c("degree <= d", "exact degree d+1"))
ggplot(residual.plot,
       aes(x = row.family.label, y = residual, fill = target)) +
  geom_boxplot(outlier.alpha = 0.35, position = position_dodge(width = 0.75)) +
  scale_y_log10() +
  facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
  labs(
    title = "Polynomial residuals by row family",
    subtitle = "Useful settings keep degree <= d residuals small while retaining discrimination against degree d+1 controls.",
    x = "row family",
    y = "residual ratio, log10 scale",
    fill = "polynomial target"
  ) +
  base.theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

The residual plot guards against a deceptive nullity result. A small nullity is
not useful if the operator no longer annihilates the intended polynomial probes.

# Global Rank Selection

The local rank-aware rule asks whether a connector improves the rank inside one
disk. The global-rank selector asks the complementary question: does the same
candidate add a linearly independent constraint to the assembled operator?
At each step it evaluates

\[
\Delta_{\mathrm{rank}}(\pi)
=
\mathrm{rank}\{[A;A_\pi]\}-\mathrm{rank}(A),
\]

and accepts only candidates with positive global rank gain and acceptable
polynomial residual. This is a deliberately expensive calibration tool, not a
proposed production solver.

The `global-rank cross` row family uses only the original skeleton-membership
connector pool. The `global-rank rich` row family uses the same selector but
adds radial, boundary-to-boundary, and annulus-to-boundary local geodesics to
the candidate pool.

```{r global-rank-trajectory, fig.height=7.0}
if (nrow(global.selection.table) > 0L) {
  global.selection.table$expected.nullity <- choose(
    global.selection.table$degree + global.selection.table$true.dimension,
    global.selection.table$true.dimension
  )
  global.selection.table$nullity.after <- global.selection.table$n -
    global.selection.table$rank.after
  ggplot(global.selection.table,
         aes(x = step, y = nullity.after,
             color = row.family,
             group = interaction(case.id, degree, skeleton.max.paths,
                                 disk.k, skeleton.max.cross.paths,
                                 row.family))) +
    geom_hline(aes(yintercept = expected.nullity),
               color = "grey35", linetype = "dashed") +
    geom_point(alpha = 0.82, size = 1.8) +
    geom_line(alpha = 0.42) +
    facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
    labs(
      title = "Global-rank selector trajectory",
      subtitle = "Dashed lines are polynomial-dimension references; each point is one accepted connector.",
      x = "accepted connector step",
      y = "global nullity after accepted connector",
      color = "row family"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No global-rank connector selections were recorded.")
}
```

This figure tests whether the global selector can make monotone progress toward
the target nullity. A flat trajectory means that the local candidate pool lacks
globally independent rows or that the residual guardrail rejects the useful
candidates.

```{r global-rank-candidates, fig.height=6.8}
if (nrow(global.candidate.table) > 0L) {
  global.candidate.table$selected.label <- ifelse(
    global.candidate.table$selected, "selected", "not selected"
  )
  ggplot(global.candidate.table,
         aes(x = global.rank.gain, y = polynomial.residual,
             color = selected.label,
             shape = candidate.type)) +
    geom_jitter(width = 0.08, height = 0, alpha = 0.55, size = 1.4) +
    scale_y_log10() +
    facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
    labs(
      title = "Global-rank candidate pool",
      subtitle = "Useful candidates have positive global rank gain while preserving polynomial annihilation.",
      x = "global rank gain",
      y = "candidate polynomial residual, log10 scale",
      color = "selection",
      shape = "candidate type"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No global-rank candidates were scored.")
}
```

The candidate-pool view is a debugging plot. If most candidates have zero global
rank gain, the path generator is producing redundant rows. If many positive-gain
candidates have large residuals, the generator is finding constraints that
increase rank by damaging the polynomial null space.

# Parameter Sensitivity

The rank-aware rule has several tuning controls. The next figure summarizes how
nullity and intrinsic residual vary with the most important ones: connector
budget, repair iterations, base skeleton size, disk radius, and length penalty.

```{r parameter-sensitivity, fig.height=7.0}
if (nrow(rank.summary) > 0L) {
  sens <- transform(
    rank.summary,
    length.penalty.label = paste0("len=", skeleton.cross.length.penalty),
    iteration.label = paste0("iter=", skeleton.cross.iterations)
  )
  ggplot(sens,
         aes(x = n.rank.connectors.added, y = nullity.gap,
             color = factor(skeleton.max.cross.paths),
             shape = factor(skeleton.max.paths))) +
    geom_point(alpha = 0.82, size = 2.1) +
    facet_grid(length.penalty.label ~ iteration.label) +
    labs(
      title = "Rank-aware connector count versus nullity gap",
      subtitle = "Good settings move toward lower nullity gap without requiring pathological row counts.",
      x = "accepted rank-aware connectors",
      y = "absolute nullity gap",
      color = "connector budget",
      shape = "base skeleton paths"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No rank-aware rows were recorded.")
}
```

Good settings should move left-to-right enough to reduce nullity, but not so far
that nullity collapses below the polynomial reference or residuals deteriorate.

```{r residual-versus-budget, fig.height=7.0}
if (nrow(rank.summary) > 0L) {
  ggplot(rank.summary,
         aes(x = factor(skeleton.max.cross.paths),
             y = intrinsic.residual,
             color = factor(skeleton.cross.iterations),
             group = interaction(skeleton.cross.iterations,
                                 skeleton.max.paths,
                                 disk.k,
                                 skeleton.cross.length.penalty))) +
    geom_point(alpha = 0.75, size = 1.9) +
    geom_line(alpha = 0.28) +
    scale_y_log10() +
    facet_grid(true.dimension + case.id ~ degree, scales = "free_y") +
    labs(
      title = "Intrinsic polynomial residual versus connector budget",
      subtitle = "Connector budgets are useful only if polynomial annihilation remains stable.",
      x = "connector budget",
      y = "intrinsic residual, log10 scale",
      color = "iterations"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No rank-aware rows were recorded.")
}
```

This plot checks the main failure mode of aggressive repair: nullity may improve
because the operator is adding many constraints, while the intended polynomial
annihilation property quietly worsens.

# What We Learned

```{r learned-table, echo=FALSE, results='asis'}
cat("The path graph remains the safety check. In this smoke calibration, path exactness was ",
    if (path.exact) "**preserved**" else "**not preserved**",
    ": the estimated nullity matched the 1D polynomial reference for every tested path setting.\\n\\n",
    sep = "")
if (nrow(rank.summary) > 0L) {
  med <- aggregate(
    cbind(n.rank.connectors.added, mean.deficit.before, mean.deficit.after,
          nullity.gap, intrinsic.residual) ~ case.id + true.dimension + degree,
    data = rank.summary,
    FUN = function(z) round(stats::median(z, na.rm = TRUE), 5)
  )
  cat("Median rank-aware behavior by case is summarized below. Connector additions and ",
      "deficit reduction show whether the local rank-repair mechanism is active; ",
      "nullity gap shows whether that local repair is enough globally.\\n\\n",
      sep = "")
  print(knitr::kable(med))
}
best.family <- best.rows.by.family(ok.summary)
if (nrow(best.family) > 0L) {
  report.best <- best.family[, c(
    "case.id", "true.dimension", "degree", "row.family", "nullity.gap",
    "nullity.1e.6", "expected.nullity", "intrinsic.residual",
    "n.rank.connectors.added"
  )]
  report.best$row.family <- label.family(report.best$row.family)
  report.best$intrinsic.residual <- signif(report.best$intrinsic.residual, 4)
  cat("\\n\\nBest observed setting within each row family is below. This table is ",
      "the compact check of whether global-rank selection improves the best ",
      "achievable nullity gap without increasing polynomial residuals.\\n\\n",
      sep = "")
  print(knitr::kable(report.best))
}
if (nrow(global.summary) > 0L) {
  global.med <- aggregate(
    cbind(global.rank.gain, global.nullity.initial, global.nullity.final,
          nullity.gap, intrinsic.residual, n.global.candidates.scored) ~
      case.id + true.dimension + degree,
    data = global.summary,
    FUN = function(z) round(stats::median(z, na.rm = TRUE), 5)
  )
  cat("\\n\\nMedian global-rank behavior is below. The important columns are ",
      "`global.rank.gain`, which measures independent rows actually added, ",
      "and the movement from initial to final global nullity.\\n\\n",
      sep = "")
  print(knitr::kable(global.med))
}
if (all(c("global.rank.cross", "global.rank.rich") %in% ok.summary$row.family)) {
  global.best <- best.rows.by.family(subset(
    ok.summary,
    row.family %in% c("global.rank.cross", "global.rank.rich")
  ))
  wide <- reshape(
    global.best[, c("case.id", "degree", "row.family", "nullity.gap",
                    "intrinsic.residual", "global.rank.gain")],
    idvar = c("case.id", "degree"),
    timevar = "row.family",
    direction = "wide"
  )
  if (nrow(wide) > 0L &&
      all(c("nullity.gap.global.rank.cross",
            "nullity.gap.global.rank.rich") %in% names(wide))) {
    wide$gap.improvement <- wide$nullity.gap.global.rank.cross -
      wide$nullity.gap.global.rank.rich
    cat("\\n\\nLanding 1 comparison: positive `gap.improvement` means the richer ",
        "candidate pool improved the best observed global-rank nullity gap ",
        "relative to the original skeleton-membership pool, using the same ",
        "selector.\\n\\n",
        sep = "")
    print(knitr::kable(wide[, c("case.id", "degree",
                                "nullity.gap.global.rank.cross",
                                "nullity.gap.global.rank.rich",
                                "gap.improvement")]))
  }
}
```

The main Landing 1 lesson is that candidate diversity was a real bottleneck.
With the selector held fixed, adding radial, boundary-to-boundary, and
annulus-to-boundary candidates gives the global-rank selector many more
independent rows to choose from. This is exactly the behavior we wanted to test:
the improvement comes from a richer pool, not from changing the selection rule.

The second lesson is that polynomial residuals are the guardrail. In the smoke
calibration, aggressive connector settings often kept intrinsic residuals small,
which means the new rows are not obviously destroying the polynomial probes.
This supports continuing with rank-aware enrichment, but only in tandem with
nullity-gap and residual diagnostics.

The third lesson is that 3D remains the hard case. The cube lattice accepts many
rank-aware connectors, but its nullity stays well above the polynomial reference
for the tested budgets. The global-rank selector directly tests whether the
available connector pool contains independent constraints; when it stalls above
the reference, the bottleneck is no longer just ranking accepted paths but the
geometric diversity of candidate paths.

# Results Summary And Discussion

This calibration separates three questions that were mixed together in the
first smoke report. First, does the local dimension/rank diagnostic actually
produce deficits? Second, do accepted connectors reduce those deficits? Third,
does the resulting operator have plausible nullity and polynomial residuals?

The settings recommended by this report should satisfy all three conditions:
visible but not excessive connector additions, lower direction deficits after
enrichment, and nullity closer to the polynomial-dimension reference without
large increases in \(\|AP_d\|_F/\|P_d\|_F\). Settings that only improve one of
these diagnostics are not yet good defaults.

# Appendix

## Input Artifact

```{r appendix-input, echo=TRUE}
rds.path
diagnostics$created
```

## Summary Table

```{r appendix-summary, echo=TRUE}
head(summary.table, 80)
```

## Direction Diagnostics

```{r appendix-direction, echo=TRUE}
head(direction.table, 80)
```

## Connector Candidates

```{r appendix-candidates, echo=TRUE}
head(candidate.table, 80)
```

## Global Rank Candidates

```{r appendix-global-candidates, echo=TRUE}
head(global.candidate.table, 80)
```

## Global Rank Selections

```{r appendix-global-selection, echo=TRUE}
head(global.selection.table, 80)
```

## Reproducibility

```{bash, eval=FALSE}
GFLOW_GEO_SKELETON_RANK_MODE=smoke \
  Rscript dev/graph-trend-filtering/render_geodesic_skeleton_rank_calibration.R
```
