---
title: "Geodesic Skeleton Row-Family Comparison"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.5,
  fig.height = 6.2,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = "ragg_png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "geodesic_annihilator_experiments.R"))
library(ggplot2)

requested.mode <- Sys.getenv("GFLOW_GEO_ANNIHILATOR_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("geodesic_annihilator_", requested.mode, ".rds"))
if (!file.exists(rds.path)) {
  rds.path <- file.path(report.dir, "geodesic_annihilator_smoke.rds")
}
diagnostics <- readRDS(rds.path)
summary.table <- diagnostics$results
path.table <- diagnostics$path.table
vertex.table <- diagnostics$vertex.table
skeleton.table <- diagnostics$skeleton.table %||% data.frame()
skeleton.coverage.table <- diagnostics$skeleton.coverage.table %||% data.frame()
skeleton.direction.table <- diagnostics$skeleton.direction.table %||% data.frame()
skeleton.connector.candidate.table <- diagnostics$skeleton.connector.candidate.table %||% data.frame()

if (!"path.family" %in% names(summary.table)) {
  summary.table$path.family <- "local.composite.geodesics"
}
if (!"path.family" %in% names(path.table)) {
  path.table$path.family <- "local.composite.geodesics"
}
if (!"path.family" %in% names(vertex.table)) {
  vertex.table$path.family <- "local.composite.geodesics"
}
if (!"skeleton.cross.rule" %in% names(summary.table)) {
  summary.table$skeleton.cross.rule <- "coverage"
}
if (!"skeleton.cross.rule" %in% names(path.table)) {
  path.table$skeleton.cross.rule <- "coverage"
}
if (!"skeleton.cross.rule" %in% names(vertex.table)) {
  vertex.table$skeleton.cross.rule <- "coverage"
}
if (nrow(skeleton.coverage.table) < 1L && all(c("skeleton.coverage", "skeleton.distance") %in% names(vertex.table))) {
  skeleton.coverage.table <- subset(
    vertex.table,
    path.family != "local.composite.geodesics",
    select = c(run.id, dataset, n, seed, graph.type, path.family, degree,
               center, skeleton.coverage, skeleton.distance, n.operator.rows)
  )
  names(skeleton.coverage.table)[names(skeleton.coverage.table) == "center"] <- "vertex"
  names(skeleton.coverage.table)[names(skeleton.coverage.table) == "n.operator.rows"] <- "row.coverage"
}
if (nrow(skeleton.coverage.table) > 0L && !"skeleton.cross.rule" %in% names(skeleton.coverage.table)) {
  skeleton.coverage.table$skeleton.cross.rule <- "coverage"
}
if (nrow(skeleton.table) > 0L && !"skeleton.cross.rule" %in% names(skeleton.table)) {
  skeleton.table$skeleton.cross.rule <- "coverage"
}

wrap.graph <- function(x) {
  gsub("delaunay.1skeleton", "Delaunay\n1-skeleton",
       gsub("symmetric.knn", "symmetric\nkNN", x))
}
wrap.path.family <- function(x) {
  x <- gsub("geodesic.skeleton.cross.enriched", "skeleton\n+ cross", x)
  x <- gsub("geodesic.skeleton", "skeleton", x)
  gsub("local.composite.geodesics", "local\ncomposite", x)
}
wrap.cross.rule <- function(x) {
  x <- gsub("local.dimension.rank", "local dimension\nrank", x)
  gsub("coverage", "coverage", x)
}
base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}

make.path.validation.graph <- function(n = 14L) {
  edges <- cbind(seq_len(n - 1L), seq_len(n - 1L) + 1L)
  make.graph.from.edges(n, edges, rep(1, nrow(edges)))
}

make.star.validation.graph <- function(arms = 3L, arm.length = 5L) {
  n <- 1L + arms * arm.length
  edges <- matrix(NA_integer_, nrow = arms * arm.length, ncol = 2L)
  row <- 0L
  for (arm in seq_len(arms)) {
    first <- 2L + (arm - 1L) * arm.length
    row <- row + 1L
    edges[row, ] <- c(1L, first)
    if (arm.length > 1L) {
      for (step in seq_len(arm.length - 1L)) {
        row <- row + 1L
        edges[row, ] <- c(first + step - 1L, first + step)
      }
    }
  }
  make.graph.from.edges(n, edges, rep(1, nrow(edges)))
}

make.grid.validation.graph <- function(nx = 6L, ny = 6L) {
  id <- matrix(seq_len(nx * ny), nrow = nx, ncol = ny)
  edges <- matrix(NA_integer_, nrow = (nx - 1L) * ny + nx * (ny - 1L), ncol = 2L)
  row <- 0L
  for (i in seq_len(nx)) {
    for (j in seq_len(ny)) {
      if (i < nx) {
        row <- row + 1L
        edges[row, ] <- c(id[i, j], id[i + 1L, j])
      }
      if (j < ny) {
        row <- row + 1L
        edges[row, ] <- c(id[i, j], id[i, j + 1L])
      }
    }
  }
  make.graph.from.edges(nx * ny, edges, rep(1, nrow(edges)))
}

validation.one <- function(graph.name, graph, family, degree, ...) {
  op <- build.geodesic.annihilator.operator(
    graph$adj.list,
    graph$weight.list,
    degree = degree,
    path.family = family,
    row.normalization = "unit.l2",
    vertex.normalization = "none",
    disk.k = 8L,
    ...
  )
  nullity <- estimate.nullity(op$A)$table
  data.frame(
    graph = graph.name,
    n = length(graph$adj.list),
    degree = degree,
    path.family = family,
    n.operator.rows = nrow(op$A),
    nullity.1e.6 = nullity$nullity[nullity$rel.tol == 1e-6],
    n.skeleton.paths = if (!is.null(op$skeleton.table)) nrow(op$skeleton.table) else NA_integer_,
    skeleton.cross.rule = if (!is.null(op$control$skeleton.cross.rule)) {
      op$control$skeleton.cross.rule
    } else {
      "coverage"
    },
    min.coverage = if (!is.null(op$skeleton.coverage)) min(op$skeleton.coverage$skeleton.coverage) else NA_real_,
    median.coverage = if (!is.null(op$skeleton.coverage)) stats::median(op$skeleton.coverage$skeleton.coverage) else NA_real_,
    max.coverage = if (!is.null(op$skeleton.coverage)) max(op$skeleton.coverage$skeleton.coverage) else NA_real_,
    stringsAsFactors = FALSE
  )
}

path.graph <- make.path.validation.graph()
star.graph <- make.star.validation.graph()
grid.graph <- make.grid.validation.graph()
validation.table <- do.call(rbind, lapply(1:2, function(degree) {
  rbind(
    validation.one(
      "path diameter only", path.graph, "geodesic.skeleton", degree,
      max.skeleton.paths = 1L, skeleton.min.coverage = 1L,
      skeleton.max.passes = 1L, skeleton.max.cross.paths = 0L
    ),
    validation.one(
      "star skeleton", star.graph, "geodesic.skeleton", degree,
      max.skeleton.paths = 20L, skeleton.min.coverage = 2L,
      skeleton.max.passes = 2L, skeleton.max.cross.paths = 0L
    ),
    validation.one(
      "grid skeleton", grid.graph, "geodesic.skeleton", degree,
      max.skeleton.paths = 35L, skeleton.min.coverage = 2L,
      skeleton.max.passes = 2L, skeleton.max.cross.paths = 0L
    ),
    validation.one(
      "grid skeleton + cross", grid.graph,
      "geodesic.skeleton.cross.enriched", degree,
      max.skeleton.paths = 35L, skeleton.min.coverage = 2L,
      skeleton.max.passes = 2L, skeleton.max.cross.paths = 12L
    ),
    validation.one(
      "grid skeleton + rank cross", grid.graph,
      "geodesic.skeleton.cross.enriched", degree,
      max.skeleton.paths = 8L, skeleton.min.coverage = 1L,
      skeleton.max.passes = 2L, skeleton.max.cross.paths = 12L,
      skeleton.cross.rule = "local.dimension.rank"
    )
  )
}))
validation.table$expected.path.nullity <- ifelse(
  validation.table$graph == "path diameter only",
  validation.table$degree + 1L,
  NA_integer_
)
```

# Purpose

This report isolates the new geodesic-skeleton row family from the denser local
composite-geodesic operator. The skeleton construction starts with a graph
diameter path, adds geodesics from poorly covered vertices back to the current
skeleton, and optionally adds local cross-connectors between nearby skeleton
paths. Along each selected geodesic, the operator places ordinary one-dimensional
sliding-window annihilator rows.

The central diagnostic is coverage multiplicity,

\[
c(v)=\#\{\gamma_s: v\in\gamma_s\},
\]

together with the resulting operator size and nullity. A useful skeleton should
be much sparser than the local composite operator while still preserving enough
coverage to constrain the graph.

# Deterministic Micro-Validation

The first check uses graphs where the expected behavior is visible by inspection:
a path graph, a three-arm star, and a rectangular grid. On the path graph, a
diameter-only skeleton should reproduce the usual 1D sliding windows. For
annihilator degree \(d\), the expected path nullity is

\[
\dim\ker(A)=d+1.
\]

The star and grid tests do not have a single obvious polynomial nullity target;
they are coverage tests. The star checks branch coverage, while the grid checks
whether cross-enrichment raises coverage beyond a tree-like skeleton.

```{r validation-summary-plot, fig.height=6.4}
validation.plot <- transform(
  validation.table,
  graph = factor(graph, levels = c(
    "path diameter only", "star skeleton", "grid skeleton",
    "grid skeleton + cross", "grid skeleton + rank cross"
  )),
  degree = factor(degree)
)
ggplot(validation.plot,
       aes(x = graph, y = nullity.1e.6, fill = degree)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68) +
  geom_point(aes(y = expected.path.nullity, group = degree),
             position = position_dodge(width = 0.75),
             shape = 4, size = 3, stroke = 1.1, na.rm = TRUE) +
  coord_flip() +
  labs(
    title = "Nullity on deterministic skeleton validation graphs",
    subtitle = "Crosses mark the exact 1D path-graph target d + 1.",
    x = NULL,
    y = "estimated nullity at relative tolerance 1e-6",
    fill = "degree"
  ) +
  base.theme()
```

The path graph is the strict parity check: if that bar misses the cross, the
skeleton row generator is not reproducing ordinary one-dimensional sliding
windows. The star and grid values should be read as structural diagnostics. A
large nullity means the skeleton is sparse and leaves many functions
unconstrained; a smaller nullity after cross-enrichment means connector paths
are adding real constraints.

```{r validation-coverage-plot, fig.height=6.4}
ggplot(validation.plot,
       aes(x = graph, y = median.coverage, fill = degree)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.68) +
  geom_errorbar(
    aes(ymin = min.coverage, ymax = max.coverage, group = degree),
    position = position_dodge(width = 0.75),
    width = 0.22
  ) +
  coord_flip() +
  labs(
    title = "Coverage multiplicity on deterministic skeleton validation graphs",
    subtitle = "Bars show median vertex coverage; whiskers show min and max coverage.",
    x = NULL,
    y = "skeleton paths containing a vertex",
    fill = "degree"
  ) +
  base.theme()
```

Coverage is the first thing to inspect before interpreting nullity. On path
graphs, one diameter path is enough. On branching or grid-like graphs, low
coverage identifies where a purely tree-like skeleton may underconstrain
transverse behavior.

# Skeleton Paths In Sampled Graphs

The experiment RDS records each selected skeleton path and its role. The first
path is the diameter. Coverage passes add geodesics from poorly covered vertices
to the skeleton, and the cross-enriched family adds local connector paths.

```{r skeleton-path-role-plot, fig.height=6.6}
if (nrow(skeleton.table) > 0L) {
  skeleton.role.plot <- transform(
    skeleton.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    path.family.label = factor(wrap.path.family(path.family),
                               levels = wrap.path.family(c(
                                 "geodesic.skeleton",
                                 "geodesic.skeleton.cross.enriched"
                               ))),
    cross.rule.label = factor(wrap.cross.rule(skeleton.cross.rule),
                              levels = wrap.cross.rule(c(
                                "coverage", "local.dimension.rank"
                              )))
  )
  ggplot(skeleton.role.plot,
         aes(x = skeleton.role, fill = cross.rule.label)) +
    geom_bar(color = "white", alpha = 0.86, position = "dodge") +
    facet_grid(dataset + path.family.label ~ graph.label, scales = "free_y") +
    labs(
      title = "Selected skeleton paths by construction role",
      subtitle = "Cross-enriched runs add local connector geodesics after the coverage passes.",
      x = "skeleton path role",
      y = "number of selected paths",
      fill = "cross rule"
    ) +
    base.theme() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  plot.new()
  text(0.5, 0.5, "No skeleton path table was recorded.")
}
```

This plot tells us whether the skeleton is dominated by the initial diameter,
coverage paths, or cross-connectors. For two- and three-dimensional sampled
graphs, a report with only diameter and coverage paths is usually too
tree-like; connector paths are the mechanism that begins to add transverse
constraints.

```{r skeleton-path-length-plot, fig.height=6.6}
if (nrow(skeleton.table) > 0L) {
  ggplot(skeleton.role.plot,
         aes(x = path.length, fill = skeleton.role)) +
    geom_histogram(bins = 26, color = "white", alpha = 0.84, position = "identity") +
    facet_grid(graph.label ~ path.family.label + cross.rule.label, scales = "free_y") +
    labs(
      title = "Metric lengths of selected skeleton geodesics",
      subtitle = "Diameter paths are long by construction; connector paths should be shorter and more local.",
      x = "geodesic path length",
      y = "number of skeleton paths",
      fill = "role"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No skeleton path table was recorded.")
}
```

The intended pattern is long diameter paths, medium coverage paths, and shorter
local cross-connectors. If connector paths are as long as the diameter, the
cross-enrichment rule is no longer local and should be tightened.

# Coverage In Sampled Graphs

The next diagnostic looks directly at vertex coverage. This is more important
than the number of paths, because a few long paths can give high path counts
while still leaving many vertices poorly constrained.

```{r sampled-coverage-plot, fig.height=6.8}
if (nrow(skeleton.coverage.table) > 0L) {
  coverage.plot <- transform(
    skeleton.coverage.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    path.family.label = factor(wrap.path.family(path.family),
                               levels = wrap.path.family(c(
                                 "geodesic.skeleton",
                                 "geodesic.skeleton.cross.enriched"
                               ))),
    cross.rule.label = factor(wrap.cross.rule(skeleton.cross.rule),
                              levels = wrap.cross.rule(c(
                                "coverage", "local.dimension.rank"
                              )))
  )
  ggplot(coverage.plot,
         aes(x = skeleton.coverage, fill = dataset)) +
    geom_histogram(binwidth = 1, boundary = -0.5, color = "white", alpha = 0.84) +
    facet_grid(graph.label ~ path.family.label + cross.rule.label, scales = "free_y") +
    labs(
      title = "Vertex coverage multiplicity in sampled graphs",
      subtitle = "The cross-enriched family should shift mass toward higher coverage.",
      x = "skeleton paths containing vertex",
      y = "number of vertices",
      fill = "dataset"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No skeleton coverage table was recorded.")
}
```

The current skeleton rule is deterministic and sparse. Its main failure mode is
therefore visible here: vertices with low coverage or zero row coverage. These
regions are the natural targets for the next cross-enrichment and thinning
experiments.

# Local Direction-Rank Enrichment

The rank-aware cross rule estimates a local disk dimension \(m_v\), embeds the
disk in local coordinates, and measures how many independent skeleton directions
are already present. If the local skeleton directions have rank \(r_v<m_v\), the
algorithm searches for connector geodesics between distinct skeleton paths and
prefers connectors that increase directional rank:

\[
\mathrm{deficit}(v)=(m_v-r_v)_+.
\]

This diagnostic shows whether the new rule is actually finding local rank
deficits and whether accepted connectors reduce those deficits.

```{r direction-rank-deficit, fig.height=6.8}
if (nrow(skeleton.direction.table) > 0L) {
  direction.plot <- transform(
    skeleton.direction.table,
    graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
      "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
    ))),
    cross.rule.label = factor(wrap.cross.rule(skeleton.cross.rule),
                              levels = wrap.cross.rule(c(
                                "coverage", "local.dimension.rank"
                              )))
  )
  deficit.long <- rbind(
    transform(direction.plot,
              deficit.stage = "before",
              direction.deficit = direction.deficit.before),
    transform(direction.plot,
              deficit.stage = "after",
              direction.deficit = direction.deficit.after)
  )
  deficit.long$deficit.stage <- factor(deficit.long$deficit.stage,
                                       levels = c("before", "after"))
  ggplot(deficit.long,
         aes(x = direction.deficit, fill = deficit.stage)) +
    geom_histogram(binwidth = 1, boundary = -0.5, color = "white",
                   alpha = 0.78, position = "dodge") +
    facet_grid(graph.label ~ cross.rule.label, scales = "free_y") +
    labs(
      title = "Local skeleton direction deficit before and after rank-aware connectors",
      subtitle = "Deficit is max(0, estimated local dimension minus local skeleton direction rank).",
      x = "direction deficit",
      y = "number of local disks",
      fill = "stage"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No local direction-rank diagnostics were recorded.")
}
```

If the after distribution shifts left, the rank-aware connectors are doing their
intended job. If many disks remain deficient, either the candidate connector
pool is too small or the local dimension estimate is asking for more directions
than the skeleton can provide at this radius.

```{r rank-aware-connector-gains, fig.height=6.4}
if (nrow(skeleton.direction.table) > 0L) {
  added.plot <- subset(direction.plot, connector.added)
  if (nrow(added.plot) > 0L) {
    ggplot(added.plot,
           aes(x = rank.gain, y = coverage.gain, color = graph.label)) +
      geom_jitter(width = 0.05, height = 0.05, alpha = 0.75, size = 2) +
      facet_grid(degree ~ cross.rule.label) +
      labs(
        title = "Accepted connector rank gain versus coverage gain",
        subtitle = "Rank-aware connectors are selected by deterministic rank, coverage, length, and path-key scoring.",
        x = "direction-rank gain",
        y = "coverage gain",
        color = "graph"
      ) +
      base.theme()
  } else {
    plot.new()
    text(0.5, 0.5, "No rank-aware connectors were accepted.")
  }
} else {
  plot.new()
  text(0.5, 0.5, "No local direction-rank diagnostics were recorded.")
}
```

Rank gain is the primary criterion. Coverage gain is still useful because a
connector that spans a missing direction but never reaches poorly covered
vertices may have little effect on the final row operator.

# Operator Size And Nullity

The skeleton family is meant to be a sparse comparator, not a drop-in
replacement for the local composite-geodesic operator. The plot below compares
operator row counts and nullities across row families.

```{r operator-size-nullity-plot, fig.height=6.8}
operator.plot <- subset(summary.table, status == "ok")
operator.plot <- transform(
  operator.plot,
  graph.label = factor(wrap.graph(graph.type), levels = wrap.graph(c(
    "delaunay.1skeleton", "symmetric.knn", "radius", "cknn"
  ))),
  path.family.label = factor(wrap.path.family(path.family),
                             levels = wrap.path.family(c(
                               "local.composite.geodesics",
                               "geodesic.skeleton",
                               "geodesic.skeleton.cross.enriched"
                             ))),
  cross.rule.label = factor(wrap.cross.rule(skeleton.cross.rule),
                            levels = wrap.cross.rule(c(
                              "coverage", "local.dimension.rank"
                            )))
)
operator.plot$row.family.label <- ifelse(
  operator.plot$path.family == "geodesic.skeleton.cross.enriched",
  paste(operator.plot$path.family.label, operator.plot$cross.rule.label, sep = "\n"),
  as.character(operator.plot$path.family.label)
)
operator.plot$row.family.label <- factor(
  operator.plot$row.family.label,
  levels = c("local\ncomposite", "skeleton",
             "skeleton\n+ cross\ncoverage",
             "skeleton\n+ cross\nlocal dimension\nrank")
)
ggplot(operator.plot,
       aes(x = row.family.label, y = n.operator.rows, fill = row.family.label)) +
  geom_boxplot(outlier.alpha = 0.35) +
  scale_y_log10() +
  facet_grid(degree ~ graph.label, scales = "free_y") +
  labs(
    title = "Operator size by row-generation family",
    subtitle = "Skeleton rows are intentionally much sparser than local composite-geodesic rows.",
    x = "row family",
    y = "operator rows, log10 scale",
    fill = "row family"
  ) +
  base.theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

The useful regime is not simply “fewest rows.” If the skeleton operator is too
sparse, nullity stays high and the graph remains underconstrained. The next
plot asks whether cross-enrichment moves nullity in the expected direction.

```{r operator-nullity-plot, fig.height=6.8}
ggplot(operator.plot,
       aes(x = row.family.label, y = nullity.1e.6, fill = row.family.label)) +
  geom_boxplot(outlier.alpha = 0.35) +
  facet_grid(degree ~ graph.label, scales = "free_y") +
  labs(
    title = "Estimated nullity by row-generation family",
    subtitle = "Cross-enrichment should reduce skeleton nullity without collapsing the intended polynomial null space.",
    x = "row family",
    y = "estimated nullity at relative tolerance 1e-6",
    fill = "row family"
  ) +
  base.theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

High skeleton nullity means the selected paths do not yet constrain enough graph
directions. Very low nullity can mean the operator is overconstraining the graph
and destroying the desired polynomial-like null space. The next methodological
step is to use local dimension diagnostics to decide where cross-connectors are
needed, instead of adding them only from graph-only coverage proxies.

# Discussion

The skeleton implementation gives us a clean experimental comparator. The
diameter-only behavior on path graphs checks the one-dimensional sliding-window
mechanism, while the star and grid examples expose the central modeling issue:
one skeleton pass behaves like a tree approximation. Coverage-based
cross-enrichment increases coverage and reduces nullity using graph-only
heuristics.

The rank-aware cross rule is the next step toward a geometry-controlled
row-generation strategy. In a disk whose estimated intrinsic dimension is
\(m_v\), it asks whether the selected skeleton directions span approximately
\(m_v\) independent directions. When they do not, it adds connector paths that
are scored by direction-rank gain, coverage gain, length, and deterministic path
keys. The main question for subsequent experiments is whether this local rank
criterion improves polynomial residuals without collapsing the intended
polynomial-like null space.

# Appendix

## Input Artifact

```{r appendix-input, echo=TRUE}
rds.path
diagnostics$created
```

## Deterministic Validation Table

```{r appendix-validation-table, echo=TRUE}
validation.table
```

## Skeleton Path Table

```{r appendix-skeleton-table, echo=TRUE}
head(skeleton.table, 60)
```

## Skeleton Coverage Table

```{r appendix-skeleton-coverage-table, echo=TRUE}
head(skeleton.coverage.table, 60)
```

## Skeleton Direction Diagnostics Table

```{r appendix-skeleton-direction-table, echo=TRUE}
head(skeleton.direction.table, 60)
```

## Skeleton Connector Candidate Table

```{r appendix-skeleton-connector-candidate-table, echo=TRUE}
head(skeleton.connector.candidate.table, 60)
```

## Reproducibility Command

```{bash, eval=FALSE}
GFLOW_GEO_ANNIHILATOR_MODE=smoke \
  Rscript dev/graph-trend-filtering/render_geodesic_skeleton_comparison.R
```
