---
title: "Phase 4 Path-Selection Diagnostics for Path-Divided-Difference Graph Trend Filtering"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 11,
  fig.height = 6.5,
  dpi = 120
)

repo.dir <- "/Users/pgajer/current_projects/gflow"
report.dir <- file.path(repo.dir, "dev/graph-trend-filtering/reports")
diagnostics.file <- file.path(report.dir, "path_selection_diagnostics.rds")
if (!file.exists(diagnostics.file)) {
  source(file.path(repo.dir, "dev/graph-trend-filtering/path_selection_diagnostics.R"))
  dir.create(report.dir, recursive = TRUE, showWarnings = FALSE)
  diagnostics <- build_phase4_path_selection_diagnostics()
  saveRDS(diagnostics, diagnostics.file)
} else {
  diagnostics <- readRDS(diagnostics.file)
}

library(ggplot2)
```

## Purpose

This report is the Phase 4 diagnostic layer for the experimental
path-divided-difference graph trend-filtering operator.

The operator begins with canonical path rows

\[
\mathcal P_r(G)=\overrightarrow{\mathcal P}_r(G)/{\sim},
\]

where \(\sim\) identifies a path with its reverse.  The open modeling question
is whether every non-duplicate simple path should become a penalty row, or
whether a smaller row set should be selected to avoid over-constraining branch
points and dense data-derived graphs.

The report compares four row-selection rules for \(r=2\), corresponding to
order \(k=1\) trend filtering:

- `all`: all canonical simple length-2 paths;
- `branch.continuation`: paths whose internal vertices are not graph-theoretic branch points;
- `overlap.thinned`: a greedy low-redundancy subset based on the path-overlap graph;
- `coverage.greedy`: a greedy subset that prioritizes vertex and edge coverage.

This is deliberately a development report, not a public API proposal.

## Diagnostic Inputs

```{r input-summary}
summaries <- diagnostics$summaries
vertex.coverage <- diagnostics$vertex.coverage
path.tables <- diagnostics$path.tables

data.frame(
  generated_at = diagnostics$generated_at,
  n_graphs = length(unique(summaries$graph)),
  n_selection_rules = length(unique(summaries$selection)),
  n_diagnostic_rows = nrow(summaries),
  n_error_rows = sum(summaries$status != "ok")
)
```

The toy graphs are intentionally small:

- `path7`: ordinary seven-vertex path graph;
- `star3_arm2`: three-arm star with two vertices per arm;
- `star3_arm4`: longer three-arm star;
- `cycle8`: small cycle;
- `grid3x3`: small two-dimensional grid;
- `sknn1d12`: symmetric 3-nearest-neighbor graph from 12 one-dimensional samples.

## Summary Table

```{r summary-table}
summaries[order(summaries$graph, summaries$selection), ]
```

## Row Counts

```{r row-counts}
ggplot(summaries, aes(x = selection, y = selected.rows, fill = selection)) +
  geom_col(color = "gray35", linewidth = 0.2) +
  facet_wrap(~ graph, scales = "free_y") +
  labs(
    title = "Selected path rows by graph and selector",
    subtitle = "All rows are canonical modulo path reversal; r = 2",
    x = NULL,
    y = "Selected path rows"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

`branch.continuation` is intentionally restrictive.  On tree-like stars this
captures arm-continuation semantics.  On the kNN graph it can produce no rows,
because high graph degree is not the same as a geometric branch point.

## Nullity

```{r nullity}
ok.summary <- subset(summaries, status == "ok")

ggplot(ok.summary, aes(x = selection, y = nullity, fill = selection)) +
  geom_col(color = "gray35", linewidth = 0.2) +
  facet_wrap(~ graph, scales = "free_y") +
  labs(
    title = "Estimated nullity of the selected penalty matrix",
    subtitle = "For r = 2, path graphs should have nullity 2; the two-step three-arm star branch model has nullity 4",
    x = NULL,
    y = "Estimated nullity"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

The star examples are the key sanity check.  `all` includes cross-arm paths and
pushes the null space toward constants.  `branch.continuation` preserves
independent arm slopes.  The two experimental selectors recover similar
behavior on `star3_arm2` because their coverage objectives naturally choose the
three arm-continuation rows.

## Coverage

```{r vertex-coverage}
ggplot(vertex.coverage, aes(x = factor(vertex), y = coverage, fill = selection)) +
  geom_col(position = "dodge", color = "gray35", linewidth = 0.15) +
  facet_wrap(~ graph, scales = "free_x") +
  labs(
    title = "Vertex coverage by selected path rows",
    x = "Vertex",
    y = "Number of selected paths containing vertex",
    fill = "Selector"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(size = 7))
```

Coverage is not itself the goal, but it is a useful guardrail.  A selector that
gets the desired nullity by dropping entire regions of the graph is not a good
operator design.

## Path-Overlap Graph Diagnostics

```{r overlap-diagnostics}
ggplot(ok.summary,
       aes(x = selected.rows, y = mean.path.overlap.degree, color = selection)) +
  geom_point(size = 2.7) +
  geom_text(aes(label = graph), nudge_y = 0.25, size = 3, show.legend = FALSE) +
  labs(
    title = "Path-row count versus mean path-overlap degree",
    subtitle = "Lower overlap generally means less redundant local curvature constraints",
    x = "Selected path rows",
    y = "Mean overlap degree in selected path-overlap graph",
    color = "Selector"
  ) +
  theme_minimal(base_size = 11)
```

The overlap-thinning rule is a first attempt to select a sparse epsilon-net-like
subset of path rows.  The current implementation is intentionally simple: it
prefers low-overlap paths and then repairs uncovered vertices.  Its value here
is diagnostic, not final.

## Representative Path Rows

```{r sample-path-rows}
sample.paths <- do.call(rbind, lapply(split(path.tables, paste(path.tables$graph, path.tables$selection)), function(tab) {
  utils::head(tab[, c("graph", "selection", "row", "vertices", "metric.length", "row.weight")], 8)
}))
rownames(sample.paths) <- NULL
sample.paths
```

## Failed or Empty Selections

```{r errors}
errors <- subset(summaries, status != "ok")
if (nrow(errors) == 0L) {
  data.frame(message = "No failed selections")
} else {
  errors[, c("graph", "selection", "message")]
}
```

The `sknn1d12` failure for `branch.continuation`, if present, is expected and
important.  It demonstrates why graph degree is not a reliable branch detector
in data-derived kNN graphs.

## Preliminary Conclusions

1. Reversal canonicalization and row selection are separate operations.
   Reversal canonicalization removes duplicate absolute-value rows; row
   selection decides which non-duplicate paths define smoothness.

2. `branch.continuation` is a useful tree/star semantics test, but it is not
   suitable as a default for kNN graphs.

3. `all.simple` is the cleanest mathematical baseline, but on branched graphs it
   adds cross-arm paths and can shrink the null space aggressively.

4. `overlap.thinned` and `coverage.greedy` are promising as diagnostic
   selectors because they keep graph coverage while reducing redundant path
   constraints.

5. The next implementation decision should not be made from nullity alone.
   We should evaluate selected-row operators in regression benchmarks and
   inspect whether the chosen rows preserve known branchwise polynomial signals.
