---
title: "Transported Hessian Transport Rules On Quadric Graphs"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11.5,
  fig.height = 7.0,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "transported_hessian_quadric_multimethod_validation.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_QUADRIC_MULTIMETHOD_VALIDATION_MODE", unset = "smoke")
rds.path <- file.path(report.dir,
                      paste0("transported_hessian_quadric_multimethod_validation_",
                             mode, ".rds"))
diagnostics <- readRDS(rds.path)
metadata <- diagnostics$metadata
summary.table <- diagnostics$summary
probe.table <- diagnostics$probes
summary.ok <- subset(summary.table, status == "ok")
summary.ok$dim.label <- paste0("intrinsic dim ", summary.ok$intrinsic.dim)
summary.ok$surface.label <- paste(summary.ok$dim.label, summary.ok$surface, sep = " | ")
summary.ok$method.label <- factor(summary.ok$method.label,
                                  levels = unique(summary.ok$method.label))

nonzero.floor <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x)) min(x) / 2 else 1e-12
}
for (nm in c("residual.affine.mean", "residual.quadratic.mean",
             "residual.cubic.mean", "residual.fractional.min")) {
  summary.ok[[paste0("plot.", nm)]] <- pmax(summary.ok[[nm]],
                                            nonzero.floor(summary.ok[[nm]]))
}

base.theme <- function(base.size = 11) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 10)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}
```

# Purpose

This report compares the currently implemented transported-Hessian transport
families on data-derived graphs sampled from quadratic hypersurfaces.  A
sample starts from intrinsic coordinates

\[
u=(u_1,\ldots,u_m)
\]

and embeds them as a quadric graph

\[
\Phi(u)=\bigl(u,\ q(u)\bigr), \qquad
q(u)=\sum_{i\le k} c_i u_i^2-\sum_{i>k} c_i u_i^2.
\]

The graph is a connected symmetric kNN graph in the ambient data coordinates
\(\Phi(u)\).  Every method builds an order-2 transported-Hessian-like operator
\(A\).  The probes are intrinsic chart functions, so the test asks whether a
transport rule behaves like an intrinsic graph second-difference operator.

The method families are:

- regression-gradient transport, including intrinsic oracle, ambient data
  coordinates, and PCA;
- soft coordinate direction matching;
- the new edge-angle hard and soft rules, which compare directions by their
  angle relative to the oriented base dart.

The exact-coordinate rule is intentionally excluded: it is a useful grid/path
reference, but generic quadric kNN graphs do not have axis-aligned direction
labels.  The graph-derived local-chart backend for regression-gradient
transport is also excluded from this particular grid because it rebuilds a local
embedding per dart; that backend is tested in the separate regression-gradient
validation report.

```{r metadata, results='asis'}
cat(sprintf(
  "**Build:** %s  \n**Mode:** %s  \n**Runs:** %d  \n**Cases:** %d  \n**Graph weights:** %s\n",
  metadata$generated.at,
  metadata$mode,
  nrow(summary.table),
  metadata$n.cases,
  paste(metadata$graph.weights, collapse = ", ")
))
```

# Affine Probe Residuals

For intrinsic affine probes \(u_i\), the target diagnostic is

\[
R_{\mathrm{aff}}
=
\frac{1}{m}\sum_{i=1}^{m}
\frac{\|A u_i\|_2}{\|u_i\|_2}.
\]

A plausible second-difference operator should make this small.  This is the
main correctness screen for whether a transport rule respects the intrinsic
local linear structure of the sampled surface.

```{r affine-residuals, fig.height=8.2}
ggplot(summary.ok,
       aes(x = curvature,
           y = plot.residual.affine.mean,
           color = method.label,
           group = method.label)) +
  geom_line(linewidth = 0.75, alpha = 0.82) +
  geom_point(size = 2.0) +
  facet_grid(intrinsic.dim ~ surface, scales = "free_y") +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Intrinsic affine residuals by transport rule",
    subtitle = "Lower is better. These probes are intrinsic chart coordinates, not ambient coordinates.",
    x = "quadric curvature coefficient",
    y = expression(R[aff]),
    color = "method"
  ) +
  base.theme(9)
```

Interpretation: regression-gradient with intrinsic or ambient coordinates is
the affine benchmark.  Methods that stay close to it are successfully preserving
the null space expected of a second-difference-like construction.  Methods that
sit orders of magnitude higher are not reliable affine annihilators on these
data-derived graphs.

# Quadratic And Cubic Probes

The operator should not annihilate all higher-order structure.  For quadratic
and cubic intrinsic probes we report

\[
R_{\mathcal F}
=
\frac{1}{|\mathcal F|}
\sum_{p\in\mathcal F}
\frac{\|A p\|_2}{\|p\|_2},
\]

where \(\mathcal F\) is either the set of intrinsic monomials \(u_i u_j\) or
\(u_i u_j u_k\).  Cubic probes are diagnostics only; this report is still
testing order-2 operators.

```{r quadratic-cubic, fig.height=7.4}
qc <- rbind(
  data.frame(summary.ok,
             probe.family = "quadratic",
             plot.residual = summary.ok$plot.residual.quadratic.mean),
  data.frame(summary.ok,
             probe.family = "cubic",
             plot.residual = summary.ok$plot.residual.cubic.mean)
)
ggplot(qc,
       aes(x = method.label, y = plot.residual, fill = method.family)) +
  geom_boxplot(outlier.alpha = 0.45, width = 0.72) +
  facet_grid(probe.family ~ intrinsic.dim, scales = "free_y") +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "Higher-order intrinsic probe residuals",
    subtitle = "Quadratic and cubic probes should generally be detected by an order-2 transported Hessian.",
    x = NULL,
    y = expression(R[probe]),
    fill = "family"
  ) +
  base.theme(9)
```

Interpretation: very small affine residual together with nonzero quadratic and
cubic residual is the desired qualitative pattern.  If a method also drives
quadratic residuals to nearly zero, it is too weak for a Hessian-like role.

# Affine Versus Quadratic Separation

A useful operator should occupy the lower-right portion of this plot: low
affine residual, but appreciable quadratic residual.  This plot therefore
checks both null-space behavior and signal detection in one view.

```{r affine-quadratic-separation, fig.height=7.4}
ggplot(summary.ok,
       aes(x = plot.residual.affine.mean,
           y = plot.residual.quadratic.mean,
           color = method.label,
           shape = factor(intrinsic.dim))) +
  geom_point(alpha = 0.78, size = 2.4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Affine null-space behavior versus quadratic detection",
    subtitle = "The preferred pattern is low x-value and non-negligible y-value.",
    x = expression(R[aff]),
    y = expression(R[quad]),
    color = "method",
    shape = "intrinsic dim"
  ) +
  base.theme(9)
```

Interpretation: this plot makes clear when a method is failing because it does
not annihilate affine functions, versus when it is failing because it has become
too weak overall.

# Fractional-Power Probe Panel

Fractional probes use the intrinsic radius

\[
\rho(u)=
\frac{\|u-\bar u\|_2}{\max_w\|w-\bar u\|_2}
\]

and evaluate \(\rho^\alpha\) for non-integer \(\alpha\).  These are not meant to
define a polynomial null space.  They are a stress test for whether the operator
has a gradual, non-integer response profile rather than an all-or-nothing
integer-degree behavior.

```{r fractional-curves, fig.height=7.6}
fractional <- subset(probe.table, grepl("^fractional_radial_", probe))
fractional$alpha <- as.numeric(sub("^fractional_radial_", "", fractional$probe))
frac.summary <- aggregate(
  residual ~ method.label + method.family + intrinsic.dim + alpha,
  data = fractional,
  FUN = median,
  na.rm = TRUE
)
frac.summary$plot.residual <- pmax(frac.summary$residual,
                                   nonzero.floor(frac.summary$residual))
ggplot(frac.summary,
       aes(x = alpha,
           y = plot.residual,
           color = method.label,
           group = method.label)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.0) +
  facet_wrap(~ intrinsic.dim, labeller = label_both) +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Fractional radial probe response",
    subtitle = "Median residual over cases; this is a diagnostic response curve, not a null-space claim.",
    x = expression(alpha),
    y = expression(median~R[rho^alpha]),
    color = "method"
  ) +
  base.theme(9)
```

Interpretation: a smooth monotone response is reassuring.  A method whose
fractional response is erratic is likely being dominated by matching ambiguity
or row loss rather than by a clean differential analogy.

# Row Retention And Matching Diagnostics

For match-based methods, low residuals can be misleading if most candidate rows
were dropped.  We therefore inspect row retention,

\[
\mathrm{retention}
=
\frac{\#\{\hbox{accepted rows}\}}
       {\#\{\hbox{candidate rows}\}}.
\]

```{r row-retention, fig.height=7.0}
ggplot(summary.ok,
       aes(x = method.label,
           y = row.retention,
           fill = method.family)) +
  geom_boxplot(outlier.alpha = 0.45, width = 0.72) +
  coord_flip() +
  labs(
    title = "Accepted-row retention",
    subtitle = "Residuals are easier to interpret when the operator retains a substantial fraction of rows.",
    x = NULL,
    y = "row retention",
    fill = "family"
  ) +
  base.theme(9)
```

Interpretation: a method that looks good only after aggressive dropping should
be treated as a diagnostic filter, not as a stable operator.

# Results Summary

```{r results-summary, results='asis'}
ranked <- aggregate(
  cbind(residual.affine.mean, residual.quadratic.mean,
        residual.cubic.mean, row.retention) ~ method.label + method.family,
  data = summary.ok,
  FUN = median,
  na.rm = TRUE
)
ranked <- ranked[order(ranked$residual.affine.mean), ]
print(knitr::kable(
  ranked,
  digits = 4,
  caption = "Median diagnostics by method, ranked by affine residual."
))

best <- ranked[1, ]
cat(sprintf(
  "\nThe lowest median affine residual in this run is produced by **%s**. ",
  as.character(best$method.label)
))
cat("The preferred transported-Hessian behavior is not only low affine residual, ")
cat("but also nonzero quadratic/cubic residual and healthy row retention. ")
cat("The edge-angle methods are intentionally included here as a new baseline: ")
cat("they test how far one can get from local angle matching alone, before moving ")
cat("to richer direction-transport rules.\n")
```

# Appendix: Run Status

```{r run-status, results='asis'}
status.table <- as.data.frame.matrix(table(summary.table$method.label,
                                           summary.table$status))
status.table$method.label <- rownames(status.table)
status.table <- status.table[, c("method.label",
                                 setdiff(names(status.table), "method.label")),
                             drop = FALSE]
print(knitr::kable(status.table, caption = "Run status by method."))
errors <- subset(summary.table, status != "ok")
if (nrow(errors)) {
  print(knitr::kable(
    errors[, c("case.id", "method.label", "message")],
    caption = "Errors."
  ))
}
```

# Appendix: Reproduction

```{r reproduction, echo=TRUE, eval=FALSE}
# Smoke render
Rscript dev/graph-trend-filtering/render_transported_hessian_quadric_multimethod_validation.R

# Full render
GFLOW_QUADRIC_MULTIMETHOD_VALIDATION_MODE=full \
  Rscript dev/graph-trend-filtering/render_transported_hessian_quadric_multimethod_validation.R
```
