---
title: "Transported Hessian Transport-Confidence Filter Validation"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11.2,
  fig.height = 6.8,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "transported_hessian_transport_filter_validation.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_TRANSPORTED_HESSIAN_FILTER_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("transported_hessian_transport_filter_validation_", mode, ".rds"))
diagnostics <- readRDS(rds.path)
metadata <- diagnostics$metadata
policy.table <- diagnostics$policies
summary.table <- diagnostics$summary
embedding.table <- diagnostics$embedding %||% data.frame()
row.table <- diagnostics$rows %||% data.frame()
dropped.table <- diagnostics$dropped %||% data.frame()

method.levels <- c("cmdscale", "mds.edge.kk", "grip.edge.kk")
filter.levels <- policy.table$transport.filter
filter.labels <- setNames(policy.table$filter.label, policy.table$transport.filter)

summary.ok <- subset(summary.table, status == "ok")
summary.ok$local.embedding.method <- factor(summary.ok$local.embedding.method,
                                            levels = method.levels)
summary.ok$transport.filter <- factor(summary.ok$transport.filter,
                                      levels = filter.levels,
                                      labels = filter.labels[filter.levels])
if (nrow(row.table)) {
  row.table$local.embedding.method <- factor(row.table$local.embedding.method,
                                             levels = method.levels)
  row.table$transport.filter <- factor(row.table$transport.filter,
                                       levels = filter.levels,
                                       labels = filter.labels[filter.levels])
}
if (nrow(dropped.table)) {
  dropped.table$local.embedding.method <- factor(dropped.table$local.embedding.method,
                                                 levels = method.levels)
  dropped.table$transport.filter <- factor(dropped.table$transport.filter,
                                           levels = filter.levels,
                                           labels = filter.labels[filter.levels])
}

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}

nonzero.floor <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x)) min(x) / 2 else 1e-12
}

residual.long <- rbind(
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method",
                            "transport.filter", "row.retention")],
             probe = "constant",
             residual = summary.ok$residual.constant),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method",
                            "transport.filter", "row.retention")],
             probe = "linear mean",
             residual = summary.ok$residual.linear.mean),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method",
                            "transport.filter", "row.retention")],
             probe = "overall",
             residual = summary.ok$residual.overall)
)
residual.long$plot.residual <- pmax(residual.long$residual,
                                    nonzero.floor(residual.long$residual))

confidence.long <- rbind(
  data.frame(summary.ok[, c("case.id", "local.embedding.method", "transport.filter")],
             diagnostic = "best-score robust z",
             value = summary.ok$mean.best.score.robust.z),
  data.frame(summary.ok[, c("case.id", "local.embedding.method", "transport.filter")],
             diagnostic = "margin robust z",
             value = summary.ok$mean.margin.robust.z),
  data.frame(summary.ok[, c("case.id", "local.embedding.method", "transport.filter")],
             diagnostic = "effective-match fraction",
             value = summary.ok$mean.effective.match.fraction),
  data.frame(summary.ok[, c("case.id", "local.embedding.method", "transport.filter")],
             diagnostic = "margin quantile",
             value = summary.ok$mean.margin.local.quantile)
)
```

# Purpose

The transported graph Hessian compares directional derivatives across a graph
edge.  For a base dart \(u\to u'\), a direction \(u\to v\), and a matched
direction \(u'\to v'\), the row is

\[
  \delta_{u'\to v'} f - \delta_{u\to v} f,
  \qquad
  \delta_{a\to b} f = \frac{f(b)-f(a)}{\ell_{ab}}.
\]

The previous backend validation showed that true edge-KK local coordinate
repair can be used and audited.  It also showed that better edge stress alone
does not guarantee a better transported Hessian.  The next question is whether
we should reject low-confidence direction transports before they become rows of
the operator.

This report evaluates five transport policies:

\[
\text{no filter},\quad
\theta_1\le \pi/3\ \text{and}\ r_\ell\le 0.50,\quad
\text{local quantile},\quad
\text{local robust-}z,\quad
\text{local robust-}z+\text{entropy fraction},
\]

where \(s_1\) and \(s_2\) are the best and second-best match scores, \(H\) is
softmax entropy, \(\theta_1\) is the best-match angle, and

\[
r_\ell =
\frac{|\ell_{\mathrm{matched}}-\ell_{\mathrm{direction}}|}
     {\ell_{\mathrm{direction}}}.
\]

The local rules compare each candidate direction match against the alternatives
available at the same transported vertex.  For sorted candidate scores \(s\),
score gaps \(\Delta s\), and robust scale
\(\operatorname{scale}(x)=1.4826\,\operatorname{MAD}(x)\) with fallbacks, the
adaptive diagnostics are

\[
  q_s = \widehat F_s(s_1), \qquad
  q_m = \widehat F_{\Delta s}(s_2-s_1),
\]

and

\[
  z_s = \frac{\operatorname{median}(s)-s_1}{\operatorname{scale}(s)}, \qquad
  z_m = \frac{(s_2-s_1)-\operatorname{median}(\Delta s)}
              {\operatorname{scale}(\Delta s)}.
\]

The entropy-fraction gate uses

\[
  \rho_H = \frac{\exp(H)}{n_{\mathrm{candidate}}},
\]

so that the same effective-match count is judged relative to local degree.

```{r metadata-summary, results='asis'}
cat(sprintf(
  "**Build:** %s  \n**Mode:** %s  \n**Local grip requested:** %s  \n**True edge-KK available:** %s  \n**Edge-KK function:** `%s`\n",
  metadata$generated.at,
  metadata$mode,
  metadata$use.local.grip,
  metadata$edge.kk.available,
  metadata$edge.kk.function %||% NA_character_
))
```

# Row Retention

Filtering can only help if it removes questionable rows without destroying the
operator.  The retention fraction is

\[
\rho =
\frac{\#\{\text{accepted Hessian rows}\}}
     {\#\{\text{candidate direction transports}\}}.
\]

High retention means the filter is conservative.  Very low retention is a
warning: polynomial residuals may look artificially good if the operator has
become too small to test anything.

```{r row-retention, fig.height=7.2}
ggplot(summary.ok,
       aes(x = transport.filter, y = row.retention,
           fill = local.embedding.method)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72,
           color = "grey25") +
  facet_wrap(~ case.id, nrow = 2) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Accepted Hessian-row fraction under each transport filter",
    subtitle = "A useful filter removes ambiguous transports without collapsing the operator.",
    x = "transport filter",
    y = "row retention fraction",
    fill = "embedding backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The retention plot is the first safety check.  A filter that drops nearly all
rows may reduce residuals for the wrong reason.  A filter that drops nothing
cannot teach us much about transport confidence.

# Why Rows Were Dropped

The confidence gates are intentionally interpretable.  A candidate can be
dropped because the best angle is too large, the best length mismatch is too
large, the local score quantile is not selective enough, the local robust
z-score is too weak, or the softmax distribution is too diffuse relative to the
local candidate count.

```{r drop-reasons, fig.height=7.0}
if (nrow(dropped.table)) {
  drop.count <- as.data.frame(
    xtabs(~ case.id + local.embedding.method + transport.filter + drop.reason,
          data = dropped.table)
  )
  names(drop.count)[names(drop.count) == "Freq"] <- "n"
  drop.count <- subset(drop.count, n > 0)
  ggplot(drop.count,
         aes(x = transport.filter, y = n, fill = drop.reason)) +
    geom_col(width = 0.72, color = "grey25") +
    facet_grid(local.embedding.method ~ case.id, scales = "free_y") +
    labs(
      title = "Drop reasons by case, backend, and filter",
      subtitle = "The dominant reason identifies which part of the transport confidence score is binding.",
      x = "transport filter",
      y = "dropped candidate rows",
      fill = "drop reason"
    ) +
    base.theme(10) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No dropped rows were observed.") +
    theme_void()
}
```

This figure explains the mechanism behind row loss.  If all drops come from
local robust-z or quantile thresholds, the local geometry produces tied or
nearly tied directions.  If drops come from angle or length, the candidate has
a clear best match but the match is geometrically implausible.  If drops come
from effective-match fraction, the soft transport remains too spread out after
local calibration.

# Confidence Diagnostics

The row filters should improve the confidence diagnostics of the rows that
remain.  We summarize the accepted rows by the adaptive quantities used by the
new gates: best-score robust \(z\), margin robust \(z\), effective-match
fraction, and margin quantile.

```{r confidence-diagnostics, fig.height=8.0}
ggplot(confidence.long,
       aes(x = transport.filter, y = value,
           fill = local.embedding.method)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72,
           color = "grey25") +
  facet_grid(diagnostic ~ case.id, scales = "free_y") +
  labs(
    title = "Confidence diagnostics among accepted transport rows",
    subtitle = "Filters should increase margin or sharpness without forcing implausible angle or length matches.",
    x = "transport filter",
    y = "mean accepted-row diagnostic",
    fill = "embedding backend"
  ) +
  base.theme(10) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

These diagnostics are descriptive rather than final objectives.  For example,
a low effective-match fraction is good only if the selected direction is a
geometrically meaningful match and the retained rows still test the polynomial
probes.

# Polynomial Residuals

The core mathematical check is how the filtered operator acts on simple
polynomial probes.  For a probe matrix \(P\), we report

\[
  R(P) = \frac{\|AP\|_F}{\|P\|_F}.
\]

For a transported second-difference operator, constants and linear coordinate
probes are the most important sanity checks.  Lower residuals are better only
when row retention remains high enough for the operator to be meaningful.

```{r polynomial-residuals, fig.height=8.0}
ggplot(residual.long,
       aes(x = transport.filter, y = plot.residual,
           fill = local.embedding.method)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72,
           color = "grey25") +
  facet_grid(probe ~ case.id, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Polynomial-probe residuals after transport filtering",
    subtitle = "Residuals are shown on a log scale; exact zeros are floored only for display.",
    x = "transport filter",
    y = "residual ||AP||F / ||P||F",
    fill = "embedding backend"
  ) +
  base.theme(10) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

This plot is the main evidence about whether confidence filtering helps the
operator mathematically.  The desired pattern is lower constant/linear
residuals with only moderate row loss.

# Nullity And Operator Size

Nullity is a coarse global check.  If a filter drops many rows, the nullity can
increase simply because the operator has fewer constraints.  We therefore read
nullity together with row retention and residuals.

```{r nullity, fig.height=7.2}
ggplot(summary.ok,
       aes(x = transport.filter, y = nullity.estimate,
           fill = local.embedding.method)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72,
           color = "grey25") +
  facet_wrap(~ case.id, scales = "free_y", nrow = 2) +
  labs(
    title = "Estimated nullity after transport filtering",
    subtitle = "Nullity increases are useful only if the remaining operator still has enough retained rows.",
    x = "transport filter",
    y = "estimated nullity",
    fill = "embedding backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The nullity plot distinguishes true improvement from degeneracy.  A filter that
raises nullity while retaining few rows is not a solved transport rule; it is a
diagnostic that we are rejecting too many constraints.

# Results Summary And Discussion

```{r compact-best-table, results='asis'}
score.table <- summary.ok
score.table$linear.with.retention <- score.table$residual.linear.mean /
  pmax(score.table$row.retention, 1e-6)
score.table <- subset(score.table, row.retention >= 0.25)
best <- score.table[order(score.table$case.id,
                          score.table$linear.with.retention,
                          score.table$residual.linear.mean), ]
best <- best[!duplicated(best$case.id),
             c("case.id", "local.embedding.method", "transport.filter",
               "row.retention", "residual.linear.mean", "nullity.estimate",
               "mean.best.score.robust.z", "mean.margin.robust.z",
               "mean.effective.match.fraction")]
knitr::kable(best, digits = 4,
             caption = "Best linear-residual-with-retention compromise by case.")
```

The experiment asks whether local confidence filtering improves the
transported Hessian in a way that survives basic mathematical checks.  The
adaptive policies are designed to avoid one global threshold pretending that a
path endpoint, a grid interior vertex, and a sampled kNN neighborhood have the
same candidate geometry.  Confidence filtering now removes low-quality rows for
auditable reasons: weak local score separation, weak local margin separation,
diffuse softmax mass, and geometrically implausible angle/length matches.

The residual plots show that filtering can change the polynomial behavior, but
row retention is the controlling caveat.  A filter that makes residuals smaller
by dropping too many rows is not a valid default.  The most promising filters
are therefore the ones that reduce linear residuals while keeping row retention
close to the unfiltered operator.

The backend interaction also matters.  Edge-KK improves local edge geometry,
but confidence filtering is still needed because edge stress and transport
confidence are different quantities.  A local embedding can preserve edge
lengths and still produce ambiguous or geometrically mismatched direction
transport.  The entropy-fraction policy is deliberately stricter on high-degree
neighborhoods, where a large absolute effective-match count can be acceptable
or unacceptable depending on how many candidate directions were available.

# What We Learned

1. Transport confidence must be treated as a first-class diagnostic.  Backend
quality alone is not enough.
2. The most interpretable quantities are best-match margin, effective match
count, best-match angle, and relative length error.
3. Filtering should be judged jointly by row retention and polynomial
residuals.  Residuals alone can be misleading when the operator is heavily
thinned.
4. Adaptive thresholds are now implemented as first-class policy choices.  The
next refinement should calibrate their defaults against separate graph families
rather than tuning them on a single report.

# Appendix

## Run Metadata

```{r appendix-metadata}
str(metadata)
```

## Filter Policies

```{r appendix-policies}
knitr::kable(policy.table, digits = 4)
```

## Summary Table

```{r appendix-summary}
knitr::kable(summary.table, digits = 5)
```

## Reproducibility

```{r appendix-command, results='asis'}
cat("Smoke render:\n\n")
cat("```bash\n")
cat("Rscript dev/graph-trend-filtering/render_transported_hessian_transport_filter_validation.R\n")
cat("```\n\n")
cat("Full render:\n\n")
cat("```bash\n")
cat("GFLOW_TRANSPORTED_HESSIAN_FILTER_MODE=full Rscript dev/graph-trend-filtering/render_transported_hessian_transport_filter_validation.R\n")
cat("```\n")
```
