---
title: "Transported Hessian Adaptive-Threshold Calibration"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11.4,
  fig.height = 6.8,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "transported_hessian_adaptive_threshold_sweep.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_ADAPTIVE_THRESHOLD_SWEEP_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("transported_hessian_adaptive_threshold_sweep_", mode, ".rds"))
diagnostics <- readRDS(rds.path)
metadata <- diagnostics$metadata
policy.table <- diagnostics$policies
summary.table <- diagnostics$summary
drop.count <- diagnostics$drop.count %||% data.frame()

summary.ok <- subset(summary.table, status == "ok")
summary.ok$rule.family <- factor(summary.ok$rule.family,
                                 levels = c("baseline", "local_quantile",
                                            "local_robust_z"))
summary.ok$plot.residual.linear <- pmax(summary.ok$residual.linear.mean,
                                         min(summary.ok$residual.linear.mean[
                                           is.finite(summary.ok$residual.linear.mean) &
                                             summary.ok$residual.linear.mean > 0
                                         ], na.rm = TRUE) / 2)
summary.ok$graph.label <- paste(summary.ok$case.id, summary.ok$local.embedding.method,
                                sep = " | ")

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}

feasible.cutoff <- 0.50
summary.ok$feasible <- summary.ok$row.retention >= feasible.cutoff &
  !summary.ok$degenerate
score.table <- subset(summary.ok, feasible)
score.table$score <- score.table$residual.linear.mean
best.feasible <- score.table[order(score.table$case.id,
                                   score.table$local.embedding.method,
                                   score.table$score,
                                   -score.table$row.retention), ]
best.feasible <- best.feasible[!duplicated(paste(best.feasible$case.id,
                                                 best.feasible$local.embedding.method)), ]

quantile.ok <- subset(summary.ok, rule.family == "local_quantile")
robust.ok <- subset(summary.ok, rule.family == "local_robust_z")

best.by.family <- do.call(rbind, lapply(split(summary.ok, summary.ok$rule.family), function(tab) {
  tab <- subset(tab, feasible)
  if (!nrow(tab)) return(data.frame())
  tab <- tab[order(tab$case.id, tab$local.embedding.method,
                   tab$residual.linear.mean, -tab$row.retention), ]
  tab[!duplicated(paste(tab$case.id, tab$local.embedding.method)), ]
}))
```

# Purpose

The transported graph Hessian creates a row by comparing a source directional
difference at \(u\) with a transported directional difference at \(u'\).  For a
candidate set at the target fiber, let

\[
  s_{(1)} \le s_{(2)} \le \cdots \le s_{(J)}
\]

be the ordered match scores.  Adaptive thresholding asks whether the best match
is locally good and locally separated enough to trust:

\[
  m=s_{(2)}-s_{(1)}.
\]

The previous report showed the main danger: adaptive rules can reduce
polynomial residuals by dropping too many rows.  This calibration report
therefore treats row retention as a constraint, not as a secondary detail.
The main question is:

\[
  \text{minimize } R_{\mathrm{linear}}
  =
  \frac{\|AP_{\mathrm{linear}}\|_F}{\|P_{\mathrm{linear}}\|_F}
  \quad
  \text{subject to}
  \quad
  \rho_{\mathrm{row}}\ge 0.50
  \text{ and nondegeneracy}.
\]

```{r metadata-summary, results='asis'}
cat(sprintf(
  "**Build:** %s  \n**Mode:** %s  \n**Backends:** `%s`  \n**Local grip requested:** %s  \n**True edge-KK available:** %s  \n**Edge-KK function:** `%s`  \n**Policies tested:** %d\n",
  metadata$generated.at,
  metadata$mode,
  paste(metadata$methods, collapse = "`, `"),
  metadata$use.local.grip,
  metadata$edge.kk.available,
  metadata$edge.kk.function %||% NA_character_,
  nrow(policy.table)
))
```

# Retention-Residual Frontier

Each point is one threshold setting on one graph.  The vertical dashed line is
the minimum retention constraint \(\rho_{\mathrm{row}}\ge 0.50\).  Points to the
left of the line are informative as diagnostics, but they are not acceptable
default candidates because the operator may have been hollowed out.

```{r frontier, fig.height=7.8}
ggplot(summary.ok,
       aes(x = row.retention, y = plot.residual.linear,
           color = rule.family, shape = degenerate)) +
  geom_vline(xintercept = feasible.cutoff, linetype = 2, color = "grey35") +
  geom_point(alpha = 0.78, size = 2.2) +
  facet_wrap(~ graph.label, scales = "free_y", ncol = 2) +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Adaptive-threshold residual versus row-retention frontier",
    subtitle = "Useful thresholds live low on the y-axis and to the right of the retention constraint.",
    x = "accepted-row fraction",
    y = expression(R[linear] == " " * norm(A * P[linear], "F") / norm(P[linear], "F")),
    color = "rule family",
    shape = "degenerate"
  ) +
  base.theme(10)
```

The frontier is the main calibration view.  A threshold family is promising only
when it improves linear residuals while staying to the right of the retention
line.  Points with low residual and very low retention are not wins; they are
operator-degeneracy warnings.

# Best Feasible Settings

The next figure shows the best feasible setting for each graph/backend pair
under the \(0.50\) row-retention constraint.  The baseline fixed angle/length
rule is included, so an adaptive rule must beat a conservative fixed gate while
retaining enough rows.

```{r best-feasible, fig.height=7.2}
if (nrow(best.feasible)) {
  ggplot(best.feasible,
         aes(x = reorder(graph.label, residual.linear.mean),
             y = residual.linear.mean,
             fill = rule.family)) +
    geom_col(color = "grey25", width = 0.74) +
    coord_flip() +
    scale_y_log10() +
    scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
    labs(
      title = "Best feasible threshold setting by graph",
      subtitle = "Feasible means row retention at least 0.50 and no degenerate operator flag.",
      x = "graph and backend",
      y = expression(R[linear]),
      fill = "winning family"
    ) +
    base.theme(11)
} else {
  ggplot() +
    annotate("text", x = 0, y = 0,
             label = "No feasible threshold setting met the retention constraint.") +
    theme_void()
}
```

This view answers the practical default question.  If the same adaptive family
wins repeatedly across graph types, it is a plausible default.  If the fixed
baseline keeps winning, adaptive gates should remain diagnostic or
case-specific until better calibration is available.

# Quantile-Gate Heatmap

The local quantile rule accepts rows when

\[
  q_s=\widehat F_s(s_{(1)})\le q_{\max},
  \qquad
  q_m=\widehat F_{\Delta s}(m)\ge q_{\min}.
\]

The heatmap reports the best feasible linear residual at each
\((q_{\max},q_{\min})\), minimizing over the optional entropy-fraction gate.
Blank cells did not meet the row-retention constraint.

```{r quantile-heatmap, fig.height=7.4}
if (nrow(quantile.ok)) {
  qtab <- subset(quantile.ok, feasible)
  qbest <- qtab[order(qtab$case.id, qtab$local.embedding.method,
                      qtab$match.score.quantile, qtab$match.margin.quantile,
                      qtab$residual.linear.mean), ]
  qbest <- qbest[!duplicated(paste(qbest$case.id,
                                   qbest$local.embedding.method,
                                   qbest$match.score.quantile,
                                   qbest$match.margin.quantile)), ]
  ggplot(qbest,
         aes(x = factor(match.score.quantile),
             y = factor(match.margin.quantile),
             fill = residual.linear.mean)) +
    geom_tile(color = "white", linewidth = 0.6) +
    facet_wrap(~ graph.label, ncol = 2) +
    scale_fill_viridis_c(option = "magma", trans = "log10", na.value = "grey90") +
    labs(
      title = "Local quantile gate: best feasible residual by threshold pair",
      subtitle = "Each cell minimizes over the optional effective-match-fraction gate.",
      x = expression(q[max]~"(best score quantile upper bound)"),
      y = expression(q[min]~"(margin quantile lower bound)"),
      fill = expression(R[linear])
    ) +
    base.theme(10)
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No local quantile runs were available.") +
    theme_void()
}
```

The quantile rule is intentionally local-degree aware, but the heatmap makes
its danger visible: as \(q_{\min}\) becomes too strict, many graphs lose too
many rows and disappear from the feasible set.

# Robust-\(z\) Gate Heatmap

The robust-\(z\) rule accepts rows when

\[
  z_s =
  \frac{\operatorname{median}(s)-s_{(1)}}{\operatorname{scale}(s)}
  \ge z_{s,\min},
  \qquad
  z_m =
  \frac{m-\operatorname{median}(\Delta s)}{\operatorname{scale}(\Delta s)}
  \ge z_{m,\min}.
\]

The heatmap reports the best feasible residual at each
\((z_{s,\min},z_{m,\min})\), again minimizing over the optional entropy-fraction
gate.

```{r robust-heatmap, fig.height=7.4}
if (nrow(robust.ok)) {
  rtab <- subset(robust.ok, feasible)
  rbest <- rtab[order(rtab$case.id, rtab$local.embedding.method,
                      rtab$min.best.score.z, rtab$min.margin.z,
                      rtab$residual.linear.mean), ]
  rbest <- rbest[!duplicated(paste(rbest$case.id,
                                   rbest$local.embedding.method,
                                   rbest$min.best.score.z,
                                   rbest$min.margin.z)), ]
  ggplot(rbest,
         aes(x = factor(min.best.score.z),
             y = factor(min.margin.z),
             fill = residual.linear.mean)) +
    geom_tile(color = "white", linewidth = 0.6) +
    facet_wrap(~ graph.label, ncol = 2) +
    scale_fill_viridis_c(option = "mako", trans = "log10", na.value = "grey90") +
    labs(
      title = "Local robust-z gate: best feasible residual by threshold pair",
      subtitle = "Each cell minimizes over the optional effective-match-fraction gate.",
      x = expression(z[s,min]~"(best-score z lower bound)"),
      y = expression(z[m,min]~"(margin z lower bound)"),
      fill = expression(R[linear])
    ) +
    base.theme(10)
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No local robust-z runs were available.") +
    theme_void()
}
```

The robust-\(z\) rule is less tied to the discreteness of empirical quantiles,
but it still needs calibration: large \(z\)-thresholds can reject exactly the
path and boundary comparisons needed to keep the operator meaningful.

# Degeneracy And Drop Mechanisms

Degeneracy is not a secondary warning; it is the failure mode that this report
is designed to expose.  A very low residual with very low row retention means
the operator stopped testing the polynomial probe.

```{r degeneracy, fig.height=7.2}
deg <- aggregate(degenerate ~ rule.family + case.id + local.embedding.method,
                 data = summary.ok, FUN = mean)
deg$graph.label <- paste(deg$case.id, deg$local.embedding.method, sep = " | ")
ggplot(deg,
       aes(x = rule.family, y = degenerate, fill = rule.family)) +
  geom_col(width = 0.74, color = "grey25") +
  facet_wrap(~ graph.label, ncol = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  scale_fill_brewer(palette = "Dark2", guide = "none", na.translate = FALSE) +
  labs(
    title = "Degenerate-operator rate by threshold family",
    subtitle = "A setting is degenerate when it leaves zero or one Hessian row.",
    x = "threshold family",
    y = "degenerate fraction"
  ) +
  base.theme(10) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r drop-mechanisms, fig.height=7.4}
if (nrow(drop.count)) {
  dtab <- subset(drop.count, n > 0 & drop.reason != "none")
  dsum <- aggregate(n ~ rule.family + drop.reason, data = dtab, FUN = sum)
  ggplot(dsum,
         aes(x = rule.family, y = n, fill = drop.reason)) +
    geom_col(width = 0.74, color = "grey25") +
    scale_y_log10() +
    labs(
      title = "Dominant row-drop mechanisms across the sweep",
      subtitle = "Counts are log-scaled because some strict rules dominate the raw totals.",
      x = "threshold family",
      y = "dropped candidate rows",
      fill = "drop reason"
    ) +
    base.theme(10) +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
} else {
  ggplot() +
    annotate("text", x = 0, y = 0, label = "No dropped rows were recorded.") +
    theme_void()
}
```

These figures distinguish a useful confidence rule from a destructive thinning
rule.  A good rule should mostly remove ambiguous local matches while keeping
enough rows to maintain a low-dimensional null space.

# Results Summary

```{r results-summary, results='asis'}
family.summary <- aggregate(cbind(row.retention, residual.linear.mean,
                                  residual.overall, degenerate) ~ rule.family,
                            data = summary.ok,
                            FUN = function(x) median(x, na.rm = TRUE))
knitr::kable(family.summary, digits = 4,
             caption = "Median behavior by threshold family.")

if (nrow(best.feasible)) {
  show <- best.feasible[, c("case.id", "local.embedding.method", "rule.family",
                            "transport.filter", "row.retention",
                            "residual.linear.mean", "nullity.estimate",
                            "mean.effective.match.fraction")]
  knitr::kable(show, digits = 4,
               caption = "Best feasible policy by graph/backend pair.")
}
```

The calibration experiment should be read conservatively.  The best threshold
is not the one with the smallest residual in isolation; it is the one with the
smallest residual among policies that retain enough rows and avoid degeneracy.
This report therefore turns adaptive thresholding from an attractive idea into
a measurable tradeoff.

```{r automatic-interpretation, results='asis'}
best.universal <- aggregate(cbind(residual.linear.mean, row.retention) ~
                              transport.filter + rule.family,
                            data = subset(summary.ok, feasible),
                            FUN = function(x) median(x, na.rm = TRUE))
feasible.rate <- aggregate(feasible ~ transport.filter + rule.family,
                           data = summary.ok, FUN = mean)
best.universal <- merge(best.universal, feasible.rate,
                        by = c("transport.filter", "rule.family"))
best.universal <- best.universal[order(-best.universal$feasible,
                                       best.universal$residual.linear.mean), ]
winner <- best.universal[1, ]

best.family.count <- if (nrow(best.feasible)) {
  as.data.frame(table(best.feasible$rule.family), stringsAsFactors = FALSE)
} else {
  data.frame(Var1 = character(), Freq = integer())
}
names(best.family.count) <- c("rule.family", "n.graphs")

cat("## What We Learned\n\n")
cat(sprintf(
  "Under the %.0f%% row-retention constraint, the best broadly feasible policy is `%s` from the `%s` family. Its median feasible linear residual is %.4g and its median row retention is %.3f.\n\n",
  feasible.cutoff * 100,
  winner$transport.filter,
  winner$rule.family,
  winner$residual.linear.mean,
  winner$row.retention
))
if (nrow(best.family.count)) {
  pieces <- sprintf("`%s`: %d", best.family.count$rule.family,
                    best.family.count$n.graphs)
  cat("Best feasible policy counts by graph/backend pair: ",
      paste(pieces, collapse = "; "), ".\n\n", sep = "")
}
cat("The practical conclusion is that adaptive gates are useful calibration tools, ",
    "but they are not yet a safe universal default.  The fixed angle/length gate ",
    "remains the conservative baseline whenever an adaptive rule improves residuals ",
    "mainly by moving leftward on the retention frontier.\n", sep = "")
```

The expected outcome is not necessarily a single universal default.  It may be
that fixed angle/length gates are safer on irregular or sampled graphs, while
adaptive gates are useful on clean path/grid references.  If the frontier shows
that pattern, the next implementation step should expose adaptive gates as
diagnostics or case-specific policies rather than making them the default.

# Appendix

## Run Metadata

```{r appendix-metadata}
str(metadata)
```

## Policy Grid

```{r appendix-policies}
knitr::kable(policy.table, digits = 4)
```

## Summary Table

```{r appendix-summary}
knitr::kable(summary.table, digits = 5)
```

## Reproducibility

```{r appendix-command, results='asis'}
cat("Smoke render:\n\n")
cat("```bash\n")
cat("Rscript dev/graph-trend-filtering/render_transported_hessian_adaptive_threshold_sweep.R\n")
cat("```\n\n")
cat("Full render:\n\n")
cat("```bash\n")
cat("GFLOW_ADAPTIVE_THRESHOLD_SWEEP_MODE=full Rscript dev/graph-trend-filtering/render_transported_hessian_adaptive_threshold_sweep.R\n")
cat("```\n")
```
