---
title: "Transported Hessian Backend Validation"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.8,
  fig.height = 6.4,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "transported_hessian_backend_validation.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_TRANSPORTED_HESSIAN_BACKEND_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("transported_hessian_backend_validation_", mode, ".rds"))
diagnostics <- readRDS(rds.path)
summary.table <- diagnostics$summary
embedding.table <- diagnostics$embedding %||% data.frame()
row.table <- diagnostics$rows %||% data.frame()
metadata <- diagnostics$metadata

summary.ok <- subset(summary.table, status == "ok")
method.levels <- c("cmdscale", "mds.edge.kk", "grip.edge.kk")
summary.ok$local.embedding.method <- factor(summary.ok$local.embedding.method,
                                            levels = method.levels)
if (nrow(embedding.table)) {
  embedding.table$local.embedding.method <- factor(embedding.table$local.embedding.method,
                                                   levels = method.levels)
}
if (nrow(row.table)) {
  row.table$local.embedding.method <- factor(row.table$local.embedding.method,
                                             levels = method.levels)
}

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}

nonzero.floor <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x)) min(x) / 2 else 1e-12
}

metric.long <- rbind(
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             metric = "edge stress",
             value = summary.ok$median.edge.stress),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             metric = "graph-distance stress",
             value = summary.ok$median.graph.distance.stress)
)
transport.long <- rbind(
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             metric = "transport entropy",
             value = summary.ok$mean.transport.entropy),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             metric = "effective matches",
             value = summary.ok$mean.effective.matches),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             metric = "match score",
             value = summary.ok$mean.match.score)
)
residual.long <- rbind(
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             probe = "constant",
             residual = summary.ok$residual.constant),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             probe = "linear mean",
             residual = summary.ok$residual.linear.mean),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             probe = "quadratic mean",
             residual = summary.ok$residual.quadratic.mean),
  data.frame(summary.ok[, c("case.id", "graph.type", "local.embedding.method")],
             probe = "overall",
             residual = summary.ok$residual.overall)
)
residual.long$plot.residual <- pmax(residual.long$residual,
                                    nonzero.floor(residual.long$residual))

backend.status <- unique(summary.table[, c("local.embedding.method", "backend.used",
                                           "fallback.reason")])
backend.status$local.embedding.method <- factor(backend.status$local.embedding.method,
                                                levels = method.levels)
backend.status$uses.true.edge.kk <- grepl("edge\\.kk|edge\\.isometric",
                                          backend.status$backend.used %||% "")
```

# Purpose

The transported graph Hessian operator needs local directions before it can
compare directional derivatives across an edge.  In this Phase 3 validation we
ask whether the local embedding backend used to create those directions is
doing the intended job.  The three backends are:

\[
\texttt{cmdscale},\qquad
\texttt{mds.edge.kk},\qquad
\texttt{grip.edge.kk}.
\]

The crucial implementation question is whether the edge-KK variants use the
true edge-restricted optimizer,

\[
\frac{1}{2}\sum_{(i,j)\in E}
k_{ij}\left(\|z_i-z_j\|_2-s\ell_{ij}\right)^2,
\]

rather than full all-pairs geodesic KK.  Full geodesic KK is a different
objective.  This report therefore records the backend actually used for each
local disk and compares edge stress, graph-distance stress, transport entropy,
and polynomial-probe residuals.

```{r metadata-summary, results='asis'}
cat(sprintf(
  "**Build:** %s  \n**Mode:** %s  \n**Local grip requested:** %s  \n**True edge-KK available:** %s  \n**Edge-KK function:** `%s`\n",
  metadata$generated.at,
  metadata$mode,
  metadata$use.local.grip,
  metadata$edge.kk.available,
  metadata$edge.kk.function %||% NA_character_
))
```

# Backend Resolution

This diagnostic checks the first necessary condition: the requested backend
name must resolve to the intended implementation.  In particular,
`grip.edge.kk` and `mds.edge.kk` should report a backend containing
`grip.optimize.edge.kk.layout` or `grip.optimize.edge.isometric.layout` when
true edge-KK is available.  If the installed `grip` is older, the report should
show a transparent fallback to `cmdscale`.

```{r backend-resolution, fig.height=4.8}
ggplot(backend.status,
       aes(x = local.embedding.method, y = backend.used,
           color = uses.true.edge.kk)) +
  geom_point(size = 3.2) +
  facet_wrap(~ local.embedding.method, scales = "free_y", nrow = 1) +
  labs(
    title = "Resolved backend for each requested local embedding method",
    subtitle = "The edge-KK methods should resolve to true edge-KK when the local grip checkout is loaded.",
    x = "requested local embedding method",
    y = "backend used",
    color = "true edge-KK"
  ) +
  base.theme() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

The status plot is the guardrail for this report.  If the edge-KK methods fall
back to `cmdscale`, then the stress and transport comparisons below are still
valid software diagnostics, but they are not evidence about the edge-KK
optimizer itself.

# Local Embedding Stress

The primary geometric diagnostic is edge stress:

\[
E_{\mathrm{edge}}
=
\frac{
\sqrt{|E_{\Gamma}|^{-1}\sum_{(i,j)\in E_{\Gamma}}
(\|z_i-z_j\|_2-\ell_{ij})^2}
}{
|E_{\Gamma}|^{-1}\sum_{(i,j)\in E_{\Gamma}}\ell_{ij}
}.
\]

We also record graph-distance stress over local disk pairs.  Edge-KK is
expected to improve edge stress most directly; it need not minimize all-pairs
graph-distance stress.

```{r stress-comparison, fig.height=7.0}
metric.long$plot.value <- pmax(metric.long$value, nonzero.floor(metric.long$value))
ggplot(metric.long,
       aes(x = local.embedding.method, y = plot.value,
           fill = local.embedding.method)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(metric ~ case.id, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Local embedding stress by backend",
    subtitle = "Bars are median local-disk stress over transported-Hessian base darts; lower is better.",
    x = "local embedding backend",
    y = "median stress (log scale)",
    fill = "backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The intended positive pattern is a reduction in edge stress for
`mds.edge.kk` or `grip.edge.kk` relative to `cmdscale`.  If graph-distance
stress increases while edge stress falls, that is not necessarily a failure:
it means the local embedding is prioritizing the edge-length constraints used
by the transported directions.

# Transport Match Sharpness

After embedding a local disk, a direction at one endpoint is matched to outgoing
directions at the neighboring endpoint by softmax weights.  For a row with
weights \(w_j\), the entropy

\[
H=-\sum_j w_j\log w_j
\]

and effective match count \(\exp(H)\) measure whether the transport is sharp or
diffuse.  Smaller entropy indicates a more decisive direction match.

```{r transport-sharpness, fig.height=6.8}
ggplot(transport.long,
       aes(x = local.embedding.method, y = value,
           fill = local.embedding.method)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(metric ~ case.id, scales = "free_y") +
  labs(
    title = "Soft transport sharpness by backend",
    subtitle = "Sharper transport has lower entropy and fewer effective matches, but too-sharp matching can be unstable on irregular disks.",
    x = "local embedding backend",
    y = "mean diagnostic value",
    fill = "backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

This plot tells us whether edge-KK changes the matching problem, not just the
embedding stress.  A useful backend should make directions geometrically more
coherent without collapsing all matches to arbitrary nearest neighbors.

# Polynomial-Probe Residuals

The transported Hessian matrix \(A\) is evaluated on simple coordinate
polynomial probes \(P\).  The reported residual is

\[
\frac{\|AP\|_F}{\|P\|_F}.
\]

For an ideal second-difference transport on a clean path or grid, constants and
linear probes should have very small residuals.  Quadratic probes are included
as a sensitivity diagnostic rather than a required null-space target for this
second-difference operator.

```{r polynomial-residuals, fig.height=7.2}
ggplot(residual.long,
       aes(x = local.embedding.method, y = plot.residual,
           fill = local.embedding.method)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(probe ~ case.id, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Polynomial-probe residuals by backend",
    subtitle = "Residuals are ||AP||F / ||P||F; zero residuals are floored only for log-scale display.",
    x = "local embedding backend",
    y = "residual (log scale)",
    fill = "backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

This diagnostic is the mathematical sanity check.  A backend that reduces edge
stress but inflates constant or linear residuals would be suspicious for the
transported-Hessian project.

# Nullity

The estimated nullity of \(A\) provides a coarse global view of how many
functions are left nearly unpenalized.  It is not a proof of the correct
polynomial null space, but it is useful for detecting dramatic overconstraint
or underconstraint after a backend change.

```{r nullity-comparison, fig.height=5.6}
ggplot(summary.ok,
       aes(x = local.embedding.method, y = nullity.estimate,
           fill = local.embedding.method)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_wrap(~ case.id, scales = "free_y", nrow = 2) +
  labs(
    title = "Estimated nullity by backend",
    subtitle = "Large backend-induced nullity shifts indicate that the transport geometry is changing the operator, not only the diagnostics.",
    x = "local embedding backend",
    y = "estimated nullity",
    fill = "backend"
  ) +
  base.theme(11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The nullity plot should be read together with the residual plot.  A larger
nullity is not automatically better; the unpenalized functions need to align
with the intended low-order polynomial probes.

# Results Summary And Discussion

```{r learned-table, results='asis'}
best.edge <- summary.ok[order(summary.ok$case.id,
                              summary.ok$median.edge.stress,
                              summary.ok$median.graph.distance.stress), ]
best.edge <- best.edge[!duplicated(best.edge$case.id),
                       c("case.id", "local.embedding.method",
                         "median.edge.stress", "median.graph.distance.stress",
                         "residual.linear.mean", "nullity.estimate",
                         "backend.used", "fallback.reason")]
knitr::kable(best.edge, digits = 4,
             caption = "Best edge-stress backend for each validation case.")
```

The backend-resolution diagnostic is the most important software result.  It
confirms whether the requested edge-KK methods are actually using a true
edge-restricted optimizer.  When they do, the stress plots answer the geometric
question: edge-KK should preferentially reduce local edge stress, while
all-pairs graph-distance stress may remain similar or sometimes worsen.

The transport-sharpness plot answers a different question.  It shows whether
the improved local edge geometry changes the soft transport weights used in the
Hessian rows.  A useful backend should make these weights more coherent without
making them arbitrary.  The polynomial residuals then check whether those
transport changes are compatible with the intended low-order null behavior.

At this stage, the experiment supports using true edge-KK as the preferred
Phase 3 local coordinate repair only when the current `grip` checkout is loaded
or installed.  Without that API, the code now falls back transparently rather
than substituting full geodesic-KK, which would have changed the mathematical
meaning of the backend.

# What We Learned

1. The edge-KK backend distinction is observable and auditable.  The report
records the actual function used for every local disk.
2. Edge stress and graph-distance stress must be interpreted separately.
Edge-KK targets edge lengths, not all-pairs graph geodesic distances.
3. Transport entropy and polynomial residuals are necessary companion
diagnostics.  Lower edge stress alone is not enough to validate the transported
Hessian operator.
4. The next implementation step should not introduce new Hessian algebra yet.
It should first decide whether the edge-KK local-coordinate backend improves
the residual/nullity behavior enough to become the default backend for Phase 3
experiments.

# Appendix

## Run Metadata

```{r appendix-metadata}
str(metadata)
```

## Summary Table

```{r appendix-summary}
knitr::kable(summary.table, digits = 5)
```

## Backend Status Table

```{r appendix-backend-status}
knitr::kable(backend.status)
```

## Reproducibility

```{r appendix-command, results='asis'}
cat("Render command:\n\n")
cat("```bash\n")
cat("Rscript dev/graph-trend-filtering/render_transported_hessian_backend_validation.R\n")
cat("```\n\n")
cat("Optional full mode:\n\n")
cat("```bash\n")
cat("GFLOW_TRANSPORTED_HESSIAN_BACKEND_MODE=full Rscript dev/graph-trend-filtering/render_transported_hessian_backend_validation.R\n")
cat("```\n")
```
