---
title: "Transported Hessian Regression-Gradient Validation"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11.2,
  fig.height = 6.8,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
dev.dir <- file.path(repo.dir, "dev/graph-trend-filtering")
report.dir <- file.path(dev.dir, "reports")
source(file.path(dev.dir, "transported_hessian_regression_gradient_validation.R"))
library(ggplot2)

mode <- Sys.getenv("GFLOW_REGRESSION_GRADIENT_VALIDATION_MODE", unset = "smoke")
rds.path <- file.path(report.dir, paste0("transported_hessian_regression_gradient_validation_", mode, ".rds"))
diagnostics <- readRDS(rds.path)
metadata <- diagnostics$metadata
policy.table <- diagnostics$policies
summary.table <- diagnostics$summary
gradient.table <- diagnostics$gradient %||% data.frame()
embedding.table <- diagnostics$embedding %||% data.frame()

summary.ok <- subset(summary.table, status == "ok")
summary.ok$policy.family <- factor(summary.ok$policy.family,
                                   levels = c("exact", "soft",
                                              "regression_gradient_ambient",
                                              "regression_gradient_local",
                                              "regression_gradient_adaptive"))
summary.ok$policy.label <- factor(summary.ok$policy.label,
                                  levels = policy.table$policy.label)
summary.ok$graph.label <- paste(summary.ok$case.id, summary.ok$graph.type,
                                sep = " | ")
summary.ok$gradient.embedding.method <- ifelse(
  is.na(summary.ok$gradient.embedding.method),
  "none",
  summary.ok$gradient.embedding.method
)
summary.ok$gradient.disk.rule <- ifelse(
  is.na(summary.ok$gradient.disk.rule),
  "hops",
  summary.ok$gradient.disk.rule
)
summary.ok$gradient.disk.hops.factor <- factor(summary.ok$gradient.disk.hops)
summary.ok$gradient.disk.label <- ifelse(
  summary.ok$gradient.disk.rule == "metric.diameter.fraction",
  paste0("r=", summary.ok$gradient.disk.radius.fraction, "D"),
  ifelse(
    summary.ok$gradient.disk.rule == "metric.local.scale",
    paste0("s=", summary.ok$gradient.disk.local.scale.multiplier),
    paste0("h=", summary.ok$gradient.disk.hops)
  )
)
summary.ok$gradient.disk.label <- factor(
  summary.ok$gradient.disk.label,
  levels = unique(summary.ok$gradient.disk.label)
)
summary.ok$local.chart.label <- ifelse(
  summary.ok$policy.family == "regression_gradient_local",
  paste0(summary.ok$gradient.embedding.method, " | ", summary.ok$gradient.disk.label),
  as.character(summary.ok$policy.label)
)
fixed.local.summary <- subset(summary.ok, policy.family == "regression_gradient_local")
adaptive.summary <- subset(summary.ok, policy.family == "regression_gradient_adaptive")
summary.ok$plot.residual.linear <- pmax(
  summary.ok$residual.linear.mean,
  min(summary.ok$residual.linear.mean[
    is.finite(summary.ok$residual.linear.mean) &
      summary.ok$residual.linear.mean > 0
  ], na.rm = TRUE) / 2
)

nonzero.floor <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (length(x)) min(x) / 2 else 1e-12
}

residual.long <- rbind(
  data.frame(summary.ok[, c("case.id", "graph.type", "graph.label",
                            "policy.label", "policy.family")],
             probe = "constant",
             residual = summary.ok$residual.constant),
  data.frame(summary.ok[, c("case.id", "graph.type", "graph.label",
                            "policy.label", "policy.family")],
             probe = "linear mean",
             residual = summary.ok$residual.linear.mean),
  data.frame(summary.ok[, c("case.id", "graph.type", "graph.label",
                            "policy.label", "policy.family")],
             probe = "quadratic mean",
             residual = summary.ok$residual.quadratic.mean),
  data.frame(summary.ok[, c("case.id", "graph.type", "graph.label",
                            "policy.label", "policy.family")],
             probe = "overall",
             residual = summary.ok$residual.overall)
)
residual.long$plot.residual <- pmax(residual.long$residual,
                                    nonzero.floor(residual.long$residual))

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(color = "grey25", margin = margin(b = 10)),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.title.x = element_text(margin = margin(t = 8))
    )
}
```

# Purpose

The transported-Hessian project now has three qualitatively different transport
mechanisms:

\[
  \delta_{u'\to v'} f-\delta_{u\to v} f
\]

from direction matching, and a regression-gradient construction.  The
regression-gradient construction estimates, at each vertex \(u\),

\[
  f(w)-f(u)\approx g_u^\top(z_w-z_u),
  \qquad w\in N(u),
\]

by local weighted least squares.  Since the fitted gradient is linear in the
response vector \(f\), we can write

\[
  g_u = B_u f.
\]

The regression-gradient transported Hessian then compares gradient components
across graph darts:

\[
  A_{u\to u',k} f
  =
  (B_{u'}f)_k-(B_u f)_k.
\]

This report asks whether that operator has the expected null-space behavior
before we use it in a trend-filtering fit.  The ambient variant computes
gradients from supplied sampling coordinates.  The graph-derived variant builds
a shared local chart around each base dart and estimates both endpoint
gradients in that chart.  In particular, constants and affine coordinate probes
should be annihilated, while quadratic probes should generally be detected.

```{r metadata-summary, results='asis'}
cat(sprintf(
  "**Build:** %s  \n**Mode:** %s  \n**Policies tested:** %d  \n**Runs:** %d\n",
  metadata$generated.at,
  metadata$mode,
  nrow(policy.table),
  nrow(summary.table)
))
```

# Polynomial Residuals

The core diagnostic is

\[
  R(P)=\frac{\|AP\|_F}{\|P\|_F},
\]

computed for constant, linear, quadratic, and combined probe matrices.  For a
valid second-difference-like operator, constant and linear residuals should be
small.  Quadratic residuals should not vanish unless the operator is too weak.

```{r polynomial-residuals, fig.height=9.0}
ggplot(residual.long,
       aes(x = policy.label, y = plot.residual,
           fill = policy.family)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(probe ~ graph.label, scales = "free_y") +
  scale_y_log10() +
  scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Polynomial-probe residuals by transported-Hessian construction",
    subtitle = "Constants and linear probes test affine null-space behavior; residuals are log-scaled.",
    x = "transport policy",
    y = expression(norm(A * P, "F") / norm(P, "F")),
    fill = "policy family"
  ) +
  base.theme(9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

This is the main mathematical check.  The regression-gradient operator should
be read primarily through the constant and linear panels.  A low quadratic
residual is not automatically good: it can mean the operator is insensitive to
curvature, especially if row counts or local ranks are poor.

# Operator Size And Nullity

The same residual can mean different things if two operators have very
different numbers of rows.  We therefore inspect row count and estimated
nullity alongside the residuals.

```{r size-nullity, fig.height=7.6}
size.long <- rbind(
  data.frame(summary.ok[, c("graph.label", "policy.label", "policy.family")],
             diagnostic = "rows",
             value = summary.ok$n.rows),
  data.frame(summary.ok[, c("graph.label", "policy.label", "policy.family")],
             diagnostic = "nullity",
             value = summary.ok$nullity.estimate)
)
ggplot(size.long,
       aes(x = policy.label, y = value,
           fill = policy.family)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(diagnostic ~ graph.label, scales = "free_y") +
  scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Operator row count and estimated nullity",
    subtitle = "Residual improvements are meaningful only when the operator remains sufficiently constrained.",
    x = "transport policy",
    y = "count",
    fill = "policy family"
  ) +
  base.theme(9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

For regression-gradient transport, the row count is approximately
\(|\overrightarrow E|m\), where \(m\) is the coordinate dimension.  Nullity is
a coarse check that the operator has not become nearly empty.

# Local Regression Diagnostics

Regression-gradient transport depends on the local least-squares design.  At
vertex \(u\), the local design matrix has rows \(z_w-z_u\) for neighbors
\(w\in N(u)\).  The key diagnostics are:

\[
  \operatorname{rank}(X_u),
  \qquad
  \kappa(X_u^\top W_u X_u+\lambda I),
  \qquad
  |N(u)|.
\]

Low rank means the local gradient is not identifiable in all coordinate
directions.  Large condition number means the gradient estimate is numerically
unstable.

```{r local-regression-diagnostics, fig.height=7.6}
grad.summary <- subset(summary.ok,
                       policy.family %in% c("regression_gradient_ambient",
                                            "regression_gradient_local"))
grad.long <- rbind(
  data.frame(grad.summary[, c("graph.label", "policy.label")],
             diagnostic = "rank-deficient fraction",
             value = grad.summary$rank.deficient.fraction),
  data.frame(grad.summary[, c("graph.label", "policy.label")],
             diagnostic = "median condition",
             value = grad.summary$median.local.condition),
  data.frame(grad.summary[, c("graph.label", "policy.label")],
             diagnostic = "median neighborhood size",
             value = grad.summary$median.local.neighborhood)
)
ggplot(grad.long,
       aes(x = policy.label, y = value,
           fill = diagnostic)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_grid(diagnostic ~ graph.label, scales = "free_y") +
  labs(
    title = "Local least-squares diagnostics for regression-gradient transport",
    subtitle = "Rank and conditioning explain when gradient transport should or should not be trusted.",
    x = "regression-gradient policy",
    y = "diagnostic value",
    fill = "diagnostic"
  ) +
  base.theme(9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

This diagnostic is specific to Phase 3c.  It tells us whether a poor residual
comes from the transport idea itself, the local chart, or a local gradient fit
that is not identifiable.

# Local Chart Stress Sweep

The graph-derived regression-gradient policies create one local embedding for
each base dart.  The local disk can now be constructed either by hop radius,
or by a weighted metric radius

\[
  r_v=\alpha\,\operatorname{diam}_G,
\]

where \(\operatorname{diam}_G\) is the weighted graph diameter.  Two stress
diagnostics are useful:

\[
  S_E
  =
  \frac{
    \sqrt{|E_\Gamma|^{-1}
      \sum_{(i,j)\in E_\Gamma}
      (\|\hat z_i-\hat z_j\|-\ell_{ij})^2}
  }{\operatorname{mean}_{(i,j)\in E_\Gamma}\ell_{ij}},
\]

which measures edge-length preservation, and a corresponding all-pairs graph
distance stress inside the local disk.  For synthetic validation, where
ambient coordinates are available, we also measure whether the local chart can
be affinely aligned back to those coordinates:

\[
  R_{\mathrm{aff}}(\Gamma)
  =
  \frac{\|Z_{\mathrm{amb}} - [\mathbf 1,\widehat Z]B\|_F}
       {\|Z_{\mathrm{amb}}-\bar Z_{\mathrm{amb}}\|_F}.
\]

This is not a graph-only diagnostic, but it tells us what the graph-only
stress diagnostics miss.

```{r local-chart-stress, fig.height=7.4}
if (nrow(fixed.local.summary)) {
  stress.long <- rbind(
    data.frame(fixed.local.summary[, c("graph.label", "gradient.embedding.method",
                                       "gradient.disk.rule", "gradient.disk.label")],
               diagnostic = "edge stress",
               value = fixed.local.summary$median.embedding.edge.stress),
    data.frame(fixed.local.summary[, c("graph.label", "gradient.embedding.method",
                                       "gradient.disk.rule", "gradient.disk.label")],
               diagnostic = "graph-distance stress",
               value = fixed.local.summary$median.embedding.distance.stress),
    data.frame(fixed.local.summary[, c("graph.label", "gradient.embedding.method",
                                       "gradient.disk.rule", "gradient.disk.label")],
               diagnostic = "ambient affine residual",
               value = fixed.local.summary$median.ambient.affine.residual)
  )
  ggplot(stress.long,
         aes(x = gradient.disk.label, y = value,
             fill = gradient.embedding.method,
             alpha = gradient.disk.rule)) +
    geom_col(width = 0.72, color = "grey25") +
    facet_grid(diagnostic ~ graph.label, scales = "free_y") +
    scale_alpha_manual(values = c(hops = 0.95,
                                  "metric.diameter.fraction" = 0.70,
                                  "metric.local.scale" = 0.82),
                       na.translate = FALSE) +
    labs(
      title = "Local embedding stress for graph-derived regression-gradient transport",
      subtitle = "Each bar is the median across base-dart charts for one backend and disk rule.",
      x = "gradient disk",
      y = "median relative stress",
      fill = "local embedding backend",
      alpha = "disk rule"
    ) +
    base.theme(9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  plot.new()
  text(0.5, 0.5, "No graph-derived regression-gradient policies were available.")
}
```

This plot should be read together with the polynomial residuals.  A local chart
can have excellent edge stress and still fail to provide a coordinate system in
which ambient affine functions look affine.  That is precisely why the report
now includes both graph-only and synthetic-coordinate diagnostics.

# Disk-Radius Sensitivity

For the graph-derived regression-gradient operator, the local disk radius is a
modeling choice.  A one-hop disk often gives the gradient fit enough neighbors,
but not necessarily enough geometric context to orient the chart.  A
metric-diameter-fraction disk instead asks for all vertices within weighted
graph distance \(\alpha\operatorname{diam}_G\) of the base dart endpoints,
subject to the same maximum-chart-size cap.  A local-scale metric disk uses

\[
  r_{u,u'}=\alpha\max(s_u,s_{u'}),
\]

where \(s_u\) is the local weighted-graph \(k\)-nearest-neighbor distance.
We therefore compare \(h=1,2,3,4,5\) hop disks,
\(\alpha\in\{0.05,0.075,0.10,0.15,0.20\}\) graph-diameter disks, and
local-scale multipliers \(\alpha\in\{1,1.5,2,3,4\}\) for each backend.

```{r disk-radius-sensitivity, fig.height=8.0}
if (nrow(fixed.local.summary)) {
  ggplot(fixed.local.summary,
         aes(x = gradient.disk.label,
             y = pmax(residual.linear.mean,
                      nonzero.floor(residual.linear.mean)),
             color = gradient.embedding.method,
             linetype = gradient.disk.rule,
             group = interaction(gradient.embedding.method,
                                 gradient.disk.rule))) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2.2) +
    facet_wrap(~ graph.label, scales = "free_y", ncol = 2) +
    scale_y_log10() +
    labs(
      title = "Affine residual sensitivity to graph-disk radius",
      subtitle = "Lower is better.  Lines separate hop, graph-diameter, and local-scale metric disks.",
      x = "gradient disk",
      y = expression(R[linear]),
      color = "local embedding backend",
      linetype = "disk rule"
    ) +
    base.theme(10) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  plot.new()
  text(0.5, 0.5, "No graph-derived regression-gradient policies were available.")
}
```

If increasing the disk radius improves affine residuals, then chart context is
the bottleneck.  If edge stress improves but affine residuals get worse, the
backend is preserving edges without preserving the coordinate structure needed
by gradient transport.

# Disk Size And Locality

Larger disks are not free. They may help orient a chart, but they also make
the chart less local. The next diagnostic records the selected disk size as a
fraction of the graph and its weighted graph radius from the base dart.

```{r disk-size-locality, fig.height=7.8}
if (nrow(fixed.local.summary)) {
  locality.long <- rbind(
    data.frame(fixed.local.summary[, c("graph.label", "gradient.embedding.method",
                                       "gradient.disk.rule", "gradient.disk.label")],
               diagnostic = "disk fraction",
               value = fixed.local.summary$median.disk.fraction),
    data.frame(fixed.local.summary[, c("graph.label", "gradient.embedding.method",
                                       "gradient.disk.rule", "gradient.disk.label")],
               diagnostic = "weighted radius",
               value = fixed.local.summary$median.disk.weighted.radius)
  )
  ggplot(locality.long,
         aes(x = gradient.disk.label, y = value,
             color = gradient.embedding.method,
             linetype = gradient.disk.rule,
             group = interaction(gradient.embedding.method,
                                 gradient.disk.rule))) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2.2) +
    facet_grid(diagnostic ~ graph.label, scales = "free_y") +
    labs(
      title = "Disk size and chart locality diagnostics",
      subtitle = "The adaptive selector should not improve affine residuals merely by making every chart global.",
      x = "gradient disk",
      y = "median locality diagnostic",
      color = "local embedding backend",
      linetype = "disk rule"
    ) +
    base.theme(9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  plot.new()
  text(0.5, 0.5, "No fixed graph-derived policies were available.")
}
```

# Chart-Level Predictors

The next plot compares chart-level synthetic diagnostics with the global
operator residual.  The diagnostic
\[
  R_{\nabla,\mathrm{lin}}(\Gamma,u,u')
\]
is the mean absolute difference between the two endpoint gradient estimates
when the response is an ambient coordinate function.  This is the most direct
chart-level version of the affine-null-space test.

```{r chart-level-predictors, fig.height=7.8}
if (nrow(fixed.local.summary)) {
  ggplot(fixed.local.summary,
         aes(x = pmax(median.chart.linear.gradient.residual,
                      nonzero.floor(median.chart.linear.gradient.residual)),
             y = pmax(residual.linear.mean,
                      nonzero.floor(residual.linear.mean)),
             color = gradient.embedding.method,
             shape = gradient.disk.rule)) +
    geom_point(size = 2.6, alpha = 0.9) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                color = "grey45") +
    facet_wrap(~ graph.label, scales = "free", ncol = 2) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
      title = "Chart-level affine-gradient residual predicts operator affine residual",
      subtitle = "Points near the dashed line indicate the chart-level diagnostic is explaining the global operator behavior.",
      x = expression("median chart " * R[nabla * "," * lin]),
      y = expression("operator " * R[linear]),
      color = "local embedding backend",
      shape = "disk rule"
    ) +
    base.theme(10)
} else {
  plot.new()
  text(0.5, 0.5, "No graph-derived regression-gradient policies were available.")
}
```

This plot is the bridge toward adaptive chart selection.  A useful graph-only
rule should eventually predict this chart-level residual without needing
ambient coordinates.

# Adaptive Chart Selection

The adaptive policy tries several graph-derived charts for each base dart and
chooses the one with the lowest graph-only score:

\[
  S
  =
  S_E + S_D
  + 10\,\mathrm{rank\_deficit}
  + 0.05\log(1+\kappa)
  + 0.05\,\mathrm{disk\_fraction}
  + 0.01\,\mathrm{weighted\_radius}.
\]

The score intentionally does not use ambient-coordinate residuals.  The report
uses ambient residuals only afterward to see whether the graph-only choice is
close to the validation oracle.

```{r adaptive-selection-distribution, fig.height=7.6}
adaptive.embedding <- subset(embedding.table,
                             policy.family == "regression_gradient_adaptive" &
                               chart.selected)
if (nrow(adaptive.embedding)) {
  adaptive.embedding$gradient.disk.rule <- ifelse(
    is.na(adaptive.embedding$gradient.disk.rule),
    "hops",
    adaptive.embedding$gradient.disk.rule
  )
  adaptive.embedding$selection.label <- ifelse(
    adaptive.embedding$gradient.disk.rule == "metric.diameter.fraction",
    paste0(adaptive.embedding$requested.method,
           " | r=", adaptive.embedding$gradient.disk.radius.fraction, "D"),
    ifelse(
      adaptive.embedding$gradient.disk.rule == "metric.local.scale",
      paste0(adaptive.embedding$requested.method,
             " | s=", adaptive.embedding$gradient.disk.local.scale.multiplier),
      paste0(adaptive.embedding$requested.method,
             " | h=", adaptive.embedding$gradient.disk.hops)
    )
  )
  ggplot(adaptive.embedding,
         aes(x = selection.label, fill = requested.method)) +
    geom_bar(color = "grey25") +
    facet_wrap(~ paste(case.id, graph.type, sep = " | "),
               scales = "free_y", ncol = 2) +
    labs(
      title = "Adaptive chart selections by graph case",
      subtitle = "Counts are base-dart charts selected by the graph-only score.",
      x = "selected backend and disk",
      y = "selected base-dart charts",
      fill = "backend"
    ) +
    base.theme(9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
} else {
  plot.new()
  text(0.5, 0.5, "No adaptive chart selections were recorded.")
}
```

```{r adaptive-versus-fixed, fig.height=7.4}
if (nrow(adaptive.summary) && nrow(fixed.local.summary)) {
  best.fixed <- fixed.local.summary[
    order(fixed.local.summary$case.id, fixed.local.summary$residual.linear.mean), ]
  best.fixed <- best.fixed[!duplicated(best.fixed$case.id), ]
  compare <- rbind(
    data.frame(best.fixed[, c("case.id", "graph.label")],
               policy = "best fixed local",
               residual = best.fixed$residual.linear.mean),
    data.frame(adaptive.summary[, c("case.id", "graph.label")],
               policy = "adaptive graph-only",
               residual = adaptive.summary$residual.linear.mean)
  )
  ggplot(compare,
         aes(x = policy,
             y = pmax(residual, nonzero.floor(residual)),
             fill = policy)) +
    geom_col(width = 0.72, color = "grey25") +
    facet_wrap(~ graph.label, scales = "free_y", ncol = 2) +
    scale_y_log10() +
    labs(
      title = "Adaptive graph-only selector versus best fixed local chart",
      subtitle = "Best fixed is selected retrospectively using operator affine residual; adaptive uses only graph diagnostics.",
      x = "policy",
      y = expression(R[linear]),
      fill = "policy"
    ) +
    base.theme(10) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1),
          legend.position = "none")
} else {
  plot.new()
  text(0.5, 0.5, "Adaptive or fixed local summaries were unavailable.")
}
```

# Coordinate Source

The next figure isolates the regression-gradient policies and compares ambient
coordinate gradients against graph-derived local-chart gradients.

```{r coordinate-source, fig.height=7.4}
rg.coord <- subset(summary.ok,
                   policy.family %in% c("regression_gradient_ambient",
                                        "regression_gradient_local",
                                        "regression_gradient_adaptive"))
ggplot(rg.coord,
       aes(x = policy.label, y = plot.residual.linear,
           fill = policy.family)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_wrap(~ graph.label, scales = "free_y", ncol = 2) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Dark2", na.translate = FALSE) +
  labs(
    title = "Ambient versus graph-derived regression-gradient transport",
    subtitle = "The graph-derived policies estimate endpoint gradients in a shared local chart for each base dart.",
    x = "regression-gradient policy",
    y = expression(R[linear]),
    fill = "coordinate source"
  ) +
  base.theme(9) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The desirable first outcome is not that the graph-derived chart matches ambient
coordinates immediately on every sampled graph.  The desirable outcome is that
its failures are visible through local chart stress, local rank, and residuals.

# Ridge Sensitivity

The ridge parameter stabilizes local gradient estimation:

\[
  (X_u^\top W_u X_u+\lambda I)^{-1}X_u^\top W_u.
\]

The next figure isolates the regression-gradient policies and shows whether
changing \(\lambda\) materially changes the affine residual.

```{r ridge-sensitivity, fig.height=7.4}
ridge <- subset(summary.ok, policy.family == "regression_gradient_ambient")
ggplot(ridge,
       aes(x = factor(policy.label), y = plot.residual.linear,
           fill = policy.label)) +
  geom_col(width = 0.72, color = "grey25") +
  facet_wrap(~ graph.label, scales = "free_y", ncol = 2) +
  scale_y_log10() +
  labs(
    title = "Ridge sensitivity of ambient regression-gradient transport",
    subtitle = "Large sensitivity indicates ill-conditioned or rank-deficient local gradient fits.",
    x = "ridge setting",
    y = expression(R[linear]),
    fill = "ridge setting"
  ) +
  base.theme(10) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

Ridge sensitivity should be small on well-conditioned grid interiors and larger
near underdetermined or one-sided neighborhoods.

# Results Summary

```{r summary-table, results='asis'}
family.summary <- aggregate(cbind(residual.constant,
                                  residual.linear.mean,
                                  residual.quadratic.mean,
                                  n.rows,
                                  nullity.estimate) ~ policy.family,
                            data = summary.ok,
                            FUN = function(x) median(x, na.rm = TRUE))
knitr::kable(family.summary, digits = 4,
             caption = "Median operator diagnostics by policy family.")

best <- summary.ok[order(summary.ok$case.id,
                         summary.ok$residual.linear.mean,
                         -summary.ok$n.rows), ]
best <- best[!duplicated(best$case.id),
             c("case.id", "policy.label", "policy.family",
               "residual.linear.mean", "residual.quadratic.mean",
               "n.rows", "nullity.estimate")]
knitr::kable(best, digits = 4,
             caption = "Lowest linear residual by graph case.")
```

```{r automatic-interpretation, results='asis'}
rg <- subset(summary.ok, policy.family %in% c("regression_gradient_ambient",
                                             "regression_gradient_local",
                                             "regression_gradient_adaptive"))
rg.ambient <- subset(summary.ok, policy.family == "regression_gradient_ambient")
rg.local <- subset(summary.ok, policy.family == "regression_gradient_local")
rg.adaptive <- subset(summary.ok, policy.family == "regression_gradient_adaptive")
soft <- subset(summary.ok, policy.family == "soft")
exact <- subset(summary.ok, policy.family == "exact")
rg.max.linear <- max(rg$residual.linear.mean, na.rm = TRUE)
rg.median.quad <- median(rg$residual.quadratic.mean, na.rm = TRUE)
rg.max.rank.def <- max(rg$rank.deficient.fraction, na.rm = TRUE)
ambient.max.linear <- max(rg.ambient$residual.linear.mean, na.rm = TRUE)
local.median.linear <- median(rg.local$residual.linear.mean, na.rm = TRUE)
local.min.linear <- min(rg.local$residual.linear.mean, na.rm = TRUE)
local.median.edge.stress <- median(rg.local$median.embedding.edge.stress, na.rm = TRUE)
adaptive.median.linear <- median(rg.adaptive$residual.linear.mean, na.rm = TRUE)
adaptive.oracle.fraction <- median(rg.adaptive$selected.oracle.fraction, na.rm = TRUE)
soft.median.linear <- median(soft$residual.linear.mean, na.rm = TRUE)
exact.errors <- subset(summary.table, policy.family == "exact" & status != "ok")

cat("## What We Learned\n\n")
cat(sprintf(
  "The ambient regression-gradient operator remains the clean reference for this family: its largest recorded mean linear residual was %.3g.  Across all regression-gradient policies, the median quadratic residual was %.3g, so the construction still detects curvature rather than collapsing to a vacuous operator.\n\n",
  ambient.max.linear,
  rg.median.quad
))
cat(sprintf(
  "The soft direction-matching family had median mean linear residual %.3g in the same validation table, which confirms that Phase 3c is not merely another gate on the soft matching method; it changes the operator behavior qualitatively.\n\n",
  soft.median.linear
))
cat(sprintf(
  "The graph-derived local-chart version is now implemented and auditable.  Its median mean linear residual was %.3g, its best recorded mean linear residual was %.3g, and its median local edge stress was %.3g.  This tells us that local chart quality is now the main bottleneck: path-like charts behave almost exactly, while grid and sampled neighborhoods need better local coordinate construction or larger local disks before they can match ambient gradients.\n\n",
  local.median.linear,
  local.min.linear,
  local.median.edge.stress
))
best.local <- rg.local[order(rg.local$residual.linear.mean), ][1, ]
cat(sprintf(
  "The best graph-derived local policy in this run was %s on %s, with mean linear residual %.3g, median ambient-affine chart residual %.3g, and median chart-level affine-gradient residual %.3g.  These chart diagnostics are now the main evidence source for deciding whether a backend/radius is worth using adaptively.\n\n",
  best.local$policy.label,
  best.local$case.id,
  best.local$residual.linear.mean,
  best.local$median.ambient.affine.residual,
  best.local$median.chart.linear.gradient.residual
))
if (nrow(rg.adaptive)) {
  cat(sprintf(
    "The adaptive graph-only selector had median mean linear residual %.3g, and its selected chart matched the synthetic validation oracle for a median fraction %.3g of base darts.  This is intentionally not expected to beat the retrospective best fixed policy yet; the goal is to test whether graph-only diagnostics can make reasonable local choices without ambient-coordinate leakage.\n\n",
    adaptive.median.linear,
    adaptive.oracle.fraction
  ))
}
cat(sprintf(
  "The largest rank-deficient fraction reported for regression-gradient local designs was %.3g.  Rank was therefore not the dominant failure mode in this experiment; chart distortion and the choice of local neighborhood explain more of the residual behavior.\n\n",
  rg.max.rank.def
))
if (nrow(exact.errors)) {
  cat(sprintf(
    "Exact coordinate transport failed in %d runs, all for non-axis-aligned jittered or sampled graphs.  These failures are expected and reinforce why exact matching is only a reference construction.\n\n",
    nrow(exact.errors)
  ))
}
cat("The main caveat is that the adaptive score is still hand-weighted.  The next methodological question is whether the graph-only score should be calibrated against synthetic chart residuals, or whether the score should include a more direct chart-orientation diagnostic that remains available without ambient coordinates.\n")
```

The validation report should be read as an operator check, not as a final
regression benchmark.  A promising Phase 3c outcome is:

1. constant and linear residuals are near zero on path and grid references;
2. sampled and jittered graphs expose rank or conditioning issues rather than
   mysterious failures;
3. quadratic residuals remain nonzero, so the operator still detects curvature;
4. ridge sensitivity is interpretable and not dominant on well-conditioned
   graphs.

# Appendix

## Run Metadata

```{r appendix-metadata}
str(metadata)
```

## Policies

```{r appendix-policies}
knitr::kable(policy.table, digits = 4)
```

## Summary Table

```{r appendix-summary}
knitr::kable(summary.table, digits = 5)
```

## Embedding Table

```{r appendix-embedding}
if (nrow(embedding.table)) {
  knitr::kable(embedding.table, digits = 5)
} else {
  cat("No embedding diagnostics were recorded.")
}
```

## Error Runs

```{r appendix-errors}
err <- subset(summary.table, status != "ok")
if (nrow(err)) {
  knitr::kable(err[, c("case.id", "policy.label", "status", "message")])
} else {
  cat("No error runs.")
}
```

## Reproducibility

```{r appendix-command, results='asis'}
cat("Smoke render:\n\n")
cat("```bash\n")
cat("Rscript dev/graph-trend-filtering/render_transported_hessian_regression_gradient_validation.R\n")
cat("```\n\n")
cat("Full render:\n\n")
cat("```bash\n")
cat("GFLOW_REGRESSION_GRADIENT_VALIDATION_MODE=full Rscript dev/graph-trend-filtering/render_transported_hessian_regression_gradient_validation.R\n")
cat("```\n")
```
