---
title: "Canonical Basin Complex Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Canonical Basin Complex Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(gflow)
```

## One constructor, five basin families

`create.basin.complex()` is the only constructor for new basin objects in
`gflow`. It validates the graph and field once, dispatches to one of five basin
construction families, and returns the same `basin_complex` schema from every
family.

| `method` | Structure computed | Typical use |
|---|---|---|
| `"trajectory_flow"` | Flow trajectories and terminal basins | Deterministic field-following basins |
| `"superlevel_merge_tree"` | Plateau-aware graph level-set merge tree | Basin births, mergers, and persistence |
| `"geodesic_reachability"` | Monotone reachability supports | Overlapping attraction supports |
| `"rtcb"` | Robust trajectory-consensus basins | Stability under local trajectory variation |
| `"overlap_cell_complex"` | Ascending-by-descending overlap cells | Joint maxima/minima decomposition |

The remainder of this vignette uses one small two-dimensional grid only to make
the interface executable and inspectable. Scientific basin validation should
use the higher-dimensional, scale-path inputs appropriate to the analysis.

## Graph and field

The graph is undirected. Each adjacency entry has a parallel edge-length entry,
and vertex identifiers are one-based.

```{r graph}
edges <- matrix(
  c(
    1, 2, 1, 1, 4, 1, 2, 3, 1, 2, 5, 1,
    3, 6, 1, 4, 5, 1, 4, 7, 1, 5, 6, 1,
    5, 8, 1, 6, 9, 1, 7, 8, 1, 8, 9, 1
  ),
  ncol = 3,
  byrow = TRUE
)

adjacency <- edge_lengths <- vector("list", 9)
for (i in seq_len(nrow(edges))) {
  from <- edges[i, 1]
  to <- edges[i, 2]
  adjacency[[from]] <- c(adjacency[[from]], to)
  adjacency[[to]] <- c(adjacency[[to]], from)
  edge_lengths[[from]] <- c(edge_lengths[[from]], edges[i, 3])
  edge_lengths[[to]] <- c(edge_lengths[[to]], edges[i, 3])
}

field <- c(0, 1, 0, 1, 3, 1, 0, 1, 2)
vertex_mass <- rep(1 / length(field), length(field))
```

## Direction, mass, and density

`direction = "max"` constructs ascending basins, `"min"` constructs descending
basins, and `"both"` requests both orientations. The overlap-cell method
requires `"both"`.

`vertex.mass` represents empirical or quadrature support. It is used for basin
mass summaries and mass-threshold refinement. `vertex.density` instead
modulates trajectory edge scores when `method.params$modulation` is
`"DENSITY"` or `"DENSITY_EDGELEN"`. The constructor never substitutes one for
the other.

```{r trajectory-flow}
flow <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "trajectory_flow",
  direction = "max",
  vertex.mass = vertex_mass,
  method.params = list(
    modulation = "CLOSEST",
    edge.length.quantile.thld = 1
  )
)

flow$status
flow$basin.table[
  ,
  c("basin.id", "type", "retained.support.mass")
]
```

For density-modulated flow, provide density explicitly:

```{r density-flow}
density_flow <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "trajectory_flow",
  direction = "max",
  vertex.density = exp(field),
  method.params = list(
    modulation = "DENSITY",
    edge.length.quantile.thld = 1
  )
)
density_flow$status
```

## Level-set merge tree

The level-set method contracts exact field plateaus and records component
births, elder-rule mergers, and component-root deaths. It is the canonical
method for following basin topology across a one-parameter conditional-
expectation scale path.

```{r merge-tree}
tree <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "superlevel_merge_tree",
  direction = "max"
)

tree$status
tree$merge.table
```

Construct one object at every retained scale value, then compare merge tables
and memberships between adjacent scale values. The scale parameter belongs to
the conditional-expectation estimator, not to `create.basin.complex()` itself;
record it alongside each returned object.

## Reachability, RTCB, and overlap cells

```{r remaining-methods}
reachability <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "geodesic_reachability",
  direction = "max",
  method.params = list(edge.length.quantile.thld = 1)
)

consensus <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "rtcb",
  direction = "max",
  method.params = list(
    edge.length.quantile.thld = 1,
    n.min = 1L,
    m.min = 0
  )
)

cells <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "overlap_cell_complex",
  direction = "both"
)

vapply(
  list(
    geodesic_reachability = reachability,
    rtcb = consensus,
    overlap_cell_complex = cells
  ),
  function(x) x$status,
  character(1)
)
```

## Membership is not assignment

`membership.table` is the authoritative many-to-many relation between vertices
and basins. Overlap is allowed. `primary.assignment` is an optional convenience
label selected by the method's `primary.assignment.policy`; it does not replace
membership and may be absent.

```{r membership}
head(reachability$membership.table)
reachability$primary.assignment
```

Downstream code that needs complete basin supports should use
`membership.table`. Code that requires exactly one basin per vertex must choose
and record an assignment policy.

## Refinement

Refinement stages are controlled with `simplify.params`. They execute in a
fixed order and are recorded in `provenance$refinement.stages`:

1. relative-value filtering;
2. extrema clustering;
3. geometric filtering;
4. support filtering; and
5. expansion.

Stage applicability depends on the construction family. Unsupported
combinations fail validation rather than silently doing nothing.

```{r refinement}
refined <- create.basin.complex(
  adjacency,
  edge_lengths,
  field,
  method = "trajectory_flow",
  direction = "max",
  method.params = list(edge.length.quantile.thld = 1),
  simplify.params = list(
    relative.value = list(enabled = TRUE)
  )
)

refined$provenance$refinement.stages
```

## Archived objects and retired constructors

The following public constructors are retired and deliberately raise a
`gflow_basin_lifecycle_error`:

* `compute.gfc()`;
* `compute.basins.of.attraction()`;
* `compute.gfc.trajectory()`;
* `compute.gfc.flow()`; and
* `create.basin.cx()`.

Use `create.basin.complex()` for all new work. `as.basin.complex()` exists only
for compatible archived results. Conversion requires the archived object and
its original graph; a field must also be supplied if the archive did not retain
one. Conversion provenance records the source class and retained archive so
that migration remains auditable.

```{r conversion-interface, eval=FALSE}
canonical <- as.basin.complex(
  archived_object,
  adj.list = original_adjacency,
  edge.length.list = original_edge_lengths,
  field = original_field
)
```

## Result checks

Every result carries construction status, resolved parameters, provenance,
diagnostics, basin summaries, membership, and optional method-specific
structures.

```{r result-checks}
objects <- list(flow, tree, reachability, consensus, cells)
stopifnot(all(vapply(
  objects,
  function(x) inherits(x, "basin_complex") && identical(x$status, "ok"),
  logical(1)
)))

names(flow)
flow$provenance$construction
```
