---
title: "Example graphs and scalar fields"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Example graphs and scalar fields}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 7, fig.height = 4)
library(gflow)
```

Small deterministic graphs make basin definitions visible. Here we separate
the graph, its edge lengths, and its scalar field, then compare selected
methods under stated controls. These examples verify interfaces and illustrate
structure; they do not validate performance on noisy or high-dimensional data.
See [Finding your way around gflow](function-guide.html) for the API catalog
and [the canonical workflow](basin_complex_workflow_vignette.html) for all five
construction families.

## Recipes at a glance

| Example | Required inputs | Question and entry points | Output |
|---|---|---|---|
| Seven-vertex path | Explicit adjacency, unit lengths, two peaks including an exact plateau | How do exact level sets and terminal flow differ? `create.basin.complex()`, `cut()` | Basin tables, branch hierarchy, assignments |
| Nine-vertex grid | Four-neighbor edges, unit lengths, peaks separated by level 1 | Which supports overlap? Compare trajectory flow and reachability; refine support | Membership rows and retained-support summaries |
| Two paths and an isolate | Six-vertex disconnected adjacency and field | Does the level-set hierarchy preserve components? | Three component roots with separately defined lifetimes |
| Noisy circle (linked below) | Seeded noisy coordinates, known generating field, observed response, selected dgraphs graph | How do graph choice and noise affect a basin/association analysis? | Refined basins and local-association summaries |

There are no public scientific datasets bundled in this package. The installed
`extdata` code manifest identifies the build; it is not a dataset. These recipes
define their inputs directly and add no generator API or external download.

## A path with two peaks and a plateau

List position is the internal vertex index. Every neighbor has a corresponding
edge length in the same position, and edges appear in both directions. Field
height and edge length use independent units. The drawing's horizontal
coordinate is vertex order; it is not an extra constructor input.

```{r path-input}
adjacency <- list(2L, c(1L, 3L), c(2L, 4L), c(3L, 5L),
                  c(4L, 6L), c(5L, 7L), 6L)
edge_lengths <- lapply(adjacency, function(v) rep(1, length(v)))
field <- c(0, 3, 3, 1, 2, 4, 0)
vertex_mass <- c(1, 2, 2, 1, 1, 2, 1)
vertex_density <- c(1, 1, 1, 2, 3, 2, 1)
knitr::kable(data.frame(vertex = seq_along(field), field,
                        mass = vertex_mass, density = vertex_density),
             caption = "Independent field, mass, and density inputs.")
```

```{r path-plot, fig.cap="Path field: vertices 2 and 3 form one connected plateau at height 3; vertex 6 peaks at height 4. The two peaks connect through vertex 4 at height 1."}
plot(seq_along(field), field, type = "b", pch = 19,
     xlab = "Vertex (unit-length path)", ylab = "Scalar field (field units)",
     ylim = c(-0.3, 4.6))
text(seq_along(field), field, labels = seq_along(field), pos = 3)
```

Mass expresses empirical or quadrature support. The constructor normalizes it
for basin mass summaries. Density can change edge scoring in density-modulated
trajectory flow; it does not weight the level-set sweep. Neither is inferred
from the field. Here we use the same mass for both constructions and no density
modulation, so differences arise from the basin definitions.

```{r path-methods}
path_tree <- create.basin.complex(
  adjacency, edge_lengths, field, method = "superlevel_merge_tree",
  direction = "max", vertex.mass = vertex_mass
)
path_flow <- create.basin.complex(
  adjacency, edge_lengths, field, method = "trajectory_flow",
  direction = "max", vertex.mass = vertex_mass,
  method.params = list(modulation = "CLOSEST", edge.length.quantile.thld = 1)
)
stopifnot(identical(path_tree$status, "ok"),
          identical(path_flow$status, "ok"))
for (object in list(path_tree, path_flow)) {
  print(get.basin.table(object)[, c("extremum.vertex", "raw.support.size",
                                    "primary.support.size", "persistence")])
}
```

CLOSEST flow contracts the exact plateau and chooses terminal paths. Its two
supports contain 4 and 3 vertices. The exact tree's root support contains all
7 vertices, including the younger plateau branch of size 2. This hierarchical
overlap is legitimate; tree support is not a partition into terminal paths.

```{r path-cut}
tree <- get.basin.merge.tree(path_tree, required = TRUE)
above_two <- cut(tree, height = 2, direction = "max")
at_merge <- cut(tree, height = 1, direction = "max")
stopifnot(above_two$n.components == 2L, at_merge$n.components == 1L)
above_two$components[, c("extremum.vertex", "support.size", "vertices")]
get.basin.assignment(path_flow)[, c("vertex", "basin.id", "next.vertex")]
```

At height 2, the active components are `{2, 3}` and `{5, 6}`. At height 1
they merge. The younger branch persists for `3 - 1 = 2` field units. The root's
finite lifetime is `4 - 0 = 4` under the component-opposite-extreme convention.
This is a structural comparison, with no significance test.

## A grid with overlapping supports

Build a 3-by-3 grid with horizontal and vertical edges only. The coordinate
matrix is used to construct and draw those edges. It is separate from the
field and would not affect the analysis if merely redrawn.

```{r grid-input}
coordinates <- as.matrix(expand.grid(x = 0:2, y = 0:2))
grid_adjacency <- lapply(seq_len(nrow(coordinates)), function(i) {
  which(rowSums(abs(sweep(coordinates, 2, coordinates[i, ], "-"))) == 1)
})
grid_lengths <- lapply(grid_adjacency, function(v) rep(1, length(v)))
grid_field <- c(0, 1, 0, 1, 3, 1, 0, 1, 2)
```

```{r grid-plot, fig.width=5, fig.cap="Four-neighbor grid, labeled vertex:field. The peaks at vertices 5 and 9 connect through level-1 vertices 6 and 8; diagonal shortcuts are absent."}
plot(coordinates, type = "n", asp = 1, xlab = "Graph coordinate x",
     ylab = "Graph coordinate y", xlim = c(-0.2, 2.2), ylim = c(-0.2, 2.3))
for (i in seq_along(grid_adjacency)) {
  for (j in grid_adjacency[[i]][grid_adjacency[[i]] > i]) {
    segments(coordinates[i, 1], coordinates[i, 2],
             coordinates[j, 1], coordinates[j, 2], col = "grey60")
  }
}
points(coordinates, pch = 21, bg = "white", cex = 1.5)
text(coordinates, labels = paste(seq_along(grid_field), grid_field, sep = ":"),
     pos = 3, cex = 0.9)
```

Use all edge lengths for both methods, ascending direction, and the same
explicit primary-assignment policy. No refinement is applied at first.

```{r grid-compare}
grid_objects <- lapply(c("trajectory_flow", "geodesic_reachability"), function(m) {
  create.basin.complex(
    grid_adjacency, grid_lengths, grid_field, method = m, direction = "max",
    method.params = list(edge.length.quantile.thld = 1,
                        primary.assignment.policy = "largest_membership_then_extremum")
  )
})
names(grid_objects) <- c("Trajectory flow", "Geodesic reachability")
stopifnot(all(vapply(grid_objects, function(x) identical(x$status, "ok"), logical(1))))
comparison <- do.call(rbind, lapply(names(grid_objects), function(method) {
  object <- grid_objects[[method]]
  membership <- get.basin.membership(object)
  data.frame(method, extrema = paste(get.basin.table(object)$extremum.vertex,
                                     collapse = ", "),
             membership.rows = nrow(membership),
             overlapping.vertices = sum(table(membership$vertex) > 1),
             uncovered.vertices = length(setdiff(seq_along(grid_field),
                                                 membership$vertex)))
}))
knitr::kable(comparison, row.names = FALSE,
             col.names = c("Method", "Peak vertices", "Membership rows",
                           "Vertices in multiple basins", "Uncovered vertices"),
             caption = "Same grid and field, different basin definitions.")
get.basin.membership(grid_objects[[2]])
get.basin.assignment(grid_objects[[2]])[, c("vertex", "basin.id", "assignment.status")]
```

Both methods identify maxima 5 and 9 and cover this grid, but reachability
has more overlapping vertices. A single label discards that multiplicity.
Compare the membership rows first and interpret assignment as the selected
policy's convenience label. Agreement across methods is not the objective.

## Refine retained support

For reachability, retain only basins with at least six vertices. The supports
initially have sizes 8 and 5, so this removes the smaller one. Expansion can
then fill uncovered retained support along the graph.

```{r grid-refinement}
refined <- create.basin.complex(
  grid_adjacency, grid_lengths, grid_field,
  method = "geodesic_reachability", direction = "max",
  method.params = list(edge.length.quantile.thld = 1,
                      primary.assignment.policy = "largest_membership_then_extremum"),
  simplify.params = list(
    support.filter = list(enabled = TRUE, min.basin.size = 6L),
    expansion = list(enabled = TRUE)
  )
)
stopifnot(identical(refined$status, "ok"))
retained <- get.basin.table(refined)
knitr::kable(retained[, c("extremum.vertex", "retained", "raw.support.size",
                         "retained.support.size")],
             caption = "Refinement preserves raw support and changes retained support.")
raw_columns <- c("vertex", "direction", "basin.id", "membership.weight")
stopifnot(identical(get.basin.membership(refined)[, raw_columns],
                    get.basin.membership(grid_objects[[2]])[, raw_columns]))
refined$provenance$refinement.stages[, c("stage", "status")]
```

Raw membership remains a record of construction, even after retained support
changes. For current coverage, take the union of `retained.support.vertices`
over retained rows; for current labels, call `get.basin.assignment(refined)`.
The canonical workflow describes the other stages and their availability.

## Disconnected components and an isolated vertex

The exact merge-tree method supports disconnected graphs. These components
are path `1--2--3`, edge `4--5`, and isolated vertex `6`.

```{r disconnected}
disconnected_adjacency <- list(2L, c(1L, 3L), 2L, 5L, 4L, integer(0))
disconnected_lengths <- lapply(disconnected_adjacency, function(v) rep(1, length(v)))
disconnected_field <- c(0, 2, 1, 5, 4, 7)
disconnected <- create.basin.complex(
  disconnected_adjacency, disconnected_lengths, disconnected_field,
  method = "superlevel_merge_tree", direction = "max"
)
stopifnot(identical(disconnected$status, "ok"))
roots <- get.basin.table(disconnected)
stopifnot(nrow(roots) == 3L, all(is.na(roots$parent.basin.id)))
knitr::kable(roots[, c("extremum.vertex", "birth.level", "death.level", "persistence")],
             caption = "Each disconnected component has its own root lifetime.")
```

No branch joins across components. The isolated vertex is a constant component
with zero lifetime; its high absolute value does not give it large persistence.
Mass was omitted, so mass summaries are unavailable. This example establishes
component handling for exact trees, not for every trajectory backend.

## From these recipes to the noisy circle

The [noisy-circle vignette](noisy_circle_core_workflow_vignette.html) distinguishes
latent circle position, noisy coordinates, the known generating field, and
the noisy observed response. The graph is chosen from the noisy coordinates;
the basin field is the observed response, not an estimated conditional mean.
Its small permutation example demonstrates computation under an unrestricted
row-permutation null, with assumptions and limits explained alongside it.

For a new scientific application, record which field was supplied, how the
graph and lengths were chosen, the direction and method, all refinement
controls, mass/density provenance, and construction status. The examples here
need no optional renderer and open no browser or graphics window during builds.
