---
title: "Retina Development: A Single-Cell gflow Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retina Development: A Single-Cell gflow Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
library(gflow)
library(ggplot2)

retina_asset_path <- function(...) {
  rel <- file.path("retina_vignette", ...)
  candidates <- c(
    system.file("extdata", rel, package = "gflow"),
    file.path("inst", "extdata", rel),
    file.path("..", "inst", "extdata", rel)
  )
  for (path in candidates) {
    if (nzchar(path) && file.exists(path)) {
      return(normalizePath(path, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Could not locate retina vignette asset: ", rel)
}

read_retina_tsv <- function(...) {
  utils::read.delim(
    retina_asset_path(...),
    sep = "\t",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
}

manifest <- read_retina_tsv("manifest.tsv")
preprocessing.views <- read_retina_tsv("tables", "preprocessing_views.tsv")
pseudotime.full <- read_retina_tsv("tables", "pseudotime_sweep_full.tsv")
targeted.k.curve <- read_retina_tsv("tables", "targeted_k_curve.tsv")
targeted.k.selection <- read_retina_tsv("tables", "targeted_k_selection.tsv")
branch.selected.k <- read_retina_tsv("tables", "branch_selected_k.tsv")
```

## Overview

This vignette shows how `gflow` can be used on single-cell RNA-seq data rather
than only on the earlier microbiome-style examples. The target dataset is the
Clark *et al.* developing mouse retina atlas (`GSE118614`), and the analysis
unit is one row per retinal progenitor cell (RPC).

The goal is not to reproduce the entire retina project inside a package
vignette. Instead, the vignette isolates the reusable workflow:

1. freeze the analysis unit and preprocessing views,
2. build an intersection-kNN graph family with `gflow`,
3. compare graph scales explicitly,
4. validate geometry against developmental age and rooted pseudotime,
5. show that the "best" graph scale can depend on the downstream objective,
6. carry the selected geometry into one downstream biological summary.

This vignette is based on a curated retina source packet assembled during the
development of the generalized `gflow` examples. The package vignette extracts
the reusable single-cell workflow from that larger retina project rather than
trying to retell every retina-specific result.

## A Note on Scope

This vignette is intentionally narrower than the full retina analysis
workspace. The upstream retina analysis project already contains a richer
retina-specific body of work:

- dataset notes and biological context,
- pseudotime benchmark reports,
- target-gene `k`-selection reports,
- Notch/cell-cycle downstream module analyses.

Here we extract the most reusable `gflow` story from that larger project.

This version reads only package-owned assets from:

- `inst/extdata/retina_vignette`

Those assets are produced by:

- `inst/scripts/build_retina_vignette_assets.R`
- `data-raw/build_retina_vignette_assets.R`

so the vignette renders quickly while still keeping a rebuild path available.

## 1) Dataset Freeze and Analysis Unit

The current retina example is built around the public Clark retina atlas:

- source study: Clark *et al.* (Neuron, 2019)
- GEO series: `GSE118614`
- analysis unit: one RPC per row
- external validation variables:
  - developmental age
  - RPC subtype labels
  - rooted pseudotime

In the current package-adjacent retina workflow, the working analysis focuses on
QC-filtered RPC cells and compares several preprocessing views rather than
committing immediately to one representation.

## 2) Preprocessing Views

The current retina benchmark uses three views:

```{r preprocessing-view-table}
knitr::kable(
  preprocessing.views[, c("branch_id", "role", "description")],
  row.names = FALSE
)
```

This is already an important generalization step. The retina example is not a
single preprocessing recipe. It is a small multi-view graph benchmark.

## 3) Build the iKNN Graph Family

The core graph construction step uses the package API directly.

```{r build-iknn-graphs, eval = FALSE}
# X_view is a cells x features matrix for one chosen preprocessing branch.
# For example, it could be the PCA representation of log-normalized HVGs.

graphs <- create.iknn.graphs(
  X = X_view,
  kmin = 9L,
  kmax = 35L,
  compute.full = TRUE,
  pca.dim = NULL,
  n.cores = 1L,
  verbose = TRUE
)
```

Once the graph family is available, the first package-level stability summary is
degree-distribution Jensen-Shannon divergence:

```{r compute-stability, eval = FALSE}
stability <- compute.stability.metrics(
  graphs = graphs,
  graph.type = "geom",
  summary = "degree_distribution",
  divergence = "js"
)
```

This gives one explicit and reusable notion of graph-family stability. In the
retina example, it is combined with domain-aware validation rather than used as
the only criterion.

## 4) Current Package-Backed Retina Summary

The current package-backed retina rerun used `gflow::create.iknn.graphs()` as
the active graph constructor and produced the following smoke-scale summary for
the two completed broad geometry branches:

```{r current-retina-summary}
retina.summary <- branch.selected.k[, c(
  "branch_id",
  "selected_k",
  "age_spearman",
  "age_order_accuracy",
  "subtype_coherence"
)]
knitr::kable(retina.summary, digits = 4, row.names = FALSE)
```

The targeted marker-panel branch remains scientifically useful, but in the
current smoke rerun it behaved more like a performance-heavy validation branch
than a stable baseline branch. That is exactly why multi-view benchmarking is
useful: views can play different roles without being forced into one ranking
story.

## 5) Pseudotime-Oriented Graph Comparison

One strong current result is that the retina graph family can be compared by how
well rooted pseudotime aligns with developmental age.

```{r fig-pseudotime-spearman, fig.cap = "Full pseudotime sweep across HVG sets and graph sizes: Spearman correlation with developmental age rank."}
ggplot(
  pseudotime.full,
  aes(x = k, y = spearman_rho, color = hvg_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.3) +
  facet_wrap(~method, nrow = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "k",
    y = "Spearman correlation with age rank",
    color = "HVG set"
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

```{r fig-pseudotime-pearson, fig.cap = "Full pseudotime sweep across HVG sets and graph sizes: Pearson correlation with developmental age rank."}
ggplot(
  pseudotime.full,
  aes(x = k, y = pearson_r, color = hvg_label)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.3) +
  facet_wrap(~method, nrow = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "k",
    y = "Pearson correlation with age rank",
    color = "HVG set"
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

Two takeaways matter for the package story:

- the `3k` HVG view is a strong primary geometry,
- potential pseudotime gives stronger linear age alignment than geodesic on the
  same graph family.

This is a good example of why the vignette should show explicit graph
comparison, not just one selected graph.

## 6) Why One "Best k" Is the Wrong Question

The retina documents also show something more subtle and more important:
different scientific objectives can prefer different graph scales.

The pseudotime-oriented benchmark prefers a relatively dense regime, but the
target-gene conditional-expectation analysis over the union of Notch and
cell-cycle genes prefers a much smaller graph scale.

```{r fig-gcv-k-selection, fig.cap = "Conditional-expectation-driven k selection for the targeted Notch-plus-cell-cycle gene set."}
ggplot(targeted.k.curve, aes(x = k, y = median_norm)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#9ecae1", alpha = 0.45) +
  geom_line(linewidth = 0.9, color = "#08519c") +
  geom_point(size = 1.6, color = "#08519c") +
  geom_vline(
    data = subset(targeted.k.selection, criterion == "min_median_norm"),
    aes(xintercept = selected_k),
    color = "#cb181d",
    linetype = "dashed",
    linewidth = 0.7,
    inherit.aes = FALSE
  ) +
  labs(
    x = "k",
    y = "Median normalized GCV",
    title = "Target-gene conditional expectation prefers a smaller graph scale"
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

The current targeted-gene report recommends:

```{r targeted-k-summary}
targeted.k <- targeted.k.selection
targeted.k$interpretation <- ifelse(
  targeted.k$criterion == "min_median_norm",
  "Primary robust choice for the target gene set.",
  "Secondary reference influenced more strongly by heavy tails."
)
knitr::kable(targeted.k, row.names = FALSE)
```

This is one of the central lessons of the retina example:

- a graph scale that is good for age-aligned pseudotime is not automatically the
  graph scale preferred for every downstream outcome,
- `gflow` should therefore support explicit graph comparison rather than one
  fixed neighborhood convention.

## 7) Carry the Geometry into a Downstream Biological Summary

For the vignette, one downstream example is enough. The current retina work
already shows that the graph geometry can support interpretable module-level
summaries downstream of graph selection.

```{r fig-module-trajectories, out.width = "100%", fig.cap = "One downstream example: module trajectories remain robust across two smoothing scales, with the sharper graph preserving more local developmental detail."}
knitr::include_graphics(
  retina_asset_path("figures", "cluster_mean_trajectories_k10_vs_k20.png")
)
```

For a package vignette, the point is not to retell the entire Notch biology
story. The point is to show that once a geometry is selected, `gflow` can carry
that geometry into downstream conditional-expectation style summaries rather
than stopping at graph construction.

The larger retina analysis goes much further than this, but this one figure is
enough to show that the package workflow remains useful after graph selection.

## 8) Minimal Response-Refit Pattern

The response-refit step can reuse the spectral structure of an existing graph
fit. In a real single-cell analysis this is how we avoid rebuilding geometry for
every outcome.

```{r refit-pattern, eval = FALSE}
# Fit one seed response on the selected graph.
fit.seed <- fit.rdgraph.regression(
  X = X_view,
  y = y_seed,
  k = selected_k,
  adj.list = selected_graph$adj_list,
  weight.list = selected_graph$weight_list,
  pca.dim = NULL,
  verbose.level = 1L
)

# Refit many new responses using the cached spectral structure.
refit.out <- refit.rdgraph.regression(
  fitted.model = fit.seed,
  y.new = Y_multiresponse,
  per.column.gcv = TRUE,
  n.cores = 1L,
  verbose = FALSE
)
```

That pattern is important in the retina setting because the same selected
geometry can be reused for:

- pseudotime-style summaries,
- developmental age,
- subtype labels,
- cell-cycle proxies,
- targeted marker or module summaries.

## 9) What This Example Teaches About `gflow`

The retina example supports a clean package-level message:

1. `gflow` can be applied to single-cell data using the same core graph APIs.
2. Multi-view graph comparison is more informative than locking in one
   preprocessing choice.
3. Stability summaries are useful, but they should be combined with
   domain-relevant validation targets when those targets exist.
4. There is no reason to expect one globally optimal `k` across all downstream
   tasks.
5. The graph geometry is worth keeping because it can be reused for multiple
   downstream summaries rather than serving only as a preprocessing step.

## 10) Next Packaging Steps

To turn this draft into a fully portable package vignette, the next steps are:

1. expand the package-owned asset bundle from one vendored downstream figure to
   fully regenerated downstream figures where practical,
2. document the rebuild entry points more explicitly for users who want to
   refresh the vignette bundle from the source retina workspaces,
3. keep the full richer Notch/cell-cycle story in a paper or external report,
   while leaving this vignette focused on the reusable workflow.

That split keeps the vignette readable while still letting the retina example
demonstrate that `gflow` is useful on real single-cell developmental data.
