---
title: "Local Dimension Diagnostic Validation"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10.5,
  fig.height = 6.2,
  fig.align = "center",
  out.width = "100%",
  dpi = 140,
  dev = "ragg_png"
)
repo.dir <- "/Users/pgajer/current_projects/gflow"
validation.dir <- file.path(repo.dir, "dev/graph-trend-filtering/local-dimension-validation")
report.dir <- file.path(validation.dir, "reports")
source(file.path(validation.dir, "compute_local_dimension_validation.R"))
library(ggplot2)

full.path <- file.path(report.dir, "local_dimension_validation_full.rds")
smoke.path <- file.path(report.dir, "local_dimension_validation_smoke.rds")
result.path <- if (file.exists(full.path)) full.path else smoke.path
diagnostics <- readRDS(result.path)
local.table <- diagnostics$local.dimension
center.table <- diagnostics$center.selection
summary.table <- diagnostics$summary

base.theme <- function(base.size = 12) {
  theme_bw(base_size = base.size) +
    theme(
      plot.title = element_text(face = "bold", margin = margin(b = 8)),
      plot.subtitle = element_text(margin = margin(b = 8)),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 10)),
      strip.text = element_text(face = "bold", size = rel(0.86)),
      legend.position = "bottom",
      plot.margin = margin(10, 28, 10, 28)
    )
}

local.table$family.label <- gsub(" ", "\n", local.table$family)
center.table$family.label <- gsub(" ", "\n", center.table$family)
strict.centers <- subset(center.table, !transition)
ok.local <- subset(local.table, status == "ok")
ok.centers <- subset(center.table, !is.na(selected.local.dim))
```

## Purpose

This report validates the local dimension diagnostic used by the geodesic
pathwise annihilator experiments.  For each local graph disk
\(B_G(v,\varepsilon_v)\), the diagnostic embeds the induced weighted graph in
dimensions \(q=1,\ldots,q_{\max}\) and measures the relative edge-length stress

\[
E_q(v)
=
\left(
\frac{\sum_{(i,j)\in E_v}(\|z_i-z_j\|-\ell_{ij})^2}
     {\sum_{(i,j)\in E_v}\ell_{ij}^2}
\right)^{1/2}.
\]

The selected dimension is obtained from an edge-stress elbow rule.  This is not
a theorem about intrinsic dimension; it is an engineering diagnostic that should
behave correctly on controlled examples before it is used to guide row thinning
or local polynomial-basis construction.

The report also compares two additional estimators.  The metric-MDS spectrum
elbow estimates dimension from the positive eigenvalue spectrum of the local
graph-geodesic distance matrix.  The graph-Laplacian Weyl estimator fits a local
log-count/log-eigenvalue slope,

\[
N(\lambda)\approx C\lambda^{m/2},
\]

and reports \(m\approx 2\,d\log N/d\log\lambda\).  These estimators are not
expected to agree on every finite graph disk, but systematic disagreement is a
useful diagnostic of estimator calibration.

The validation ladder includes 1D path/cycle graphs, noisy circle graphs,
regular and warped grids, sampled planar graphs, quadric surface graphs, and
3D cube graphs.  The default validation uses `cmdscale`, while the optional
`hybrid.edge.kk` method uses weighted GRIP where appropriate and metric MDS
followed by edge-KK for dimensions above three when the installed `grip`
namespace provides those functions.  Noisy circles are treated as transition
cases: as radial noise increases, a growing fraction of dimension-2 selections
is reasonable.

## Stress Curves

The first check looks at the full stress profile

\[
E_1(v),\quad E_2(v),\quad E_3(v)
\]

for sampled local disks.  A 1D graph should be well represented in dimension
one.  A 2D graph should usually show a sharp drop from dimension one to two and
little additional gain afterward.  A 3D graph should often continue improving
through dimension three and then stabilize.

```{r stress-curves, fig.height=7.4}
ggplot(ok.local,
       aes(x = factor(embedding.dim), y = edge.rrmse,
           group = interaction(case.id, center), color = factor(true.dimension))) +
  geom_line(alpha = 0.28) +
  geom_point(alpha = 0.55, size = 1.3) +
  facet_wrap(~ family.label, scales = "free_y") +
  labs(
    title = "Local edge-length stress by embedding dimension",
    subtitle = "Each line is one sampled graph disk. The intended elbow is near the known intrinsic dimension.",
    x = "embedding dimension",
    y = "relative edge-length error",
    color = "known dimension"
  ) +
  base.theme()
```

The most important visual cue is the shape of the curve, not only the selected
dimension.  Strong 1-to-2 improvement in surface-like examples supports a local
2D interpretation.  Continued improvement into dimension three is expected for
volumetric examples and should be treated as a cap warning if the final stress
remains high.

## Selected Dimension

The next figure summarizes the selected local dimension directly.  Strict
accuracy is meaningful for the non-transition cases.  For noisy circles, the
selected dimension is interpreted as a noise-response curve rather than a hard
classification target.

```{r selected-dimension, fig.height=6.8}
ggplot(ok.centers,
       aes(x = factor(selected.local.dim), fill = factor(true.dimension))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ family.label, scales = "free_y") +
  labs(
    title = "Selected local dimension by validation family",
    subtitle = "Noisy circles are transition cases; exact labels are mainly meaningful for path, grid, quadric, and cube families.",
    x = "selected local dimension",
    y = "number of sampled graph disks",
    fill = "known dimension"
  ) +
  base.theme()
```

For clean path and cycle graphs, dimension one should dominate.  For grids,
sampled planar graphs, and quadric surfaces, dimension two should dominate,
especially away from boundaries.  Cube examples should produce many
dimension-three selections.

## Estimator Votes

The following plot compares the edge-stress elbow, metric-MDS spectrum elbow,
graph-Laplacian Weyl estimator, and simple consensus vote.  The aim is not to
declare one estimator universally correct; it is to see which controlled graph
families produce stable agreement and which ones expose calibration problems.

```{r estimator-votes, fig.height=7.0}
vote.long <- reshape(
  center.table[, c("case.id", "family", "true.dimension", "transition", "center",
                   "selected.local.dim", "mds.spectrum.local.dim",
                   "graph.spectrum.local.dim", "consensus.local.dim")],
  varying = c("selected.local.dim", "mds.spectrum.local.dim",
              "graph.spectrum.local.dim", "consensus.local.dim"),
  v.names = "estimated.dimension",
  timevar = "estimator",
  times = c("edge stress", "MDS spectrum", "Laplacian Weyl", "consensus"),
  direction = "long"
)
vote.long$family.label <- gsub(" ", "\n", vote.long$family)
ggplot(subset(vote.long, !is.na(estimated.dimension)),
       aes(x = factor(estimated.dimension), fill = estimator)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ family.label, scales = "free_y") +
  labs(
    title = "Dimension estimates from multiple local estimators",
    subtitle = "Agreement across estimators is stronger evidence than any single elbow rule.",
    x = "estimated dimension",
    y = "number of sampled graph disks",
    fill = "estimator"
  ) +
  base.theme()
```

The MDS spectrum estimator asks how many positive metric directions are needed
to explain the local graph-geodesic distance geometry.  The Laplacian estimator
asks how local diffusion eigenvalues scale.  The edge-stress estimator asks how
well embedded Euclidean edge lengths are realized.  These are related but not
identical notions of dimension, so disagreement should be inspected rather than
automatically averaged away.

## Confusion Matrix

For non-transition examples, the selected dimension can be compared directly
with the known dimension.

```{r confusion, fig.height=5.8}
confusion <- as.data.frame(with(
  subset(center.table, !transition & !is.na(selected.local.dim)),
  table(known = true.dimension, selected = selected.local.dim)
))
ggplot(confusion, aes(x = selected, y = known, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), fontface = "bold") +
  scale_fill_gradient(low = "#f7fbff", high = "#2166ac") +
  labs(
    title = "Selected dimension versus known dimension",
    subtitle = "Transition cases such as noisy circles are omitted from this strict classification view.",
    x = "selected dimension",
    y = "known dimension",
    fill = "count"
  ) +
  base.theme()
```

This plot is the strictest validation summary.  Off-diagonal mass identifies
where the stress-elbow rule is miscalibrated, where graph construction obscures
local geometry, or where the disk radius mixes boundary and interior behavior.

## Boundary Versus Interior

Boundary neighborhoods are typically less balanced than interior neighborhoods.
This figure separates the selected-dimension distribution by center region.

```{r boundary-interior, fig.height=6.6}
ggplot(subset(ok.centers, !transition),
       aes(x = factor(selected.local.dim), fill = center.region)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ family.label, scales = "free_y") +
  labs(
    title = "Selected dimension at boundary and interior centers",
    subtitle = "Boundary disks are expected to be less stable because local neighborhoods are one-sided.",
    x = "selected local dimension",
    y = "number of sampled graph disks",
    fill = "center region"
  ) +
  base.theme()
```

This check is important for deciding when mirror extensions or adaptive radius
rules should be used.  A method that works only on interiors but fails near
boundaries can still be useful, but the boundary behavior needs to be explicit.

## Noisy Circle Transition

The noisy circle benchmark is intentionally not a strict 1D/2D classification
problem.  Samples are generated near a circle, and a symmetric kNN graph is
constructed from the noisy coordinates.  As the radial noise level \(\sigma\)
increases, local disks can become less curve-like and more locally two
dimensional.

```{r noisy-circle-transition, fig.height=5.8}
noisy <- subset(center.table, transition & !is.na(selected.local.dim))
if (nrow(noisy) > 0L) {
  noisy.summary <- aggregate(
    selected.local.dim ~ noise.sigma,
    data = noisy,
    FUN = function(z) mean(z == 2L, na.rm = TRUE)
  )
  names(noisy.summary)[2L] <- "fraction.dim2"
  ggplot(noisy.summary, aes(x = noise.sigma, y = fraction.dim2)) +
    geom_line(linewidth = 0.9, color = "#2166ac") +
    geom_point(size = 2.3, color = "#2166ac") +
    scale_y_continuous(limits = c(0, 1)) +
    labs(
      title = "Noisy circle transition toward dimension two",
      subtitle = "A monotone or near-monotone increase is more meaningful than exact dimension-1 accuracy.",
      x = "radial noise scale",
      y = "fraction selected as dimension 2"
    ) +
    base.theme()
} else {
  plot.new()
  text(0.5, 0.5, "No noisy-circle transition cases were recorded.")
}
```

This transition plot tests calibration.  The diagnostic should not be forced to
call every noisy circle disk one-dimensional if the graph geometry itself has
become locally thickened.

## Stress Improvement Ratios

A compact way to inspect the elbow rule is to compare relative stress
improvements

\[
\Delta_q(v)=\frac{E_{q-1}(v)-E_q(v)}{E_{q-1}(v)}.
\]

Large \(\Delta_2\) and small \(\Delta_3\) is the expected surface signature.
Large \(\Delta_3\) is the expected volumetric signature.

```{r stress-improvements, fig.height=6.4}
wide <- reshape(
  ok.local[, c("case.id", "family", "true.dimension", "transition", "center",
               "embedding.dim", "edge.rrmse")],
  idvar = c("case.id", "family", "true.dimension", "transition", "center"),
  timevar = "embedding.dim",
  direction = "wide"
)
dim.cols <- sort(as.integer(sub("edge.rrmse.", "", grep("^edge\\.rrmse\\.", names(wide), value = TRUE))))
improve.parts <- lapply(dim.cols[-1L], function(q) {
  prev <- wide[[paste0("edge.rrmse.", q - 1L)]]
  curr <- wide[[paste0("edge.rrmse.", q)]]
  data.frame(
    wide[, c("case.id", "family", "true.dimension", "transition", "center")],
    transition.to = sprintf("%d -> %d", q - 1L, q),
    improvement = (prev - curr) / prev
  )
})
improve <- do.call(rbind, improve.parts)
improve$family.label <- gsub(" ", "\n", improve$family)
ggplot(improve, aes(x = transition.to, y = improvement,
                    fill = factor(true.dimension))) +
  geom_boxplot(outlier.alpha = 0.25) +
  facet_wrap(~ family.label, scales = "free_y") +
  labs(
    title = "Relative stress improvement by added embedding dimension",
    subtitle = "Surface-like disks should improve strongly from 1 to 2 and less from 2 to 3.",
    x = "dimension transition",
    y = "relative stress improvement",
    fill = "known dimension"
  ) +
  base.theme()
```

## Results Summary

```{r summary-text, results='asis'}
if (nrow(summary.table) > 0L) {
  strict <- subset(summary.table, !transition)
  if (nrow(strict) > 0L) {
    cat(sprintf(
      "Across non-transition validation families, the median strict accuracy is %.3f. ",
      stats::median(strict$strict.accuracy, na.rm = TRUE)
    ))
    cat(sprintf(
      "The median cap-hit rate is %.3f. ",
      stats::median(strict$cap.hit.rate, na.rm = TRUE)
    ))
  }
  if (nrow(noisy) > 0L) {
    cat("The noisy-circle cases should be read as a calibration curve: dimension-2 selections are allowed to increase with noise. ")
  }
}
```

The diagnostic is considered useful if it satisfies three conditions.  First,
the clean 1D, 2D, and 3D examples should produce the corresponding selected
dimension for most sampled disks.  Second, boundary errors should be visible
rather than hidden in a single global score.  Third, noisy-circle selections
should vary smoothly with noise instead of behaving erratically.

# Appendix

## Run Metadata

```{r appendix-metadata, echo=TRUE}
result.path
diagnostics$mode
diagnostics$created
diagnostics$control
```

## Summary Table

```{r appendix-summary, echo=TRUE}
summary.table
```

## Center Selection Table

```{r appendix-centers, echo=TRUE}
head(center.table, 80)
```

## Local Dimension Table

```{r appendix-local, echo=TRUE}
head(local.table, 120)
```

## Reproducibility Commands

```{r appendix-commands, echo=TRUE, eval=FALSE}
# Smoke run
GFLOW_LOCAL_DIM_VALIDATION_MODE=smoke \
  Rscript dev/graph-trend-filtering/local-dimension-validation/render_local_dimension_validation.R

# Full run
GFLOW_LOCAL_DIM_VALIDATION_MODE=full \
  Rscript dev/graph-trend-filtering/local-dimension-validation/render_local_dimension_validation.R
```
