\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts,mathtools}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations.pathreplacing,calc}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!55!black,
  urlcolor=blue!55!black,
  citecolor=blue!55!black
}

\newcommand{\R}{\mathbb R}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\reportbuilddatetime}{2026-05-16 10:10:45 EDT}

\title{Trend Filtering on Graphs: A Self-Contained Explainer for \texttt{gflow}}
\author{\texttt{gflow} graph trend filtering project notes}
\date{Build \reportbuilddatetime}

\begin{document}
\maketitle

\begin{abstract}
This note explains the main ideas behind one-dimensional trend filtering, graph
trend filtering, graph fused lasso, and related recent work in a way intended
to guide a future \texttt{gflow} implementation.  The emphasis is on accessible
mathematics: each formula is introduced by the modeling question it answers,
then interpreted in terms of what the estimator does to a signal.  The central
theme is the contrast between quadratic graph smoothing, which spreads
smoothness globally, and \(\ell_1\)-adaptive graph smoothing, which can preserve
localized changes by making graph differences sparse.
\end{abstract}

\section{Purpose and Scope}

The immediate motivation for this document is practical.  In the path-graph
benchmark for \texttt{fit.metric.graph.lowpass()}, the strongest classical
one-dimensional baseline was a wrapper around
\texttt{genlasso::trendfilter()}.  That result suggests that a graph analogue
of trend filtering may be an important comparator to add to \texttt{gflow}.

This document therefore has four goals:
\begin{enumerate}
  \item explain one-dimensional trend filtering from first principles;
  \item show how ordered finite differences become graph incidence and
        Laplacian operators;
  \item summarize the main published ideas on trend filtering and fused lasso
        over general graphs;
  \item identify the implementation decisions that matter for \texttt{gflow},
        especially for weighted metric graphs.
\end{enumerate}

The document is not a full literature review.  It is a design-oriented
mathematical explainer.

\section{Why Trend Filtering?}

Suppose we observe noisy values \(y_i\) of an unknown signal.  A smoother has to
decide how much of the variation in \(y_i\) is real signal and how much is
noise.  Standard quadratic smoothers penalize squared roughness.  This is
powerful, but it tends to smooth everywhere.  Trend filtering uses a different
principle: penalize the absolute values of discrete derivatives.

\[
  \text{quadratic smoothing:}\quad \text{many small penalties everywhere}
\]
\[
  \ell_1\text{ trend filtering:}\quad \text{many exact zeros, a few active bends}
\]

That exact-zero behavior is the lasso mechanism.  In ordinary lasso, an
\(\ell_1\) penalty sets many regression coefficients to zero.  In trend
filtering, an \(\ell_1\) penalty sets many discrete derivative values to zero.
The result is a curve that is simple over long stretches but can bend locally
where the data demand it.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,>=Latex]
  \draw[->] (0,0) -- (10.3,0) node[right] {\(x\)};
  \draw[->] (0,0) -- (0,4.0) node[above] {\(y\)};
  \foreach \x/\y in {0.4/0.45,0.9/0.8,1.3/1.25,1.9/2.0,2.5/1.8,3.0/1.25,
                    3.6/0.9,4.2/0.75,4.8/1.05,5.4/1.65,6.0/2.55,6.7/2.95,
                    7.3/2.55,7.9/1.85,8.5/1.15,9.1/0.75,9.7/0.5}
    \fill[black] (\x,\y) circle (1.35pt);

  \draw[orange!80!black,very thick,smooth]
    plot coordinates {(0.3,0.55) (1.0,0.9) (1.7,1.35) (2.4,1.48)
                      (3.1,1.32) (3.8,1.25) (4.5,1.32) (5.2,1.65)
                      (5.9,2.05) (6.6,2.25) (7.3,2.15) (8.0,1.75)
                      (8.7,1.18) (9.6,0.63)};

  \draw[blue,very thick]
    plot[smooth] coordinates {(0.3,0.42) (0.9,0.80) (1.5,1.45) (2.2,1.95) (2.8,1.62)};
  \draw[blue,very thick]
    plot[smooth] coordinates {(2.8,1.62) (3.5,0.98) (4.2,0.78) (4.9,1.12)};
  \draw[blue,very thick]
    plot[smooth] coordinates {(4.9,1.12) (5.5,1.85) (6.2,2.78) (6.9,2.85)};
  \draw[blue,very thick]
    plot[smooth] coordinates {(6.9,2.85) (7.6,2.20) (8.4,1.22) (9.6,0.48)};

  \foreach \x in {2.8,4.9,6.9}
    \draw[red,dashed,thick] (\x,0) -- (\x,3.35);

  \node[orange!80!black,anchor=west] at (0.35,3.75) {quadratic smoother};
  \node[orange!80!black,anchor=west] at (0.35,3.45) {bends broadly};
  \node[blue,anchor=west] at (6.05,3.75) {trend filtering};
  \node[blue,anchor=west] at (6.05,3.45) {sparse bends};
  \node[red] at (7.70,3.08) {selected bend locations};
\end{tikzpicture}
\caption{A conceptual comparison.  A quadratic smoother spreads curvature
continuously; trend filtering keeps many higher-order differences exactly zero
and concentrates bends at selected locations.}
\end{figure}

\section{One-Dimensional Trend Filtering}

\subsection{Data and Fitted Values}

Start with ordered observations
\[
  x_1 < x_2 < \cdots < x_n,
  \qquad
  y_i = f(x_i) + \varepsilon_i.
\]
Trend filtering estimates fitted values
\[
  \theta = (\theta_1,\ldots,\theta_n)^T,
  \qquad
  \theta_i \approx f(x_i).
\]
For now, assume the \(x_i\) are equally spaced.  Unequal spacing can be handled
by divided-difference versions of the same operators.

\subsection{Finite Differences}

The first difference measures local change:
\[
  (D^{(1)}\theta)_i = \theta_{i+1}-\theta_i.
\]
The second difference measures local curvature:
\[
  (D^{(2)}\theta)_i
  =
  \theta_{i+2}-2\theta_{i+1}+\theta_i.
\]
The third difference measures change in curvature:
\[
  (D^{(3)}\theta)_i
  =
  \theta_{i+3}-3\theta_{i+2}+3\theta_{i+1}-\theta_i.
\]

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1.25cm,y=1cm,>=Latex]
  \foreach \i/\x/\lab in {1/0/\theta_i,2/1/\theta_{i+1},3/2/\theta_{i+2},4/3/\theta_{i+3}} {
    \fill[black] (\x,0) circle (2pt);
    \node[below] at (\x,-0.05) {\(\lab\)};
  }
  \draw[blue,->,thick] (0,0.35) -- (1,0.35) node[midway,above] {\(\theta_{i+1}-\theta_i\)};
  \draw[orange!80!black,decorate,decoration={brace,amplitude=5pt}] (0,0.95) -- (2,0.95)
    node[midway,above=6pt] {second difference};
  \draw[red,decorate,decoration={brace,amplitude=5pt}] (0,1.65) -- (3,1.65)
    node[midway,above=6pt] {third difference};
\end{tikzpicture}
\caption{Trend filtering penalizes finite differences.  Higher order
differences measure higher order changes in the fitted curve.}
\end{figure}

\subsection{The Optimization Problem}

Order-\(k\) trend filtering solves
\[
  \widehat{\theta}_{\lambda}
  =
  \argmin_{\theta\in\R^n}
  \left\{
  \frac{1}{2}\norm{y-\theta}_2^2
  +
  \lambda\norm{D^{(k+1)}\theta}_1
  \right\}.
\]
The first term is data fidelity.  The second term is an \(\ell_1\) penalty on
the \((k+1)\)st discrete derivative.

\begin{center}
\begin{tabular}{lll}
\toprule
Order \(k\) & Penalized quantity & Shape of fitted signal\\
\midrule
\(0\) & first differences & piecewise constant\\
\(1\) & second differences & piecewise linear\\
\(2\) & third differences & piecewise quadratic\\
\bottomrule
\end{tabular}
\end{center}

The path-graph benchmark used \texttt{genlasso::trendfilter(y, pos = x, ord =
2)}, so it used \(k=2\).  That means it estimated a locally piecewise quadratic
curve and selected the amount of smoothing by cross-validation.

\section{From a Line to a Graph}

The path graph is the bridge between one-dimensional trend filtering and graph
trend filtering.  On a line, each point has a left and right neighbor.  On a
graph, each vertex has graph neighbors.  The replacement is:
\[
  \text{ordered finite differences}
  \quad\longrightarrow\quad
  \text{edge differences}.
\]

Choose an arbitrary orientation for every undirected edge.  If edge \(e\) is
oriented from vertex \(i\) to vertex \(j\), define one row of the incidence
matrix \(D\) by
\[
  D_{e,i}=-1,\qquad D_{e,j}=1,\qquad D_{e,\ell}=0
  \quad(\ell\notin\{i,j\}).
\]
Then
\[
  (D\beta)_e = \beta_j-\beta_i.
\]

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1.2cm,y=1cm,>=Latex]
  \node[circle,draw,minimum size=7mm] (v1) at (0,0) {\(1\)};
  \node[circle,draw,minimum size=7mm] (v2) at (1.8,0.65) {\(2\)};
  \node[circle,draw,minimum size=7mm] (v3) at (3.6,0) {\(3\)};
  \node[circle,draw,minimum size=7mm] (v4) at (1.8,-1.0) {\(4\)};
  \draw[->,thick] (v1) -- node[above] {\(e_1\)} (v2);
  \draw[->,thick] (v2) -- node[above] {\(e_2\)} (v3);
  \draw[->,thick] (v1) -- node[left] {\(e_3\)} (v4);
  \draw[->,thick] (v4) -- node[right] {\(e_4\)} (v3);
  \draw[->,thick] (v2) -- node[right] {\(e_5\)} (v4);

  \node[anchor=west,align=left] at (5.1,0.2) {
    \(D\beta\) has one entry per edge.\\[1mm]
    If \(e_1:1\to2\), then\\
    \((D\beta)_{e_1}=\beta_2-\beta_1\).\\[1mm]
    The chosen orientation only sets signs;\\
    \(\ell_1\) penalties use absolute values.
  };
\end{tikzpicture}
\caption{The incidence matrix turns a vertex signal into edge differences.}
\end{figure}

The graph Laplacian is
\[
  L = D^T D.
\]
For an unweighted graph, \(L\beta\) compares each vertex value to the values of
its neighbors:
\[
  (L\beta)_i
  =
  \sum_{j\sim i}(\beta_i-\beta_j).
\]

\section{Graph Trend Filtering}

Wang, Sharpnack, Smola, and Tibshirani define graph trend filtering by
constructing higher-order graph difference operators from \(D\) and \(L\).  The
estimator is
\[
  \widehat\beta_\lambda
  =
  \argmin_{\beta\in\R^n}
  \left\{
  \frac{1}{2}\norm{y-\beta}_2^2
  +
  \lambda\norm{\Delta^{(k+1)}\beta}_1
  \right\}.
\]

One convenient expression for their graph difference operators is
\[
  \Delta^{(k+1)}
  =
  \begin{cases}
  L^{(k+1)/2}, & k \text{ odd},\\[1mm]
  D L^{k/2}, & k \text{ even}.
  \end{cases}
\]

\begin{center}
\begin{tabular}{llll}
\toprule
Order \(k\) & Operator & Penalized object & Graph interpretation\\
\midrule
\(0\) & \(D\) & edge differences & piecewise constant regions\\
\(1\) & \(L\) & vertex Laplacian values & graph piecewise linear analogue\\
\(2\) & \(DL\) & edge differences of Laplacian values & graph piecewise quadratic analogue\\
\bottomrule
\end{tabular}
\end{center}

\subsection{\(k=0\): Graph Fused Lasso}

For \(k=0\), \(\Delta^{(1)}=D\), so
\[
  \norm{\Delta^{(1)}\beta}_1
  =
  \sum_{(i,j)\in E}|\beta_i-\beta_j|.
\]
The estimator is
\[
  \widehat\beta
  =
  \argmin_{\beta\in\R^n}
  \left\{
  \frac{1}{2}\norm{y-\beta}_2^2
  +
  \lambda\sum_{(i,j)\in E}|\beta_i-\beta_j|
  \right\}.
\]
This is graph fused lasso or graph total variation denoising.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[scale=0.95,>=Latex]
  \tikzstyle{vtx}=[circle,draw,minimum size=7mm,inner sep=0pt]
  \foreach \name/\x/\y/\lab in {a/0/1.1/1,b/1.3/1.9/2,c/2.7/1.2/3,d/1.2/0/4,
                               e/4.1/1.9/5,f/4.1/0.2/6,g/5.5/1.1/7} {
    \node[vtx] (\name) at (\x,\y) {\(\lab\)};
  }
  \draw[thick] (a)--(b)--(c)--(d)--(a);
  \draw[thick] (b)--(d);
  \draw[thick] (c)--(e)--(g)--(f)--(c);
  \draw[thick] (e)--(f);
  \draw[red,very thick] (c)--(e);
  \node[align=center,blue!70!black] at (1.25,2.85) {region with\\similar fitted values};
  \node[align=center,blue!70!black] at (4.85,2.85) {another\\region};
  \node[red,align=left,anchor=west,fill=white,inner sep=1pt] (jump) at (6.05,1.55)
    {active\\jump};
  \draw[red,->,thick] (jump.west) -- ($(c)!0.5!(e)$);
\end{tikzpicture}
\caption{The \(k=0\) penalty encourages piecewise constant graph regions while
allowing selected edges to carry jumps.}
\end{figure}

\subsection{\(k=1\): Penalizing \(L\beta\)}

For \(k=1\), the operator is
\[
  \Delta^{(2)} = L.
\]
The penalty is
\[
  \norm{L\beta}_1
  =
  \sum_i \left|\sum_{j\sim i}(\beta_i-\beta_j)\right|.
\]
On a path graph, \(L\beta\) is essentially a second difference.  On a general
graph, it measures whether a vertex value agrees with a local average of its
neighbors.  Sparsity in \(L\beta\) means many vertices behave locally like
graph-harmonic points.

\subsection{\(k=2\): Penalizing \(DL\beta\)}

For \(k=2\), the operator is
\[
  \Delta^{(3)} = DL.
\]
This takes the graph Laplacian values \(L\beta\) at vertices and then computes
edge differences between those values.  On a path graph this corresponds to a
third-difference penalty, which is why the analogy is piecewise quadratic.

\section{Weighted Graphs and Metric Conductances}

The published graph trend filtering construction is often introduced using an
unweighted incidence matrix.  In \texttt{gflow}, graphs usually carry metric
edge lengths or conductances, so weighting is not a detail.  It is a modeling
choice.

For edge \(e=(i,j)\), a weighted incidence operator can be written as
\[
  (D_w\beta)_e
  =
  \omega_e(\beta_j-\beta_i).
\]
Then the graph fused lasso penalty becomes
\[
  \norm{D_w\beta}_1
  =
  \sum_{e=(i,j)}\omega_e|\beta_j-\beta_i|.
\]

If \(c_e\) is a conductance, two natural conventions are:
\[
  \omega_e = c_e
  \qquad\text{or}\qquad
  \omega_e = \sqrt{c_e}.
\]
The first convention makes the \(\ell_1\) edge penalty scale directly with
conductance:
\[
  \sum_e c_e|\beta_j-\beta_i|.
\]
The second convention aligns with the quadratic Laplacian identity:
\[
  \norm{D_w\beta}_2^2 = \beta^T L_w\beta,
  \qquad
  L_w = D_w^T D_w.
\]

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,>=Latex]
  \node[circle,draw,minimum size=8mm] (i) at (0,0) {\(i\)};
  \node[circle,draw,minimum size=8mm] (j) at (4.2,0) {\(j\)};
  \draw[very thick,->] (i) -- node[above] {\(e\)} node[below] {\(\ell_e,\;c_e=\phi(\ell_e)\)} (j);
  \node[align=left,anchor=west] at (5.0,0.65) {
    metric edge length: \(\ell_e\)\\
    conductance transform: \(c_e=\phi(\ell_e)\)\\
    graph trend weight: \(\omega_e=c_e\) or \(\sqrt{c_e}\)
  };
  \node[align=left,anchor=west] at (5.0,-0.85) {
    penalty contribution:\\
    \(\omega_e|\beta_j-\beta_i|\)
  };
\end{tikzpicture}
\caption{For metric graph trend filtering, \texttt{gflow} must decide how
metric lengths become \(\ell_1\) edge-difference weights.}
\end{figure}

This choice should probably be explicit in the API.  A first implementation
could expose a weighting option such as
\[
  \omega_e \in \{1,\;c_e,\;\sqrt{c_e}\}.
\]

\section{Relationship to Metric Graph Low-Pass Smoothing}

The metric graph low-pass smoother is fundamentally \(\ell_2\)-spectral.  A
closely related variational form is
\[
  \min_f
  \left\{
  \sum_i(y_i-f_i)^2+\eta f^T L f
  \right\}.
\]
The graph trend filtering analogue is
\[
  \min_\beta
  \left\{
  \sum_i(y_i-\beta_i)^2+\lambda\norm{\Delta^{(k+1)}\beta}_1
  \right\}.
\]

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,>=Latex]
  \node[draw,rounded corners,align=center,minimum width=4.0cm,minimum height=1.4cm] (l2) at (0,0)
    {\(\ell_2\) Laplacian smoothing\\[1mm]\(\eta f^TLf\)};
  \node[draw,rounded corners,align=center,minimum width=4.0cm,minimum height=1.4cm] (l1) at (7.0,0)
    {\(\ell_1\) graph trend filtering\\[1mm]\(\lambda\norm{\Delta^{(k+1)}\beta}_1\)};
  \node[align=center] at (0,-1.35) {shrinks high-frequency\\variation broadly};
  \node[align=center] at (7.0,-1.35) {sets many graph differences\\exactly to zero};
  \draw[->,thick] (l2) -- node[above,align=center] {comparator,\\not replacement} (l1);
\end{tikzpicture}
\caption{Graph trend filtering is the \(\ell_1\)-adaptive counterpart to
\(\ell_2\) graph low-pass smoothing.}
\end{figure}

\section{Published Ideas to Build On}

\subsection{Trend Filtering on Graphs}

Yu-Xiang Wang, James Sharpnack, Alex Smola, and Ryan J. Tibshirani introduced
the graph trend filtering estimator and studied its statistical behavior.

\begin{itemize}
  \item arXiv: \url{https://arxiv.org/abs/1410.7690}
  \item JMLR PDF: \url{https://jmlr.csail.mit.edu/papers/volume17/15-147/15-147.pdf}
  \item Main contribution: graph difference operators \(\Delta^{(k+1)}\) and
        locally adaptive graph denoising.
\end{itemize}

\subsection{The DFS Fused Lasso}

Madrid Padilla, Sharpnack, Scott, and Tibshirani focused on scalable graph
fused lasso, corresponding mainly to the \(k=0\) case.  The idea is to use a
depth-first-search traversal to reduce a graph denoising problem to a chain-like
problem while retaining useful guarantees.

\begin{itemize}
  \item JMLR page: \url{https://www.jmlr.org/beta/papers/v18/16-532.html}
  \item Main contribution: fast graph total variation denoising over general
        graphs.
\end{itemize}

\subsection{Fused \(\ell_1\) Trend Filtering on Graphs}

Vladimir Pastukhov's recent arXiv work studies fused \(\ell_1\) trend
filtering on graphs, including combinations of trend filtering and fusion
regularizers.

\begin{itemize}
  \item arXiv: \url{https://arxiv.org/abs/2401.05250}
  \item Main contribution: broader fused graph trend filtering objectives and
        scalable iterative schemes.
\end{itemize}

\section{Implementation Implications for \texttt{gflow}}

A conservative implementation path is:
\begin{enumerate}
  \item Start with \(k=0\), the weighted graph fused lasso:
  \[
    \widehat\beta
    =
    \argmin_{\beta\in\R^n}
    \left\{
    \frac{1}{2}\norm{y-\beta}_2^2
    +
    \lambda\norm{D_w\beta}_1
    \right\}.
  \]
  \item Add an operator-construction helper that returns the edge table,
        orientation, weights, incidence triplets, and diagnostics.
  \item Make the metric length to trend weight convention explicit.
  \item Add path-graph checks against one-dimensional fused-lasso or trend
        filtering references.
  \item Add higher-order \(k=1\) and \(k=2\) operators once the solver and
        weighting conventions are stable.
\end{enumerate}

\section{Open Design Questions}

\begin{enumerate}
  \item Should edge weights in \(D_w\) be \(c_e\), \(\sqrt{c_e}\), or another
        transform of metric length?
  \item Should the first solver be exact for small/medium graphs, approximate
        for large graphs, or both?
  \item Should disconnected graphs be fit componentwise?
  \item Should the API expose both fixed-\(\lambda\) refits and cross-validated
        \(\lambda\) selection?
  \item Should DFS fused lasso be an approximation mode or a separate
        estimator?
\end{enumerate}

\section{Summary}

One-dimensional trend filtering works by penalizing discrete derivatives with
an \(\ell_1\) norm.  Graph trend filtering replaces ordered finite differences
with graph incidence and Laplacian-based difference operators.  The resulting
estimators are locally adaptive graph smoothers: they can be simple over large
regions while preserving localized changes.  That makes them natural
comparators to \texttt{gflow}'s metric graph low-pass smoother and a promising
next project for graph-based regression experiments.

\end{document}
