---
title: "Graph Trend Filtering on Curated Gaussian-Mixture Subset"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 11,
  fig.height = 7,
  dpi = 120
)

repo.dir <- "/Users/pgajer/current_projects/gflow"
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(repo.dir, quiet = TRUE)
} else {
  library(gflow)
}

library(gflow)
library(ggplot2)
source(file.path(repo.dir, "dev/metric-graph-lowpass-benchmarks/multi_scenario_1d_mixture_helpers.R"))
```

## Purpose

This report applies the phase-2 `fit.graph.trend.filtering()` implementation to
a curated subset of the 240-dataset one-dimensional Gaussian-mixture benchmark.
The goal is visual and diagnostic: do the new graph trend-filtering comparators
capture the true Gaussian-mixture signal, and how do they compare with the best
method from the earlier path-graph benchmark, `genlasso::trendfilter(..., ord = 2)`?

The comparison is intentionally not a final verdict on graph trend filtering.
The current `gflow` function implements

\[
\widehat\beta_\lambda =
\arg\min_{\beta}
\left\{
\frac{1}{2}\|y-\beta\|_2^2
+ \lambda\|\Delta_w^{(k+1)}\beta\|_1
\right\},
\]

with

\[
k=0:\Delta_w^{(1)}=D_w,\qquad
k=1:\Delta_w^{(2)}=L_w,\qquad
k=2:\Delta_w^{(3)}=D_wL_w.
\]

For \(k=0\), this is graph fused lasso. For \(k=1,2\), this follows the graph
trend-filtering operator construction from Wang, Sharpnack, Smola, and
Tibshirani. The benchmark winner, `genlasso::trendfilter(..., ord = 2)`, is a
specialized one-dimensional divided-difference trend filter, so the comparison
is still a diagnostic rather than exact path-graph parity.

## Benchmark Design

```{r load}
results.file <- file.path(
  repo.dir,
  "dev/metric-graph-lowpass-benchmarks/reports/graph_trend_filtering_subset_results.rds"
)
if (!file.exists(results.file)) {
  stop("Missing precomputed results file. Run compute_graph_trend_filtering_subset.R first.")
}

bench <- readRDS(results.file)
results <- bench$results
predictions <- bench$predictions

data.frame(
  n.datasets = length(unique(results$dataset.id)),
  n.methods = length(unique(results$method)),
  n.result.rows = nrow(results),
  n.prediction.rows = nrow(predictions),
  n.errors = sum(results$status != "ok"),
  results.file = results.file
)
```

The subset uses eight shape families, all three sampling/noise variants, and
one replicate per shape/variant pair. Every fit uses the sorted one-dimensional
path graph. Adjacent samples in sorted \(x\) are connected.

The graph trend-filtering variants are:

- `fit.graph.trend.filtering.order0.unit`: \(k=0\), unweighted path graph.
- `fit.graph.trend.filtering.order0.conductance`: \(k=0\), metric edge lengths are first
  transformed into conductances with `exp.length, sigma=edge.q0.75`, then
  floored at \(10^{-4}\), median-normalized, and used as \(\omega_e=c_e\).
- `fit.graph.trend.filtering.order0.sqrt.conductance`: \(k=0\), the same
  conductances are used, but \(\omega_e=\sqrt{c_e}\).
- `fit.graph.trend.filtering.order1.unit`: \(k=1\), unweighted \(L_w\) penalty.
- `fit.graph.trend.filtering.order2.unit`: \(k=2\), unweighted \(D_wL_w\) penalty.
- `fit.graph.trend.filtering.order2.sqrt.conductance`: \(k=2\), square-root
  metric-conductance weights.

The higher-order graph fits use `approx = TRUE` in the `genlasso` backend for
this report. This is a pragmatic smoke-benchmark choice: exact dense/SVD
`order = 1` paths are much slower at \(n=250\).

## Overall Scores

```{r summaries}
ok <- subset(results, status == "ok")
ok$method <- factor(ok$method, levels = c(
  "genlasso.trendfilter.ord2.cv",
  "fit.graph.trend.filtering.order0.unit",
  "fit.graph.trend.filtering.order0.conductance",
  "fit.graph.trend.filtering.order0.sqrt.conductance",
  "fit.graph.trend.filtering.order1.unit",
  "fit.graph.trend.filtering.order2.unit",
  "fit.graph.trend.filtering.order2.sqrt.conductance"
))

summary.table <- aggregate(
  rmse.truth ~ method,
  data = ok,
  FUN = function(z) c(
    median = median(z),
    mean = mean(z),
    q25 = unname(quantile(z, 0.25)),
    q75 = unname(quantile(z, 0.75)),
    q90 = unname(quantile(z, 0.90))
  )
)
summary.table <- do.call(data.frame, summary.table)
names(summary.table) <- sub("rmse.truth.", "", names(summary.table), fixed = TRUE)
summary.table[order(summary.table$median), ]
```

```{r rmse-violin}
ggplot(ok, aes(x = method, y = rmse.truth, fill = method)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.65, color = "gray35") +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2.3,
               fill = "white", color = "black") +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "RMSE to truth on curated Gaussian-mixture subset",
    subtitle = "White diamonds mark medians; y-axis is log scaled",
    x = NULL,
    y = "RMSE to truth, log scale"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r paired-ratios}
paired_ratio <- function(dat, reference) {
  ref <- dat[dat$method == reference, c("dataset.id", "rmse.truth")]
  names(ref)[2] <- "ref.rmse"
  out <- lapply(setdiff(unique(as.character(dat$method)), reference), function(method) {
    cur <- dat[dat$method == method, c("dataset.id", "rmse.truth")]
    names(cur)[2] <- "cur.rmse"
    m <- merge(ref, cur, by = "dataset.id")
    d <- log(m$cur.rmse) - log(m$ref.rmse)
    tt <- stats::t.test(d)
    data.frame(
      method = method,
      reference = reference,
      n = length(d),
      mean.rmse.ratio = exp(mean(d)),
      ratio.conf.low = exp(unname(tt$conf.int[1])),
      ratio.conf.high = exp(unname(tt$conf.int[2])),
      median.rmse.ratio = median(m$cur.rmse / m$ref.rmse),
      p.value = tt$p.value,
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, out)
}

ratio.table <- paired_ratio(ok, "genlasso.trendfilter.ord2.cv")
ratio.table[order(ratio.table$mean.rmse.ratio), ]
```

```{r ratio-forest}
ratio.plot <- ratio.table
ggplot(ratio.plot, aes(x = mean.rmse.ratio, y = reorder(method, mean.rmse.ratio))) +
  geom_vline(xintercept = 1, linetype = 2, color = "gray45") +
  geom_errorbarh(aes(xmin = ratio.conf.low, xmax = ratio.conf.high),
                 height = 0.18, linewidth = 0.7, color = "#2c7fb8") +
  geom_point(size = 2.5, color = "#2c7fb8") +
  scale_x_log10() +
  labs(
    title = "Paired RMSE ratios versus quadratic trend filtering",
    subtitle = "Values above 1 mean higher RMSE than the ord=2 trend filter",
    x = "RMSE ratio versus genlasso ord=2, log scale",
    y = NULL
  ) +
  theme_minimal(base_size = 11)
```

## Example Fits

```{r prepare-predictions}
predictions$method <- factor(as.character(predictions$method), levels = levels(ok$method))

dataset.order <- unique(results$dataset.id)
predictions$dataset.label <- paste0(
  predictions$dataset.id, "\n",
  predictions$shape.name, "\n",
  predictions$variant.name
)
label.levels <- unique(predictions$dataset.label[match(dataset.order, predictions$dataset.id)])
predictions$dataset.label <- factor(predictions$dataset.label, levels = label.levels)

data.points <- unique(predictions[, c("dataset.id", "dataset.label", "x", "y", "truth")])
truth.lines <- data.points[order(data.points$dataset.id, data.points$x), ]

method.colors <- c(
  "genlasso.trendfilter.ord2.cv" = "#1b9e77",
  "fit.graph.trend.filtering.order0.unit" = "#2b6cb0",
  "fit.graph.trend.filtering.order0.conductance" = "#d95f02",
  "fit.graph.trend.filtering.order0.sqrt.conductance" = "#7570b3",
  "fit.graph.trend.filtering.order1.unit" = "#66a61e",
  "fit.graph.trend.filtering.order2.unit" = "#a6761d",
  "fit.graph.trend.filtering.order2.sqrt.conductance" = "#1f78b4"
)
```

```{r all-panels, fig.height=13}
ggplot() +
  geom_point(data = data.points, aes(x = x, y = y),
             color = "black", alpha = 0.38, size = 0.45) +
  geom_line(data = truth.lines, aes(x = x, y = truth),
            color = "#d73027", linewidth = 0.75) +
  geom_line(data = predictions[order(predictions$dataset.id, predictions$method, predictions$x), ],
            aes(x = x, y = fitted, color = method),
            linewidth = 0.55, alpha = 0.90) +
  facet_wrap(~ dataset.label, ncol = 3) +
  scale_color_manual(values = method.colors) +
  labs(
    title = "Graph trend filtering versus quadratic 1D trend filtering",
    subtitle = "Red: true mixture; black: noisy observations; colored lines: fitted methods",
    x = "x",
    y = "response",
    color = "Fit"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7.5)
  )
```

## Where Does Graph Trend Filtering Look Best?

```{r best-graph}
graph.methods <- grep("^fit\\.graph\\.trend\\.filtering", as.character(ok$method), value = TRUE)
best.graph <- ok[ok$method %in% graph.methods, ]
best.graph <- best.graph[order(best.graph$dataset.id, best.graph$rmse.truth), ]
best.graph <- best.graph[!duplicated(best.graph$dataset.id), ]

ord2 <- ok[ok$method == "genlasso.trendfilter.ord2.cv", c("dataset.id", "rmse.truth")]
names(ord2)[2] <- "ord2.rmse"
best.graph <- merge(best.graph, ord2, by = "dataset.id")
best.graph$ratio.to.ord2 <- best.graph$rmse.truth / best.graph$ord2.rmse
best.graph[order(best.graph$ratio.to.ord2),
           c("dataset.id", "shape.name", "variant.name", "method",
             "rmse.truth", "ord2.rmse", "ratio.to.ord2", "lambda")]
```

```{r best-graph-panels, fig.height=9}
top.graph.datasets <- head(best.graph$dataset.id[order(best.graph$ratio.to.ord2)], 9)
top.pred <- subset(predictions, dataset.id %in% top.graph.datasets)
top.points <- subset(data.points, dataset.id %in% top.graph.datasets)
top.truth <- subset(truth.lines, dataset.id %in% top.graph.datasets)

ggplot() +
  geom_point(data = top.points, aes(x = x, y = y),
             color = "black", alpha = 0.42, size = 0.55) +
  geom_line(data = top.truth, aes(x = x, y = truth),
            color = "#d73027", linewidth = 0.8) +
  geom_line(data = top.pred[order(top.pred$dataset.id, top.pred$method, top.pred$x), ],
            aes(x = x, y = fitted, color = method),
            linewidth = 0.65, alpha = 0.92) +
  facet_wrap(~ dataset.label, ncol = 3) +
  scale_color_manual(values = method.colors) +
  labs(
    title = "Datasets where the best graph trend-filtering variant is closest to ord=2 trend filtering",
    x = "x",
    y = "response",
    color = "Fit"
  ) +
  theme_minimal(base_size = 9) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 7.5)
  )
```

## Interpretation

The \(k=0\) graph trend-filtering fits should be read as graph fused-lasso
baselines: they can create useful plateaus, but smooth Gaussian mixtures often
need higher-order behavior. The \(k=1\) and \(k=2\) fits are the first direct
graph higher-order comparators in `gflow`. If they close the gap with
`genlasso::trendfilter(..., ord = 2)`, that supports continuing with solver and
weighting refinements. If they remain worse but visually sensible, the likely
next issue is not the operator definition alone but the solver, path-graph
boundary behavior, and metric weighting convention.
