---
title: "Metric Graph Low-Pass: Curated 1D Gaussian-Mixture Benchmark"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 120
)

repo.dir <- "/Users/pgajer/current_projects/gflow"
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(repo.dir, quiet = TRUE)
} else {
  library(gflow)
}

library(gflow)
library(ggplot2)
source(file.path(repo.dir, "dev/metric-graph-lowpass-benchmarks/multi_scenario_1d_mixture_helpers.R"))
```

## Purpose

This benchmark expands the single-scenario smoke test to a curated set of
1D Gaussian-mixture regression problems over \([0,1]\). The goal is to compare
direct metric-conductance low-pass smoothing against the existing
`fit.rdgraph.regression()` baseline over many visibly distinct mixture shapes
and controlled sampling/noise patterns.

The benchmark uses:

- 16 curated mixture shapes: 10 two-Gaussian and 6 three-Gaussian signals.
- 3 sampling/noise variants per shape.
- 5 random replicates per shape/variant pair.
- \(16 \times 3 \times 5 = 240\) datasets, each with \(n = 250\).
- `laplacian.type = "unnormalized"` only for `fit.metric.graph.lowpass()`.

For fairness, both the metric smoother and rdgraph baseline use the same
MST-repaired symmetric kNN graph for each dataset:

```r
g <- create.sknn.graph(X, k = graph.k, connect.components = TRUE)
```

The metric smoother receives `g$adj_list` and `g$weight_list` as metric edge
lengths. The rdgraph baseline receives the same graph as a precomputed graph.

## Conductance Rules

For metric-lowpass fits, `weight.list` values are metric lengths
\(\ell_{ij}\). The benchmark uses these conductance transforms:

\[
c_{ij} = (\ell_{ij} + \epsilon)^{-\alpha}
\]

\[
c_{ij} = \exp(-\ell_{ij}/\sigma)
\]

\[
c_{ij} = \exp(-\ell_{ij}^{2}/\sigma^{2})
\]

\[
c_{ij} = \exp\{-\ell_{ij}^{2}/(\sigma_i\sigma_j)\}.
\]

## Helpers

```{r helpers, eval=FALSE, include=FALSE}
normalize_to_unit_max <- function(y) {
  if (max(y) <= 0) y else y / max(y)
}

mixture_signal <- function(x, means, sds, weights) {
  y <- numeric(length(x))
  for (j in seq_along(means)) {
    y <- y + weights[j] * dnorm(x, mean = means[j], sd = sds[j])
  }
  normalize_to_unit_max(y)
}

shape_catalog <- function() {
  data.frame(
    shape.id = sprintf("S%02d", 1:16),
    shape.name = c(
      "two separated narrow equal",
      "two separated unequal",
      "two close overlapping equal",
      "two close overlapping unequal",
      "narrow left plus broad right",
      "broad left plus narrow right",
      "left boundary plus interior",
      "interior plus right boundary",
      "deep valley separated",
      "nearly unimodal overlap",
      "three separated peaks",
      "dominant center side bumps",
      "two close plus distant",
      "narrow center spike",
      "left-heavy triple",
      "boundary interior boundary"
    ),
    means = I(list(
      c(0.28, 0.72),
      c(0.25, 0.74),
      c(0.43, 0.56),
      c(0.40, 0.54),
      c(0.30, 0.69),
      c(0.31, 0.70),
      c(0.08, 0.48),
      c(0.52, 0.93),
      c(0.22, 0.78),
      c(0.45, 0.56),
      c(0.20, 0.50, 0.80),
      c(0.23, 0.52, 0.79),
      c(0.35, 0.46, 0.82),
      c(0.24, 0.50, 0.76),
      c(0.16, 0.38, 0.70),
      c(0.08, 0.50, 0.92)
    )),
    sds = I(list(
      c(0.045, 0.045),
      c(0.050, 0.070),
      c(0.080, 0.080),
      c(0.060, 0.095),
      c(0.035, 0.140),
      c(0.140, 0.035),
      c(0.035, 0.090),
      c(0.090, 0.035),
      c(0.040, 0.040),
      c(0.115, 0.115),
      c(0.040, 0.050, 0.040),
      c(0.060, 0.045, 0.060),
      c(0.045, 0.045, 0.060),
      c(0.090, 0.025, 0.090),
      c(0.040, 0.060, 0.080),
      c(0.035, 0.075, 0.035)
    )),
    weights = I(list(
      c(1.00, 1.00),
      c(1.00, 0.55),
      c(1.00, 1.00),
      c(1.00, 0.45),
      c(1.00, 0.85),
      c(0.85, 1.00),
      c(0.85, 1.00),
      c(1.00, 0.85),
      c(1.00, 1.00),
      c(1.00, 0.90),
      c(0.85, 1.00, 0.75),
      c(0.45, 1.00, 0.45),
      c(0.90, 0.75, 1.00),
      c(0.55, 1.00, 0.55),
      c(1.00, 0.75, 0.45),
      c(0.70, 1.00, 0.70)
    )),
    stringsAsFactors = FALSE
  )
}

variant_catalog <- function() {
  data.frame(
    variant.id = c("V1", "V2", "V3"),
    variant.name = c(
      "uniform gaussian",
      "center-biased heteroskedastic",
      "gap-biased laplace outliers"
    ),
    stringsAsFactors = FALSE
  )
}

sample_x <- function(n, variant.id) {
  if (identical(variant.id, "V1")) {
    return(sort(runif(n)))
  }
  if (identical(variant.id, "V2")) {
    p0 <- pnorm(0, mean = 0.50, sd = 0.20)
    p1 <- pnorm(1, mean = 0.50, sd = 0.20)
    return(sort(qnorm(runif(n, p0, p1), mean = 0.50, sd = 0.20)))
  }

  n.left <- floor(0.42 * n)
  n.right <- n - n.left
  x <- c(runif(n.left, 0.02, 0.42), runif(n.right, 0.58, 0.98))
  sort(x)
}

rlaplace <- function(n, scale) {
  u <- runif(n) - 0.5
  -scale * sign(u) * log1p(-2 * abs(u))
}

add_noise <- function(x, truth, variant.id) {
  n <- length(x)
  if (identical(variant.id, "V1")) {
    return(truth + rnorm(n, sd = 0.08))
  }
  if (identical(variant.id, "V2")) {
    sd.x <- 0.045 + 0.14 * truth
    return(truth + rnorm(n, sd = sd.x))
  }

  y <- truth + rlaplace(n, scale = 0.055)
  n.out <- max(1L, round(0.04 * n))
  idx <- sample.int(n, n.out)
  y[idx] <- y[idx] + rnorm(n.out, mean = 0, sd = 0.45)
  y
}

make_dataset <- function(shape, variant, replicate, n = 250L, base.seed = 73001L) {
  seed <- base.seed + as.integer(sub("^S", "", shape$shape.id)) * 1000L +
    as.integer(sub("^V", "", variant$variant.id)) * 100L + replicate
  set.seed(seed)
  x <- sample_x(n, variant$variant.id)
  truth <- mixture_signal(x, shape$means[[1]], shape$sds[[1]], shape$weights[[1]])
  y <- add_noise(x, truth, variant$variant.id)
  list(
    seed = seed,
    x = x,
    X = matrix(x, ncol = 1),
    truth = truth,
    y = y
  )
}

safe_min <- function(x) if (all(is.na(x))) NA_real_ else min(x, na.rm = TRUE)
safe_median <- function(x) if (all(is.na(x))) NA_real_ else median(x, na.rm = TRUE)
safe_max <- function(x) if (all(is.na(x))) NA_real_ else max(x, na.rm = TRUE)

conductance_label <- function(rule, alpha, sigma.rule, sigma.quantile, local.k) {
  if (identical(rule, "inverse.length.power")) {
    return(sprintf("%s alpha=%s", rule, signif(alpha, 3)))
  }
  if (identical(rule, "exp.length") || identical(rule, "exp.length.squared")) {
    if (identical(sigma.rule, "edge.quantile")) {
      return(sprintf("%s sigma=edge.q%s", rule, signif(sigma.quantile, 3)))
    }
    return(sprintf("%s sigma=%s", rule, sigma.rule))
  }
  if (identical(rule, "self.tuned.gaussian")) {
    return(sprintf("%s local.k=%d", rule, as.integer(local.k)))
  }
  rule
}

metric_grid <- function() {
  out <- rbind(
    data.frame(conductance.rule = "inverse.length.power",
               alpha = c(0.5, 1, 1.5, 2, 3),
               sigma.rule = NA_character_,
               sigma.quantile = NA_real_,
               local.k = NA_integer_),
    data.frame(conductance.rule = c("exp.length", "exp.length",
                                    "exp.length.squared", "exp.length.squared"),
               alpha = NA_real_,
               sigma.rule = c("median", "edge.quantile", "median", "edge.quantile"),
               sigma.quantile = c(NA_real_, 0.75, NA_real_, 0.75),
               local.k = NA_integer_),
    data.frame(conductance.rule = "self.tuned.gaussian",
               alpha = NA_real_,
               sigma.rule = NA_character_,
               sigma.quantile = NA_real_,
               local.k = c(3L, 5L, 10L))
  )
  out$laplacian.type <- "unnormalized"
  out$filter.type <- "heat_kernel"
  out$n.eigenpairs <- 60L
  out$n.candidates <- 30L
  out$config.id <- sprintf("M%02d", seq_len(nrow(out)))
  out$config.label <- mapply(
    conductance_label,
    out$conductance.rule,
    out$alpha,
    out$sigma.rule,
    out$sigma.quantile,
    out$local.k,
    USE.NAMES = FALSE
  )
  out
}

fit_metric_one <- function(spec, graph, y, n) {
  args <- list(
    adj.list = graph$adj_list,
    weight.list = graph$weight_list,
    y = y,
    conductance.rule = spec$conductance.rule,
    laplacian.type = "unnormalized",
    filter.type = spec$filter.type,
    n.eigenpairs = min(spec$n.eigenpairs, n - 1L),
    n.candidates = spec$n.candidates,
    eigen.solver = "auto",
    dense.eigen.threshold = 200L,
    dense.fallback = "auto",
    dense.fallback.threshold = 5000L,
    verbose = FALSE
  )
  if (identical(spec$conductance.rule, "inverse.length.power")) {
    args$conductance.alpha <- spec$alpha
  }
  if (identical(spec$conductance.rule, "exp.length") ||
      identical(spec$conductance.rule, "exp.length.squared")) {
    args$conductance.sigma.rule <- spec$sigma.rule
    if (!is.na(spec$sigma.quantile)) {
      args$conductance.sigma.quantile <- spec$sigma.quantile
    }
  }
  if (identical(spec$conductance.rule, "self.tuned.gaussian")) {
    args$conductance.local.k <- spec$local.k
  }
  start <- proc.time()
  out <- tryCatch(do.call(fit.metric.graph.lowpass, args), error = function(e) e)
  elapsed <- unname((proc.time() - start)[["elapsed"]])
  list(result = out, elapsed = elapsed)
}

score_fit <- function(method, fit, elapsed, y, truth) {
  if (inherits(fit, "error")) {
    return(data.frame(
      method = method,
      status = "error",
      rmse.truth = NA_real_,
      rmse.observed = NA_real_,
      mae.truth = NA_real_,
      runtime.sec = elapsed,
      eta = NA_real_,
      gcv = NA_real_,
      effective.df = NA_real_,
      eigen.backend = NA_character_,
      conductance.median = NA_real_,
      message = conditionMessage(fit),
      stringsAsFactors = FALSE
    ))
  }
  yhat <- as.numeric(fit$fitted.values)
  if (inherits(fit, "metric.graph.lowpass.fit")) {
    conductance <- fit$operator$edge.table$conductance
    eta <- fit$gcv$eta.optimal
    gcv <- fit$gcv$gcv.optimal
    edf <- fit$gcv$effective.df
    backend <- fit$spectral$backend
  } else {
    conductance <- NA_real_
    eta <- tail(fit$gcv$eta.optimal, 1)
    gcv <- tail(fit$gcv$gcv.optimal, 1)
    edf <- NA_real_
    backend <- if (!is.null(fit$spectral$backend)) fit$spectral$backend else NA_character_
  }
  data.frame(
    method = method,
    status = "ok",
    rmse.truth = sqrt(mean((yhat - truth)^2)),
    rmse.observed = sqrt(mean((yhat - y)^2)),
    mae.truth = mean(abs(yhat - truth)),
    runtime.sec = elapsed,
    eta = eta,
    gcv = gcv,
    effective.df = edf,
    eigen.backend = backend,
    conductance.median = safe_median(conductance),
    message = "",
    stringsAsFactors = FALSE
  )
}
```

## Scenario Catalog

```{r scenario-catalog}
shapes <- shape_catalog()
variants <- variant_catalog()
replicates <- 1:5
graph.k <- 5L
n <- 250L

scenario.grid <- merge(
  merge(shapes[, c("shape.id", "shape.name")], variants, by = NULL),
  data.frame(replicate = replicates),
  by = NULL
)
scenario.grid$dataset.id <- sprintf(
  "%s_%s_R%02d",
  scenario.grid$shape.id,
  scenario.grid$variant.id,
  scenario.grid$replicate
)

data.frame(
  n.shapes = nrow(shapes),
  n.variants = nrow(variants),
  n.replicates = length(replicates),
  n.datasets = nrow(scenario.grid),
  graph.k = graph.k,
  n.per.dataset = n
)
```

```{r shape-catalog-table}
shapes[, c("shape.id", "shape.name", "means", "sds", "weights")]
```

```{r shape-catalog-plot}
x.grid <- seq(0, 1, length.out = 500)
shape.curves <- do.call(rbind, lapply(seq_len(nrow(shapes)), function(i) {
  data.frame(
    shape.id = shapes$shape.id[i],
    shape.name = shapes$shape.name[i],
    x = x.grid,
    truth = mixture_signal(x.grid, shapes$means[[i]], shapes$sds[[i]], shapes$weights[[i]])
  )
}))

ggplot(shape.curves, aes(x, truth)) +
  geom_line(color = "red", linewidth = 0.8) +
  facet_wrap(~ shape.id + shape.name, ncol = 4) +
  labs(title = "Curated Gaussian-mixture shape catalog", x = "x", y = "scaled truth") +
  theme_minimal(base_size = 10)
```

## Model Grid

```{r model-grid}
metric.configs <- metric_grid()
metric.configs[, c("config.id", "config.label", "filter.type", "n.eigenpairs")]
```

## Run Benchmark

The full fit table is generated outside this report by:

```r
Rscript dev/metric-graph-lowpass-benchmarks/compute_multi_scenario_1d_mixture_results.R
```

The report reads the precomputed table below.

```{r load-results}
results.file <- file.path(
  repo.dir,
  "dev/metric-graph-lowpass-benchmarks/reports/multi_scenario_1d_mixture_results.csv"
)
if (!file.exists(results.file)) {
  stop("Missing precomputed results file. Run compute_multi_scenario_1d_mixture_results.R first.")
}
results <- read.csv(results.file, stringsAsFactors = FALSE)

data.frame(
  n.rows = nrow(results),
  n.datasets = length(unique(results$dataset.id)),
  n.metric.rows = sum(results$method == "fit.metric.graph.lowpass"),
  n.rdgraph.rows = sum(results$method == "fit.rdgraph.regression"),
  n.errors = sum(results$status != "ok"),
  results.file = results.file
)
```

```{r run-benchmark-disabled, eval=FALSE, include=FALSE}
results <- vector("list", nrow(scenario.grid) * (nrow(metric.configs) + 1L))
example.fits <- list()
idx <- 0L

for (s in seq_len(nrow(scenario.grid))) {
  shape <- shapes[match(scenario.grid$shape.id[s], shapes$shape.id), ]
  variant <- variants[match(scenario.grid$variant.id[s], variants$variant.id), ]
  dat <- make_dataset(shape, variant, scenario.grid$replicate[s], n = n)

  graph.start <- proc.time()
  graph <- tryCatch(
    create.sknn.graph(dat$X, k = graph.k, connect.components = TRUE,
                      connect.method = "component.mst"),
    error = function(e) e
  )
  graph.elapsed <- unname((proc.time() - graph.start)[["elapsed"]])

  if (inherits(graph, "error")) {
    idx <- idx + 1L
    results[[idx]] <- cbind(
      scenario.grid[s, ],
      data.frame(
        graph.k = graph.k,
        n.components.before = NA_integer_,
        n.components.after = NA_integer_,
        n.mst.edges.added = NA_integer_,
        graph.runtime.sec = graph.elapsed,
        config.id = "rdgraph",
        config.label = "fit.rdgraph.regression",
        score_fit("fit.rdgraph.regression", graph, NA_real_, dat$y, dat$truth),
        stringsAsFactors = FALSE
      )
    )
    next
  }

  rd.start <- proc.time()
  rd.fit <- tryCatch(
    fit.rdgraph.regression(
      X = dat$X,
      y = dat$y,
      k = graph.k,
      adj.list = graph$adj_list,
      weight.list = graph$weight_list,
      verbose.level = 0L
    ),
    error = function(e) e
  )
  rd.elapsed <- unname((proc.time() - rd.start)[["elapsed"]])
  idx <- idx + 1L
  results[[idx]] <- cbind(
    scenario.grid[s, ],
    data.frame(
      graph.k = graph.k,
      n.components.before = graph$n_components_before,
      n.components.after = graph$n_components_after,
      n.mst.edges.added = graph$n_mst_edges_added,
      graph.runtime.sec = graph.elapsed,
      config.id = "rdgraph",
      config.label = "fit.rdgraph.regression",
      score_fit("fit.rdgraph.regression", rd.fit, rd.elapsed, dat$y, dat$truth),
      stringsAsFactors = FALSE
    )
  )

  for (m in seq_len(nrow(metric.configs))) {
    spec <- metric.configs[m, ]
    metric.out <- fit_metric_one(spec, graph, dat$y, n = n)
    idx <- idx + 1L
    results[[idx]] <- cbind(
      scenario.grid[s, ],
      data.frame(
        graph.k = graph.k,
        n.components.before = graph$n_components_before,
        n.components.after = graph$n_components_after,
        n.mst.edges.added = graph$n_mst_edges_added,
        graph.runtime.sec = graph.elapsed,
        config.id = spec$config.id,
        config.label = spec$config.label,
        score_fit("fit.metric.graph.lowpass", metric.out$result,
                  metric.out$elapsed, dat$y, dat$truth),
        stringsAsFactors = FALSE
      )
    )

    if (scenario.grid$replicate[s] == 1L &&
        scenario.grid$shape.id[s] %in% c("S01", "S05", "S10", "S14") &&
        scenario.grid$variant.id[s] == "V1" &&
        spec$config.id %in% c("M01", "M02", "M06", "M10")) {
      example.fits[[length(example.fits) + 1L]] <- list(
        dataset = scenario.grid[s, ],
        spec = spec,
        x = dat$x,
        y = dat$y,
        truth = dat$truth,
        fit = metric.out$result
      )
    }
  }
}

results <- do.call(rbind, results[seq_len(idx)])
rownames(results) <- NULL

write.csv(
  results,
  file = file.path(repo.dir, "dev/metric-graph-lowpass-benchmarks/reports/multi_scenario_1d_mixture_results.csv"),
  row.names = FALSE
)

data.frame(
  n.rows = nrow(results),
  n.datasets = length(unique(results$dataset.id)),
  n.metric.rows = sum(results$method == "fit.metric.graph.lowpass"),
  n.rdgraph.rows = sum(results$method == "fit.rdgraph.regression"),
  n.errors = sum(results$status != "ok")
)
```

## Overall Performance

```{r overall-summary}
ok <- subset(results, status == "ok")
metric.ok <- subset(ok, method == "fit.metric.graph.lowpass")
rd.ok <- subset(ok, method == "fit.rdgraph.regression")

overall <- aggregate(
  rmse.truth ~ method,
  data = ok,
  FUN = function(z) c(median = median(z), q25 = quantile(z, 0.25),
                      q75 = quantile(z, 0.75), mean = mean(z))
)
do.call(data.frame, overall)
```

```{r config-ranking}
config.ranking <- aggregate(
  rmse.truth ~ config.id + config.label + method,
  data = metric.ok,
  FUN = function(z) c(median = median(z), mean = mean(z), q90 = quantile(z, 0.90))
)
config.ranking <- do.call(data.frame, config.ranking)
names(config.ranking) <- sub("rmse.truth.", "", names(config.ranking), fixed = TRUE)
config.ranking <- config.ranking[order(config.ranking$median), ]
head(config.ranking, 12)
```

```{r overall-boxplot}
plot.df <- ok
plot.df$display.method <- ifelse(
  plot.df$method == "fit.rdgraph.regression",
  "rdgraph baseline",
  plot.df$config.label
)
top.configs <- head(config.ranking$config.label, 8)
plot.df <- subset(plot.df, method == "fit.rdgraph.regression" | config.label %in% top.configs)

ggplot(plot.df, aes(x = reorder(display.method, rmse.truth, median), y = rmse.truth,
                    fill = method)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.65, color = "gray35") +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2.2,
               fill = "white", color = "black") +
  scale_y_log10() +
  coord_flip() +
  labs(
    title = "RMSE to truth across 240 curated datasets",
    subtitle = "Violin distributions for top 8 metric configurations by median RMSE plus rdgraph baseline; white diamonds mark medians",
    x = NULL,
    y = "RMSE to truth, log10 scale",
    fill = "Method family"
  ) +
  theme_minimal(base_size = 11)
```

## Per-Scenario Winners

```{r per-dataset-best}
best.metric.by.dataset <- metric.ok[order(metric.ok$dataset.id, metric.ok$rmse.truth), ]
best.metric.by.dataset <- best.metric.by.dataset[!duplicated(best.metric.by.dataset$dataset.id), ]
rd.by.dataset <- rd.ok[, c("dataset.id", "rmse.truth")]
names(rd.by.dataset)[2] <- "rdgraph.rmse.truth"
best.compare <- merge(best.metric.by.dataset, rd.by.dataset, by = "dataset.id", all.x = TRUE)
best.compare$metric.beats.rdgraph <- best.compare$rmse.truth < best.compare$rdgraph.rmse.truth

data.frame(
  datasets = nrow(best.compare),
  metric.beats.rdgraph = sum(best.compare$metric.beats.rdgraph, na.rm = TRUE),
  metric.beats.rdgraph.frac = mean(best.compare$metric.beats.rdgraph, na.rm = TRUE),
  median.best.metric.rmse = median(best.compare$rmse.truth, na.rm = TRUE),
  median.rdgraph.rmse = median(best.compare$rdgraph.rmse.truth, na.rm = TRUE)
)
```

```{r winners-by-variant}
winner.summary <- aggregate(
  metric.beats.rdgraph ~ variant.id + variant.name,
  data = best.compare,
  FUN = mean
)
winner.summary
```

```{r best-metric-vs-rdgraph}
ggplot(best.compare, aes(x = rdgraph.rmse.truth, y = rmse.truth, color = variant.name)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  geom_point(alpha = 0.75, size = 1.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Best metric low-pass fit versus rdgraph baseline by dataset",
    x = "rdgraph RMSE to truth, log10 scale",
    y = "Best metric RMSE to truth, log10 scale",
    color = "Variant"
  ) +
  theme_minimal(base_size = 12)
```

## Scenario and Noise Pattern Effects

```{r scenario-heatmap}
shape.variant.summary <- aggregate(
  rmse.truth ~ shape.id + shape.name + variant.id + variant.name + method,
  data = ok,
  FUN = median
)
shape.variant.best.metric <- aggregate(
  rmse.truth ~ shape.id + shape.name + variant.id + variant.name,
  data = metric.ok,
  FUN = min
)
shape.variant.rd <- subset(shape.variant.summary, method == "fit.rdgraph.regression")
names(shape.variant.rd)[names(shape.variant.rd) == "rmse.truth"] <- "rdgraph.median.rmse"
shape.variant.compare <- merge(shape.variant.best.metric, shape.variant.rd,
                               by = c("shape.id", "shape.name", "variant.id", "variant.name"),
                               all.x = TRUE)
shape.variant.compare$log10.metric.over.rdgraph <- log10(shape.variant.compare$rmse.truth /
                                                           shape.variant.compare$rdgraph.median.rmse)

ggplot(shape.variant.compare,
       aes(x = variant.id, y = shape.id, fill = log10.metric.over.rdgraph)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0) +
  labs(
    title = "Best metric RMSE relative to rdgraph by shape and variant",
    subtitle = "Blue means best metric low-pass has lower RMSE than rdgraph",
    x = "Variant",
    y = "Shape",
    fill = "log10(metric / rdgraph)"
  ) +
  theme_minimal(base_size = 12)
```

```{r conductance-by-variant}
metric.ok$conductance.family <- sub(" .*", "", metric.ok$config.label)
ggplot(metric.ok, aes(x = conductance.family, y = rmse.truth, fill = variant.name)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_y_log10() +
  labs(
    title = "Metric low-pass RMSE by conductance family and noise/sampling variant",
    x = "Conductance family",
    y = "RMSE to truth, log10 scale",
    fill = "Variant"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Example Fits

```{r example-fits}
example.scenarios <- subset(
  scenario.grid,
  replicate == 1L & variant.id == "V1" & shape.id %in% c("S01", "S05", "S10", "S14")
)
example.config.labels <- head(config.ranking$config.label, 3)
example.configs <- metric.configs[match(example.config.labels, metric.configs$config.label), ]

example.df <- do.call(rbind, lapply(seq_len(nrow(example.scenarios)), function(i) {
  shape <- shapes[match(example.scenarios$shape.id[i], shapes$shape.id), ]
  variant <- variants[match(example.scenarios$variant.id[i], variants$variant.id), ]
  dat <- make_dataset(shape, variant, example.scenarios$replicate[i], n = n)
  graph <- create.sknn.graph(dat$X, k = graph.k, connect.components = TRUE,
                             connect.method = "component.mst")
  do.call(rbind, lapply(seq_len(nrow(example.configs)), function(j) {
    spec <- example.configs[j, ]
    fit <- fit_metric_one(spec, graph, dat$y, n = n)$result
    if (inherits(fit, "error")) return(NULL)
    data.frame(
      dataset.id = example.scenarios$dataset.id[i],
      shape.id = example.scenarios$shape.id[i],
      shape.name = example.scenarios$shape.name[i],
      config.label = spec$config.label,
      x = dat$x,
      y = dat$y,
      truth = dat$truth,
      fitted = fit$fitted.values
    )
  }))
}))

if (!is.null(example.df) && nrow(example.df) > 0) {
  ggplot(example.df, aes(x, y)) +
    geom_point(color = "black", alpha = 0.35, size = 0.8) +
    geom_line(aes(y = truth), color = "red", linewidth = 0.7) +
    geom_line(aes(y = fitted, color = config.label), linewidth = 0.7) +
    facet_wrap(~ shape.id + shape.name, scales = "free_y", ncol = 2) +
    labs(
      title = "Representative metric low-pass fits",
      x = "x",
      y = "response",
      color = "Metric configuration"
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
}
```

## Runtime and Failures

```{r runtime}
ggplot(ok, aes(x = method, y = runtime.sec, fill = method)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_log10() +
  labs(
    title = "Runtime per fit",
    x = NULL,
    y = "Runtime seconds, log10 scale"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r failure-table}
failure.table <- subset(results, status != "ok")
if (nrow(failure.table) == 0) {
  data.frame(message = "No fit failures")
} else {
  aggregate(dataset.id ~ method + config.label + message, data = failure.table, FUN = length)
}
```

## Full Result Location

The full result table was written to:

```{r result-file}
file.path(repo.dir, "dev/metric-graph-lowpass-benchmarks/reports/multi_scenario_1d_mixture_results.csv")
```
